---
title: "Hands-on Exercise 2 - Beyond ggplot2 Fundamentals"
author: "Amelia Rosalind"
date: "April 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting started

## Installing and loading required libraries

For this exercise, besides tidyverse, four R packages will be used. They are:

-   ggrepel: provides geoms for ggplot2 to repel overlapping text labels
-   ggthemes: provides some extra themes, geoms, and scales for ggplot2
-   hrbrthemes: provides typography-centric thems and theme components for ggplot2
-   patchwork: for preparing composite figure created using ggplot2

The code chunk below will check if the packages needed are installed and load them into R environment.

```{r}
pacman::p_load(ggrepel, patchwork, ggthemes,
               hrbrthemes, tidyverse)
```

## Importing data

The data file used is called Exam_data which consists of year end examination grades of a cohort of primary 3 students from a local school. It is in csv file format.

```{r}
exam_data <- read_csv("Exam_data.csv")
summary(exam_data)
```

From the summary, there area total of seven attributes in the exam_data data frame. Four of them are categorical data type and the other three are continuous data type.

-   Categorical attributes are: ID, CLASS, GENDER and RACE
-   Continuous attributes are: MATHS, ENGLISH and SCIENCE

# Beyond ggplot2 Annotation: ggrepel

One of the challenge in plotting statistical graph is annotation, especially with large number of data points. See the graph below for reference.

```{r}
ggplot(data=exam_data,
        aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label(aes(label = ID),
             hjust = .5,
             vjust = -.5) +
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores verses Maths scores for Primary 3")
```

[ggrepel](https://ggrepel.slowkow.com) is an extension of ggplot2 package which provides geoms for ggplot2 to repel overlapping text. We simply replace `geom_text()` by geom_text_repel() and `geom_label()` by [geom_label_repel()](https://ggrepel.slowkow.com).

```{r}
ggplot(data=exam_data,
        aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label_repel(aes(label = ID),
                   fontface = 'bold') +
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores verses Maths scores for Primary 3")
```

# Beyond ggplot2 Themes

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) which are: theme_gray(), theme_bw(), theme_classic(), theme_dark(), theme_light(), theme_linedraw(), theme_minimal() and theme_void().

::: panel-tabset
## Without

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20, 
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths score")
```

## With theme_dark()

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20, 
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  theme_dark() +
  ggtitle("Distribution of Maths score")
```
:::

## ggtheme package

[ggtheme](https://cran.r-project.org/web/packages/ggthemes/index.html) provides ggplot2 themes that replicate the look fo plots by Edward Tufte, Stephen Few, Fivethirteight, The Economist, Stata, Excel and The Wall Street Journal among others.

The code below used The Exonomist theme.

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20, 
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths score") +
  theme_economist()
```

::: callout-tip
It has many themes and provides some extra geoms and scales for ggplot2. Consult [this site](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) to see the various options.
:::

## hrbrthemes packages

hrbrthemes package provides base theme that focuses on typograohic elements including where various lables are placed as well as the fonts that are used.

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20, 
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths score") +
  theme_ipsum()
```

The second goal centers around productivity for a production workflow where presenting something needs to be visually clean and readable to increase reader's engagement. Consult [this site](https://hrbrmstr.github.io/hrbrthemes/articles/why_hrbrthemes.html) to learn more.

```{r}
ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20, 
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths score") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = 'Y')
```

::: {.callout-tip title="What can be learnt from the code chunk above?"}
-   `axis_title_size`: controls font size of the axis title
-   `base_size`: controls the size of axis label
-   `grid`: controls the lines in the graph, in this case remove the x-axis grid lines
:::

# Beyond single graph

It is possible to create composite plot by combining multiple graphs. First, lets create three statistical graphics.

::: panel-tabset
## Maths

```{r}
p1 <- ggplot(data=exam_data,
             aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  coord_cartesian(xlim = c(0,100)) +
  ggtitle("Distributions of Maths scores")

p1
```

## English

```{r}
p2 <- ggplot(data=exam_data,
             aes(x = ENGLISH)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  coord_cartesian(xlim = c(0,100)) +
  ggtitle("Distributions of English scores")
 
p2
```

## English vs Maths

```{r}
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

p3
```
:::

## patchwork methods

There are several extension function that support composite figure by combining several graphs such as `grid.arrange()` of **gridExtra** package and `plot_grid()` of **cowplot** package. This section will focus on an extension called [patchwork](https://patchwork.data-imaginist.com) which is specially designed for combining ggplot2 graphs into a single figure.

Patchwok package has a very simple syntax wherewe can create layouts super easily. Here's the general syntax:

-   Two-column layout using the plus sign +
-   Subplot group ysing paranthesis ()
-   Two-row layout using division sign /
-   Plot beside each other using \|

Examples:

::: panel-tabset
## Combine 2 ggplot graphs

```{r}
#| fig-width: 12
#| fig-height: 5
p1 + p2
```

## combine 3 ggplot graphs

```{r}
#| fig-width: 12
#| fig-height: 8
(p1 / p2) | p3
```
:::

::: callout-tip
Another way is to use `wrap_plots()` to directly plot all the graphs into one section. Use `#| fig-width` and `#| fig-height` to change the size or scale of the figure. This one uses 15 and 7 respectively.

```{r}
#| fig-width: 15
#| fig-height: 7
wrap_plots(p1, p2, p3)
```
:::

To learn more, refer to [Plot Assembly](https://patchwork.data-imaginist.com/articles/guides/assembly.html). The site has information about adding a table or some text and how to place it to the left of the graph.

## Creating composite figure with tag

In order to identify subplots in text, patchwork also provides auto-tagging capabilities as shown below.

```{r}
#| fig-width: 12
#| fig-height: 8
((p1 / p2) | p3) +
  plot_annotation(tag_levels = 'I')
```

## Creating figure with inser

It is also possible to insert one or more plots or graphic elements freely on top or below another plot using [inset_element()](https://patchwork.data-imaginist.com/reference/inset_element.html).

```{r}
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1,
                   align_to = 'full')
```

::: callout-note
`align_to` specifies where the element should be relative to. 'panel' is the default, 'full' is for the whole image, 'plot' is for the plot area of the main layer.
:::

## Creating composite figure using patchwork and ggtheme

```{r}
#| fig-width: 15
#| fig-height: 8
patchwork <- ((p1 / p2) | p3) +
  plot_annotation(tag_levels = "1")
patchwork & theme_economist(dkpanel=TRUE) 
```

::: callout-tip
It is not possible to add both ggtheme and hrbrtheme as both as solid themes so adding both would result on the last theme added. For example if the code was `pathwork & theme_economist() & theme_ipsum()` the output will have the ipsum theme and no economist theme.
:::
