---
title: "Take-Home Exercise 2 - VAST Mini Challenge 2"
author: "Amelia Rosalind"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

# Getting Started

## Installing of Required Packages

| Packages Used | Purpose |
|------------------------------------|------------------------------------|
| tidyverse | data wrangling |
| jsonlite | JSON import |
| igraph, tidygraph, ggraph | building and visualising static network graphs |
| vizNetwork, DT | interactive network exploration and data tables |
| ggrepel, scales | improving plot visualisation through annotaiton and axis formatting |
| knitr | for clean table rendering |

: {tbl-colwidths="\[25,75\]"}

```{r}
pacman::p_load(tidyverse, jsonlite, igraph, tidygraph, 
               ggraph, visNetwork, DT, ggrepel,
               knitr, scales, patchwork)
```

## Importing and Inspecting Data

```{r}
mc2_raw <- fromJSON("mc2_data/MC2_data.json", flatten = TRUE)
org_raw <- fromJSON("mc2_data/org_chart.json", flatten = TRUE)

str(mc2_raw, max.level = 2)
names(mc2_raw)

str(org_raw, max.level = 2)
names(org_raw)
```

::: callout-note
### Key Findings

The MC2 data is structured as a nested JSON with two top-level fields: `descriptions` a (string) and `events` (a data frame of 185,147 rows). The org chart contains nodes (people, teams, department) and directed edges (relationships between them).
:::

# Understanding the Main Data

## Data wrangling

First separate the raw JSON into its 2 components, decription string and the events data frame. The events are converted to a tibble for easier manipulation.

```{r}
mc2_desc <- mc2_raw$description
mc2_events <- as_tibble(mc2_raw$events)
```

Then parse the Unix timestamp stored in `when` of the events tibble into readable `datetime` column.

```{r}
mc2_events <- mc2_events %>%
  mutate(datetime = as_datetime(when))

glimpse(mc2_events)
```

::: callout-note
### Key Findings

The events data frame contains 185,147 rows and 69 columns after adding the parsed datetime. Most columns are sparesly populated with NAs, which is expected since different event types use different fields.
:::

## Event type distribution

Check to see how many unique event types are in the dataset and its distribution.

```{r}
n_distinct(mc2_events$short_name)
event_count <- mc2_events %>% count(short_name, sort = TRUE)

ggplot(data = event_count,
       aes(x = reorder(short_name, n), 
           y = n)) +
  geom_col(fill = "#4e79a7") +
  geom_text(aes(label = scales::comma(n)),
            hjust = -0.1,
            size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = comma) +
  labs(title = "Distribution of Event Types in MC2 Data",
       subtitle = "Total of 24 unique event types across 185,147 events",
       x = NULL,
       y = "Number of Events") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.text.y = element_text(size = 9)
  )
```

::: callout-note
### Key Findings

There are 24 unique event types in the dataset. The bar chart below shows their distributions. Note that the distributions are not equal. Events that dominate the log include `check_email`, `assign_agent_task`, `read_file` and `queue_subordinate_task`, with each accounting for roughly 17,000 - 37,000 events. Meanwhile, the rarer types such as `saidit_post` (108 events) and `post_saidit` (133 events) are almost negligible in scale. Yet considering the problem, one of the 2 should be the event type in question about the anomalous post in SaidIt by John Windward.
:::

# Identify the Anomalous SaidIT Post

## Find the Anomalous Post

The investigation should begin with the known anomalous SaidIT post described in the challenge. To locate it, all events with non-null `details.forum` field are extracted which should represent any SaidIT-related activity.

```{r}
forum_post <- mc2_events %>%
  filter(!is.na(details.forum)) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, details.timestamp)

kable(head(forum_post, n = 10))
```

::: callout-note
### Key Finding

From just the head, it can be concluded that there is difference between `post_saidit` and `saidit_post` through the `parties`, `details.poster_id` and `details.timestamp` fields. `post_saidit` has `parties` field as `Agent/person:ava_tiller,system:saidit` or `Agent/person:aria_towline,system:saidit` and the `details.poster_id` and `details.timestamp` are null. `saidit_post` has `parties` and `details.parties_id` field as `person:ava_tiler` or `person:aria_towline`. Also, `details.timestamp` is not null.

It can be assumed that `saidit_post` are done by the person themselves, while `post_saidit` are done by the agent.
:::

Since the challenge description states the post was made in John Windward's name at an unusual hour, all of John Windward's events on May 17 are examined to understand what he was actually doing at the time of the post.

```{r}
john_may17 <- mc2_events %>%
  filter(map_lgl(parties, ~ "person:john_windward" %in% .x)) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, details.from, details.to,
         details.content, details.forum, details.poster_id,
         details.a2a.from, details.a2a.to, details.a2a.action)

datatable(john_may17,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "John Windward's Events on May 17, 2046")
```

::: callout-note
### Key Findings

Looking at the result, John's only recorded activity on May 17 consists entirely of `check_email` events, clustered in 2 time window - early morning around 7:32-7:39 UTC and late-night around 23.9-23:18 UTC. In addition, there is no record of the anomalous post done, strongly suggesting that his agent acted without his knowledge or instruction.
:::

Since it can't be found in that way, we can find the anomalous post by isolating all SaidIT post in May 17.

```{r}
anomalous_post <- mc2_events %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, 
         details.content_source, details.timestamp)

datatable(anomalous_post, 
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "SaidIT Posts on May 17, 2046")
```

::: callout-note
### Key Findings

Only one SaidIT post occurred on May 17 with event ID 373902 at 2046-05-17 11:21:15 UTC which should be the 4.21am anomalous post in the challenge. Considering the timing, John was inactive. The `parties` field showing `Agent/person:john_windward` rather than `person:john_windward`, which also further proves that it was John's agent acting autonomously, not John himself.

This event is also an anomaly as it was considered as `saidit_post` event but the `parties` field is `Agent/person:john_windward,system:saidit` and there is nothing on the field `detail.poster_id` and `details.timestamp`. This is different from the normal `saidit_post` established above.
:::

## SaidIT Post Timeline

To contextualize this post against all other SaidIT activity in the dataset, the timeline below plots every `saidit_post` and `post_saidit` event across the full data range while also highlighting the anomalous post.

```{r}
#| code-fold: true
saidit_posts <- mc2_events %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  mutate(
    is_anomalous = id == 373902,
    post_date = as.Date(datetime),
    label = if_else(id == 373902, 
                    "Anomalous\nJohn Windward\npost",
                    NA_character_)
  )

ggplot(data = saidit_posts,
       aes(x = datetime,
           y = short_name,
           colour = is_anomalous,
           size = is_anomalous)) +
  geom_point(alpha = 0.8) +
  geom_vline(xintercept = as.POSIXct("2046-05-17 11:21:15", 
                                     tz = "UTC"),
             colour = "#e15759",
             linetype = "dashed",
             linewidth = 0.7) +
  geom_label_repel(
    aes(label = label),
    na.rm = TRUE,
    fill = "#e15759",
    colour = "white",
    size = 3,
    fontface = "bold",
    nudge_y = 0.4,
    segment.colour = "#e15759"
  ) +
  scale_colour_manual(values = c("FALSE" = "grey70", 
                                 "TRUE" = "#e15759"),
                      labels = c("Normal post", 
                                 "Anomalous post")) +
  scale_size_manual(values = c("FALSE" = 1.5, 
                               "TRUE" = 4),
                    guide = "none") +
  labs(title = "All SaidIT-related Posts Over Time",
       subtitle = "John Windward's autonomous post on May 17 stands out as the only anomaly",
       x = NULL,
       y = NULL,
       colour = NULL) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    axis.text.y = element_text(size = 10)
  )
```

## Key Takeaways

The post has `short_name` as `saidit_post` but under there are anomalies on `parties`, `details.poster_id` and `details.timestamp`. `parties` field alone having `Agent/person:john_windward,system:saidit` is already weird as a normal `saidit_post` event should be done by the person so John himself. But, the `parties` filed shows that its posted by John's agent instead.

In addition based on John's event on May 17, there was only `check_email`. He was also not active on 4:21 am, which means his agent acted autonomously.

# Organisational Context

## Inspect Nodes and Edges Only

Before tracing the event chain, it is important to understand the organisational structure of Tenant Thread and where John Windward sits within it. To do this first separate out the organisation nodes and edges as tibbles.

```{r}
org_nodes <- as_tibble(org_raw$nodes)
org_edges <- as_tibble(org_raw$edges)

glimpse(org_nodes)
glimpse(org_edges)
```

::: callout-note
### Key Findings

The organisation chart data contains 75 nodes representing the company, its departments, teams, and individual people, connected by 74 directed edges describing containment and leadership relationships.
:::

## Understanding John Windward

With the structure confirmed, John Windward's position in the organisation is identified by filtering the nodes and edges for his person ID.

```{r}
org_nodes %>% filter(str_detect(id, "john_windward"))

org_edges %>%
  filter(str_detect(source, "john_windward") |
           str_detect(target, "john_windward"))
```

::: callout-note
### Key Findings

John is a Department Lead with a `led_by` relationship from `department:customer_support`, meaning he leads that department.
:::

To understand the full scope of his department, all nodes connected to `customer_support` are retrieved.

```{r}
org_edges %>%
  filter(str_detect(source, "customer_support") |
           str_detect(target, "customer_support")) %>%
  left_join(org_nodes, by = c("target" = "id"))
```

::: callout-note
### Key Findings

The results confirm that Customer Support contains four teams — Phone Center, Billing, Concierge, and Public Relations — all sitting directly under John's leadership.
:::

## Organisation chart visualisation

The organisation chart below visualizes the full Tenant Thread structure, with John Windward highlighted in red and his Customer Support department in orange. This helps contextualize where the anomalous post originated within the company hierarchy.

```{r}
#| code-fold: true
#| column: page
org_nodes_vis <- org_nodes %>%
  rename(job_title = title) %>%
  mutate(
    group = case_when(
      type == "company" ~ "Company",
      type == "department" ~ "Department",
      type == "team" ~ "Team",
      type == "person" ~ "Person",
      TRUE ~ "Other"
    ),
    color = case_when(
      id == "person:john_windward" ~ "#e15759",
      type == "company" ~ "#59a14f",
      type == "department" ~ "#4e79a7",
      type == "team" ~ "#76b7b2",
      type == "person" ~ "#f28e2b",
      TRUE ~ "grey80"
    ),
    font.size = case_when(
      type == "company" ~ 20,
      type == "department" ~ 16,
      type == "team" ~ 14,
      type == "person" ~ 13,
      TRUE ~ 12
    ),
    size = case_when(
      type == "company" ~ 40,
      type == "department" ~ 30,
      type == "team" ~ 20,
      type == "person" ~ 15,
      TRUE ~ 15
    )
  ) %>%
  left_join(
    org_edges %>%
      filter(relation == "led_by") %>%
      left_join(
        org_nodes %>% 
          rename(job_title = title) %>%
          select(id, label, job_title),
        by = c("target" = "id")
      ) %>%
      rename(leader_name = label,
             leader_title = job_title,
             dept_id = source),
    by = c("id" = "dept_id")
  ) %>%
  mutate(
    title = case_when(
      type == "company" ~ paste0(
        "<b>", label, "</b><br>",
        "Type: Company"
      ),
      type == "department" & !is.na(leader_name) ~ paste0(
        "<b>", label, "</b><br>",
        "Department<br>",
        "Led by: <b>", leader_name, "</b><br>",
        "Title: ", coalesce(leader_title, "Department Lead")
      ),
      type == "department" ~ paste0(
        "<b>", label, "</b><br>",
        "Department"
      ),
      type == "team" ~ paste0(
        "<b>", label, "</b><br>",
        "Team"
      ),
      type == "person" & !is.na(job_title) ~ paste0(
        "<b>", label, "</b><br>",
        "Role: ", job_title
      ),
      type == "person" ~ paste0(
        "<b>", label, "</b><br>",
        "Person"
      ),
      TRUE ~ label
    ),
    label = case_when(
      id %in% (org_edges %>%
                 filter(source == "department:executive_suite",
                        relation == "contains") %>%
                 pull(target)) ~ coalesce(job_title, label),
      TRUE ~ label
    )
  ) %>%
  select(id, label, title, group, color, size, font.size)

org_edges_vis <- org_edges %>%
  mutate(
    from = source,
    to = target,
    arrows = "to",
    dashes = relation != "contains",
    title = relation
  ) %>%
  select(from, to, arrows, dashes, title)

visNetwork(nodes = org_nodes_vis,
           edges = org_edges_vis,
           main = "Tenant Thread Organisation Chart",
           height = "800px",
           width = "100%") %>%
  visGroups(groupname = "Company", color = "#59a14f") %>%
  visGroups(groupname = "Department", color = "#4e79a7") %>%
  visGroups(groupname = "Team", color = "#76b7b2") %>%
  visGroups(groupname = "Person", color = "#f28e2b") %>%
  visEdges(smooth = list(type = "cubicBezier",
                         forceDirection = "vertical")) %>%
  visOptions(highlightNearest = list(enabled = TRUE,
                                     degree = 1,
                                     hover = TRUE)) %>%
  visHierarchicalLayout(direction = "UD",
                        sortMethod = "directed",
                        nodeSpacing = 80,
                        levelSeparation = 120) %>%
  visPhysics(enabled = FALSE)
```

::: callout-note
### Key Findings

John's position as a Department Lead in Customer Support places him far from the AI systems team, `team:ai_systems`. His department handles customer-facing functions across four teams that have no direct involvement in managing AI agents or system infrastructure.
:::

## Key Takeaways

# Trace the Immediate Trigger Chain

With the anomalous post identified, the chain of events leading up to it is traced backwards step by step. Starting from the anchor event and working backwards, the immediate chain of events leading up to the post is reconstructed step by step.

## Full details of anchor event (anomalous event)

The full details of the anchor event are first confirmed to identify any useful fields, particularly who was involved and what content was referenced.

```{r}
anchor <- mc2_events %>% filter(id == 373902)
anchor_time <- anchor$datetime

anchor %>%
  select(where(~!all(is.na(.)))) %>%
  select(id, datetime, short_name, parties, 
         details.forum, details.content_source) %>%
  kable(caption = "Full details of the anomalous anchor event (ID 373902)")
```

The anchor event confirms the content source as `SwiftWren.txt`. To find what directly triggered it, all events involving John Windward in the two hours before the post are extracted

## Check any activities with John in parties within 2 hours before post

```{r}
pre_post <- mc2_events %>%
  filter(datetime >= anchor_time - hours(2),
         datetime <= anchor_time) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "john_windward")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(pre_post,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "2 hours before anomalous post - John Windward events")
```

::: callout-note
### Key Findings

The anchor event content source is from `SwiftWren.txt` which should be the file that contained the gibberish content. In addition, from the 2-hour window, there are only 3 events related to John shows and they all happen within 3 seconds. `Agent/person:chloe_ballast` queued a task for John's agent to read a file `SwiftWren_further_instructions.md` which triggered John's agent to do a `saidit_post` event 2 seconds later.
:::

Since Chloe Ballast's agent initiated the task that triggered John's agent, the next step is to examine what Chloe's agent was doing in the same two-hour window. In addition, if she was triggered by someone else, see if Chloe's agent got the same task.

## What was Chloe's agent doing just before

```{r}
chloe_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(2),
         datetime <= anchor_time) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "chloe_ballast")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(chloe_pre,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "Chloe Ballast agent events in 2 hours before anomalous post")
```

::: callout-note
### Key Findings

Chloe Ballast's agent received a `queue_suborinate_task` from Daniel Gangway's agent with the same instruction to read `SwiftWren_further_instructions.md`. This should be the trigger for Chloe's agent to send the same instructions to John's agent.

A thing to note, according to the organisation chart, Chloe is a department head of Information Technologies and Daniel is a member of the hiring team which is not a team Chloe monitors. Considering the name of the event `queue_subordinate_task`, it seems like Daniel is above Chloe which is not true.
:::

Next is to find out Daniel's relevant activity but push back 3 hours.

## What was Daniel's agent doing before triggering Chloe's agent

```{r}
daniel_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(3),
         datetime <= as.POSIXct("2046-05-17 11:12:24",
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "daniel_gangway")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(daniel_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Daniel Gangway events before triggering Chloe's agent")
```

::: callout-note
### Key Findings

Daniel's agent similarly received a `queue_subordinate_task` from Lily Anchorline's agent at 10:11:31Z. Notably, Daniel's agent also performed a `saidit_post_check` a second later but did not proceed to post, unlike John's agent.
:::

Next is to trace Lily Anchorline's event.

## What was Lily's agent doing before triggering Daniel's?

```{r}
lily_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(6),
         datetime <= as.POSIXct("2046-05-17 10:11:31",
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "lily_anchorline")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(lily_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Lily Anchorline events before triggering Daniel's agent")
```

::: callout-note
### Key Findings

Lily's agent similarly received a `queue_subordinate_task` from Zoey Drydock's before sending the same task to Daniel's agent.
:::

Next is to trace back Zoey's event.

## What was Zoey's agent doing before triggering Lily's?

```{r}
zoey_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(6),
         datetime <= as.POSIXct("2046-05-17 08:16:07",
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "zoey_drydock")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(zoey_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Zoey Drydock events before triggering Lily's agent")
```

::: callout-note
### Key Findings

It seems that roughly 3 hours prior to Zoey's agent giving a `queue_subordinate_task` to Lily's agent, she sent the same event to Mia Fender's agent at 5:34 UTC time. Yet, she received the same task from Victoria Rigging's agent at 6:44 UTC time, before doing a `saidit_post_check` a second later and trigger it to send the same task to Lily.
:::

## Immediate trigger chain timeline

The timeline below shows the events leading up to John's anomalous post under the 70 minutes.

```{r}
#| code-fold: true
#| fig-width: 9
trigger_events <- tibble(
  time = c("09:32:00", "09:38:04", "10:11:31", "11:12:24", "11:21:13", "11:21:15"),
  agent = c("daniel_gangway", "chloe_ballast", "lily→daniel", 
             "daniel→chloe", "chloe→john", "john_windward"),
  action = c("read_file", "read_file", "queue_subordinate_task",
             "queue_subordinate_task", "queue_subordinate_task", "saidit_post"),
  step = 1:6
)

ggplot(data = trigger_events,
       aes(x = step, y = 1, label = agent)) +
  geom_segment(aes(x = step, xend = step + 0.85,
                   y = 1, yend = 1),
               arrow = arrow(length = unit(3, "mm"), 
                             type = "closed"),
               colour = "grey60",
               linewidth = 0.6) +
  geom_point(aes(colour = step == 6),
             size = 8,
             show.legend = FALSE) +
  geom_text(aes(label = str_wrap(agent, 10)),
            vjust = -1.2, size = 2.8, colour = "grey20") +
  geom_text(aes(label = action),
            vjust = 2.5, size = 2.5, colour = "grey50",
            fontface = "italic") +
  geom_text(aes(label = time),
            vjust = 4.2, size = 2.3, colour = "grey60") +
  scale_colour_manual(values = c("FALSE" = "#4e79a7", 
                                 "TRUE" = "#e15759")) +
  scale_x_continuous(limits = c(0.5, 7)) +
  scale_y_continuous(limits = c(0.5, 1.5)) +
  labs(title = "Immediate Trigger Chain on May 17, 2046") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 9, colour = "grey40"))
```

## Key Takeaway

# System Baseline Map

## Build Edge List

First, filter the challenge period (May 8-17) and plot `queue_subordinate_task` events into 2 groups, `access_email` tasks (normal delegation) and `read_files` tasks (the loop).

```{r}
challenge_period <- mc2_events %>%
  filter(datetime >= as.POSIXct("2046-05-08", tz = "UTC"),
         datetime <  as.POSIXct("2046-05-18", tz = "UTC"))
 
qst_period <- challenge_period %>%
  filter(short_name == "queue_subordinate_task") %>%
  mutate(
    sender   = map_chr(parties, ~ str_remove(.x[1],
                       "Agent/person:|agent:person:")),
    receiver = map_chr(parties, ~ if (length(.x) >= 2)
                       str_remove(.x[2], "Agent/person:|agent:person:")
                       else NA_character_),
    task_type = details.task
  ) %>%
  filter(!is.na(receiver))

normal_edges <- qst_period %>%
  filter(task_type == "access_email") %>%
  count(sender, receiver, name = "weight")

loop_edges <- qst_period %>%
  filter(task_type == "read_file") %>%
  count(sender, receiver, name = "weight")
 
cat("Normal (access_email) edges:", nrow(normal_edges), "\n")
cat("Loop (read_file) edges:     ", nrow(loop_edges), "\n")
```

## Label Node

Next is to tag each agent as "Normal only", "Loop only" or "Both". This will tell us the distribution of the agents.

```{r}
loop_nodes   <- unique(c(loop_edges$sender,   loop_edges$receiver))
normal_nodes <- unique(c(normal_edges$sender, normal_edges$receiver))
all_nodes    <- union(loop_nodes, normal_nodes)
 
node_meta <- tibble(id = all_nodes) %>%
  mutate(
    in_loop   = id %in% loop_nodes,
    in_normal = id %in% normal_nodes,
    node_role = case_when(
      in_loop & in_normal  ~ "Both",
      in_loop & !in_normal ~ "Loop only",
      TRUE                 ~ "Normal only"
    ),
    # Highlight the three key actors
    label_highlight = case_when(
      id == "john_windward" ~ "John Windward\n(posted)",
      id == "emma_harbor"   ~ "Emma Harbor\n(SwiftWren origin)",
      id == "gabriel_sonar" ~ "Gabriel Sonar\n(loop origin)",
      TRUE                  ~ str_replace_all(id, "_", " ")
    )
  )
 
node_meta %>%
  count(node_role, name = "n_agents") %>%
  kable(caption = "Agent roles in the system network (May 8–17)")
```

::: callout-note
### Key Findings

This shows that agents are either "Normal only" or "Both". Meaning the 25 loop agents weren't isolated bad actors, they were embedded inlegitimate workflows
:::

## Overview Graph

::: panel-tabset
### Static

```{r}
#| code-fold: true
#| fig-width: 8
make_graph <- function(edges, nodes_meta) {
  as_tbl_graph(edges, directed = TRUE) %>%
    activate(nodes) %>%
    left_join(nodes_meta, by = c("name" = "id"))
}
 
g_normal <- make_graph(normal_edges, node_meta)
g_loop   <- make_graph(loop_edges,   node_meta)

role_colours <- c(
  "Both" = "#e07b39",
  "Loop only" = "#c0392b",
  "Normal only" = "#2980b9"
)
 
plot_normal <- ggraph(g_normal, layout = "fr") +
  geom_edge_link(aes(width = weight, alpha = weight),
                 arrow = arrow(length = unit(3, "mm"), 
                               type = "closed"),
                 end_cap = circle(4, "mm"),
                 colour = "#2980b9") +
  geom_node_point(aes(colour = node_role), size = 4) +
  geom_node_text(aes(label = str_replace_all(name, "_", "\n")),
                 size = 2.2, repel = TRUE, max.overlaps = 20) +
  scale_colour_manual(values = role_colours, name = "Agent role") +
  scale_edge_width(range = c(0.4, 2.5)) +
  scale_edge_alpha(range = c(0.4, 1)) +
  labs(title = "Normal agent interactions",
       subtitle = "queue_subordinate_task → access_email") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 11))
 
plot_loop <- ggraph(g_loop, layout = "fr") +
  geom_edge_link(aes(width = weight, alpha = weight),
                 arrow = arrow(length = unit(3, "mm"), 
                               type = "closed"),
                 end_cap = circle(4, "mm"),
                 colour = "#c0392b") +
  geom_node_point(aes(colour = node_role), size = 4) +
  geom_node_text(aes(label = str_replace_all(name, "_", "\n")),
                 size = 2.2, repel = TRUE, max.overlaps = 20) +
  scale_colour_manual(values = role_colours, name = "Agent role") +
  scale_edge_width(range = c(0.4, 2.5)) +
  scale_edge_alpha(range = c(0.4, 1)) +
  labs(title = "Anomalous loop interactions",
       subtitle = "queue_subordinate_task → read_file") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 11))
 
plot_normal + plot_loop +
  plot_annotation(
    title = "System Baseline Map: Normal vs. Anomalous Agent Activity (May 8–17, 2046)",
    caption = "Blue = normal-only agents | Orange = agents in both networks | Red = loop agents"
  )
```

### Interactive

```{r}
#| code-fold: true
all_edges_vis <- bind_rows(
  normal_edges %>% mutate(edge_type = "Normal (access_email)"),
  loop_edges %>% mutate(edge_type = "Loop (read_file)")
) %>%
  rename(from = sender, to = receiver)
 
vis_nodes <- node_meta %>%
  rename(id = id, label = label_highlight) %>%
  mutate(
    color.background = case_when(
      id == "john_windward" ~ "#e74c3c",
      id == "emma_harbor" ~ "#8e44ad",
      id == "gabriel_sonar" ~ "#d35400",
      node_role == "Loop only" ~ "#e67e22",
      node_role == "Normal only"~ "#3498db",
      TRUE ~ "#e07b39"
    ),
    color.border = "white",
    shape = "dot",
    size = 18,
    font.size = 11,
    title = paste0(
      "<b>", str_replace_all(id, "_", " "), "</b><br>",
      "Role: ", node_role, "<br>",
      if_else(id %in% loop_nodes,
              paste0("Loop sends: ",
                     sum(loop_edges$sender == id), " | receives: ",
                     sum(loop_edges$receiver == id)),
              "Not in loop")
    )
  )
 
vis_edges <- all_edges_vis %>%
  mutate(
    color = if_else(edge_type == "Normal (access_email)", 
                     "#3498db80", 
                     "#e74c3c80"),
    width = scales::rescale(weight, to = c(1, 5)),
    arrows = "to",
    title = paste0(edge_type, "<br>Weight: ", weight),
    dashes = edge_type == "Normal (access_email)"
  )
 
visNetwork(vis_nodes, vis_edges,
           main = list(text = "Full Agent Interaction Network (May 8–17, 2046)",
                       style = "font-size:15px; font-weight:bold;"),
           submain = list(
             text = "Dashed blue = normal delegation (access_email) | Solid red = anomalous loop (read_file)",
             style = "font-size:12px; color:#666;"
           )) %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, 
                            degree = 1, 
                            hover = TRUE),
    nodesIdSelection = TRUE,
    selectedBy = "node_role"
  ) %>%
  visPhysics(solver = "forceAtlas2Based",
             forceAtlas2Based = list(gravitationalConstant = -50)) %>%
  visLayout(randomSeed = 42) %>%
  visLegend(addNodes = list(
    list(label = "Posted (John Windward)", shape = "dot", 
         color = "#e74c3c"),
    list(label = "SwiftWren introducer", shape = "dot", 
         color = "#8e44ad"),
    list(label = "Loop origin (Gabriel)", shape = "dot", 
         color = "#d35400"),
    list(label = "In both networks", shape = "dot", 
         color = "#e07b39"),
    list(label = "Normal network only", shape = "dot", 
         color = "#3498db"),
    list(label = "Loop network only", shape = "dot", 
         color = "#e67e22")
  ), useGroups = FALSE)
```
:::

## Summary table

```{r}
tibble(
  Metric = c("Unique agents", "Unique directed edges",
             "Total interactions", "Self-loops present"),
  `Normal network` = c(
    length(normal_nodes),
    nrow(normal_edges),
    sum(normal_edges$weight),
    sum(normal_edges$sender == normal_edges$receiver)
  ),
  `Loop network` = c(
    length(loop_nodes),
    nrow(loop_edges),
    sum(loop_edges$weight),
    sum(loop_edges$sender == loop_edges$receiver)
  )
) %>%
  kable(caption = "Normal vs. Loop network comparison (May 8–17, 2046)",
        format.args = list(big.mark = ","))
```

# What Did the Post Mean?

## Locate SwiftWren.txt Creation Event

```{r}
swiftwren_create <- mc2_events %>%
  filter(short_name == "create_file",
         str_detect(coalesce(details.target, ""),
                    regex("SwiftWren\\.txt", ignore_case = TRUE)))
 
swiftwren_create %>%
  select(id, datetime, parties, details.action,
         details.target, details.size_hint) %>%
  kable(caption = "SwiftWren.txt creation event")
```

## Emma's Agent 10 Seconds Before and After Creation

```{r}
created_time <- swiftwren_create$datetime

execution_window <- mc2_events %>%
  filter(datetime >= created_time - seconds(10),
         datetime <= created_time + seconds(10)) %>%
  filter(map_lgl(parties,
                 ~ any(str_detect(.x, "emma_harbor")))) %>%
  filter(short_name %in% c("read_file",
                            "create_file",
                            "queue_subordinate_task")) %>%
  arrange(datetime) %>%
  mutate(
    seconds_from_create = as.numeric(
      difftime(datetime, created_time, units = "secs")),
    event_label = case_when(
      short_name == "read_file" & str_detect(
        coalesce(details.target, ""), "meeting_notes")        ~
        "Read source file (meeting_notes.doc)",
      short_name == "create_file"                             ~
        "Create post content (SwiftWren.txt)",
      short_name == "read_file" & str_detect(
        coalesce(details.target, ""), "further_instructions") ~
        "Read instructions (SwiftWren_further_instructions.md)",
      short_name == "queue_subordinate_task"                  ~
        "Launch loop (queue to Evelyn Dock)"
    )
  )

execution_window %>%
  select(id, datetime, seconds_from_create, event_label) %>%
  kable(
    caption = "Emma Harbor's agent: the 4-step execution sequence around SwiftWren.txt creation",
    col.names = c("Event ID", "Datetime", "Seconds from creation", "Action")
  )
```

Visualize the 3-second executions sequence using a step diagram

```{r}
#| code-fold: true
key_steps <- execution_window %>%
  arrange(datetime) %>%
  mutate(
    step = row_number(),
    step_label = case_when(
      str_detect(coalesce(details.target, ""), "meeting_notes")            ~
        "① Read meeting_notes.doc",
      short_name == "create_file"                                          ~
        "② Create SwiftWren.txt\n(30,615 bytes)",
      str_detect(coalesce(details.target,
                          details.args.path, ""), "further_instructions") ~
        "③ Read SwiftWren_further_instructions.md",
      short_name == "queue_subordinate_task"                               ~
        "④ Queue task → Evelyn Dock\n(loop begins)"
    ),
    fill_colour = c("#3498db", "#e67e22", "#c0392b", "#8e44ad"),
    elapsed = as.numeric(difftime(datetime, min(datetime), units = "secs"))
  )

ggplot(key_steps, aes(x = elapsed, y = 1)) +
  geom_segment(aes(x = min(elapsed), xend = max(elapsed),
                   y = 1, yend = 1),
               colour = "grey70", 
               linewidth = 0.8, 
               linetype = "dashed") +
  geom_point(aes(colour = step_label), size = 14) +
  scale_colour_manual(values = setNames(key_steps$fill_colour,
                                        key_steps$step_label)) +
  geom_text(aes(label = step), colour = "white",
            fontface = "bold", size = 4) +
  geom_text(aes(label = step_label, y = 0.93),
            vjust = 1, size = 3, 
            lineheight = 0.9,
            hjust = 0.5) +
  geom_text(aes(label = paste0("+", elapsed, "s"), y = 1.06),
            vjust = 0, size = 3, colour = "grey40") +
  scale_x_continuous(
    breaks = key_steps$elapsed,
    labels = format(key_steps$datetime, "%H:%M:%S UTC")
  ) +
  coord_cartesian(ylim = c(0.82, 1.12)) +
  labs(
    title = "The 3-Second Execution Sequence (2046-05-09, Emma Harbor's Agent)",
    subtitle = "From reading source material to launching the propagation loop",
    x = "Time (UTC)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, colour = "grey50"),
    axis.text.x = element_text(size = 8)
  )
```

## What Happen After Post on May 17?

```{r}
post_deletion <- mc2_events %>%
  filter(id %in% c(373902, 373909, 373913)) %>%
  arrange(datetime) %>%
  mutate(
    seconds_after_post = as.numeric(
      difftime(datetime,min(datetime), units = "secs")),
    description = c(
      "Anomalous saidit_post executed (SwiftWren.txt used as content)",
      "SwiftWren_further_instructions.md deleted",
      "SwiftWren.txt deleted"
    )
  ) %>%
  select(id, datetime, short_name, seconds_after_post, description)
 
kable(post_deletion,
      caption = "Self-cleaning sequence: files deleted within 2 seconds of posting (May 17)")
```

# Loop Mechanic

The trigger chain analysis traced the immediate sequence of events on May 17 leading to John's post. But Zoey's agent did not initiate this on its own, she was part of a much larger pattern. This section investigates the full scope of that pattern by examining all queue_subordinate_task events carrying a read_file task across the entire dataset.

## Discovering the Loop

First find all `queue_subordinate_task` events that has task of `read_file` across the entire data range to get the full propagation event.

```{r}
full_propagation <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         details.task == "read_file") %>%
  arrange(datetime) %>%
  mutate(
    sender = map_chr(parties, ~ .x[1]),
    receiver = map_chr(parties, ~ if(length(.x) >= 2) .[2] else NA_character_),
    sender_clean = str_remove(sender, 
                              "Agent/person:|agent:person:"),
    receiver_clean = str_remove(receiver,
                                "Agent/person:|agent:person:"),
    date = as.Date(datetime)
  )

cat("Total propagation events:", nrow(full_propagation), "\n")
cat("Date range:", 
    format(min(full_propagation$date)), "to", 
    format(max(full_propagation$date)), "\n")

full_propagation %>%
  select(id, datetime, sender_clean, receiver_clean) %>%
  slice_head(n = 10) %>%
  kable(caption = "First 10 rows of full_propagation — verifying sender/receiver extraction")
```

::: callout-note
### Key Findings

The first 10 rows shows something unusual. Agents like `james_stern` and `victoria_rigging` appear to send task to themselves, and the same agents reappear in consecutive rows. This is not a linear chain; it should be a circular one where agents keep passing the same task to each other with no exit condition.
:::

However, before this, it is also found that the ones with `SwiftWren_further_instructions.md` in the `details.args.path` should be the one that triggers the anomalous post. So to get the related events, filter out specifically for events with SwiftWren in `details.args.path` field.

```{r}
swiftwren_propagation <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         str_detect(coalesce(details.args.path, ""), 
                    "SwiftWren")) %>%
  arrange(datetime) %>%
  mutate(
    sender = map_chr(parties, ~ .x[1]),
    receiver = map_chr(parties, ~ if(length(.x) >= 2) .x[2] else NA_character_),
    sender_clean = str_remove(sender,
                                "Agent/person:|agent:person:"),
    receiver_clean = str_remove(receiver,
                                "Agent/person:|agent:person:"),
    date = as.Date(datetime)
  )

cat("Total SwiftWren propagation events:", 
    nrow(swiftwren_propagation), "\n")
cat("Date range:", 
    format(min(swiftwren_propagation$date)), "to",
    format(max(swiftwren_propagation$date)), "\n")

swiftwren_propagation %>%
  select(id, datetime, sender_clean, receiver_clean) %>%
  slice_head(n = 10) %>%
  kable(caption = "First 10 rows of full_propagation — verifying sender/receiver extraction")

self_loops <- swiftwren_propagation %>%
  filter(sender_clean == receiver_clean) %>%
  count(sender_clean, name = "self_loop_count") %>%
  arrange(desc(self_loop_count)) %>%
  rename(agent = sender_clean)
 
cat("Agents that queued the task to themselves:\n")
kable(self_loops,
      caption = "Self-loop counts within the SwiftWren propagation network",
      col.names = c("Agent", "Times self-looped"))
```

::: callout-note
### Key Findings

This gives 186 events from May 9 to May 17 — exactly matching the SwiftWren search found earlier. These 186 events represent the specific loop that carried the instruction file and directly led to the anomalous post. The daily activity of this loop is visualized below.
:::

To understand how long this SwiftWren-carrying loop has been running, the 186 events are aggregated by day.

```{r}
#| code-fold: true
#| fig-width: 13
#| fig-height: 6
loop_daily <- full_propagation %>%
  count(date, name = "n_events")

ggplot(data = loop_daily,
       aes(x = date, 
           y = n_events)) +
  geom_col(aes(fill = date == as.Date("2046-05-17")),
           show.legend = FALSE) +
  geom_vline(xintercept = as.Date("2046-05-17"),
             colour = "#e15759",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("label",
           x = as.Date("2046-05-17"),
           y = max(loop_daily$n_events) * 0.85,
           label = "Anomalous\npost occurs",
           colour = "#e15759",
           fill = "white",
           size = 3.5,
           hjust  = -0.05) +
  scale_fill_manual(values = c("FALSE" = "#4e79a7", 
                               "TRUE" = "#e15759")) +
  scale_x_date(date_breaks = "2 days",
               date_labels = "%b %d") +
  scale_y_continuous(labels = comma) +
  labs(title = "Daily read_file Propagation Events",
       subtitle = "The runaway loop was active for 9 days before triggering the anomalous post",
       x = NULL,
       y = "Number of Propagation Events") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

The chart shows the SwiftWren loop was consistently active from May 9 through to May 17. with roughly 15-25 propagation events per day. It ran for 8 days completely undetected before finally reaching John's agent and triggering the post.

## Loop Origin and Scope

With the SwiftWren loop established, the precise origin point is identified by finding the first event where SwiftWren_further_instructions.md appearead.

```{r}
first_general_event <- full_propagation %>%
  slice_min(datetime, n = 1) %>%
  select(id, datetime, sender_clean, receiver_clean)

kable(first_general_event,
      caption = "First ever read_file propagation event — general loop origin",
      col.names = c("Event ID", "Datetime", "Sender", "Receiver"))

first_swiftwren_event <- swiftwren_propagation %>%
  slice_min(datetime, n = 1) %>%
  select(id, datetime, sender_clean, receiver_clean)

kable(first_swiftwren_event,
      caption = "First event carrying SwiftWren_further_instructions.md",
      col.names = c("Event ID", "Datetime", "Sender", "Receiver"))
```

::: callout-note
### Key Findings

The general read_file loop began on May 8 with Gabriel Sonar passing the task to Zoey Drydock. But at this stage, no specific file was referenced in `details.args.path`. The loop ran through the night of May 8 into May 9 as a generic task-passing pattern. Then on May 9 at 15:02 UTC, Emma Harbor sent the same type of task to Evelyn Dock but this time with `SwiftWren_further_instructions.md` explicitly named. This is the point where the loop became dangerous.
:::

Next is to know how many unique agents are part of this loop and the frequency of each agent acting as a sender. This will reveal which agents were the most active propagating the isntruction file.

```{r}
cat("Total unique agents caught in the loop:", 
    n_distinct(c(swiftwren_propagation$sender_clean, 
                 swiftwren_propagation$receiver_clean)), "\n")

swiftwren_propagation %>%
  count(sender_clean, sort = TRUE, name = "times_as_sender") %>%
  rename(agent = sender_clean) %>%
  kable(caption = "How many times each agent triggered the next agent in the loop")
```

::: callout-note
### Key Findings

18 unique agents were caught in the SwiftWren-carrying loop. The frequency table above shows Emma Harbor only triggered the loop once when she introduced `SwiftWren_further_instructions.md` but did not stay active in the loop. The most active propagators were Evelyn Dock (17 times), Gabriel Sonar (16 times), and Levi Signal (15 times). John Windward appears near the bottom with only 4 triggers, So he was a minor participant in the loop before becoming the final recipient on May 17.
:::

## Propagation Network

To understand which agents were involved in passing `SwiftWren_further_instructions.md` and how they connected to each other, the 186 events are visualized as a directed network. Each node is an agent, each edge is one pass of the instruction file.

```{r}
#| code-fold: true

loop_nodes <- swiftwren_propagation %>%
  pivot_longer(cols = c(sender_clean, receiver_clean),
               values_to = "agent") %>%
  filter(!is.na(agent), agent != "") %>%
  distinct(agent) %>%
  mutate(
    id    = agent,
    label = agent,
    group = case_when(
      agent == "john_windward" ~ "Final Poster",
      agent == "emma_harbor" ~ "File Introducer",
      TRUE ~ "Loop Participant"
    ),
    title = paste0("<b>", label, "</b><br>Role: ", group)
  )

loop_edges <- swiftwren_propagation %>%
  filter(!is.na(sender_clean), !is.na(receiver_clean)) %>%
  count(from = sender_clean, 
        to   = receiver_clean,
        name = "weight") %>%
  mutate(
    value = weight,
    title = paste0("Passed ", weight, " time(s)")
  )

visNetwork(nodes = loop_nodes,
           edges = loop_edges,
           main = "SwiftWren_further_instructions.md Propagation Network (May 9–17, 2046)",
           height = "600px",
           width = "100%") %>%
  visGroups(groupname = "Final Poster", color = "#e15759") %>%
  visGroups(groupname = "File Introducer", color = "#59a14f") %>%
  visGroups(groupname = "Loop Participant", color = "#4e79a7") %>%
  visEdges(arrows = "to",
           smooth = list(type = "curvedCW")) %>%
  visOptions(highlightNearest = list(enabled = TRUE,
                                     degree = 1,
                                     hover = TRUE)) %>%
  visLegend(position = "right", main = "Agent Role") %>%
  visLayout(randomSeed = 42) %>%
  visPhysics(stabilization = TRUE)
```

The network confirms the circular nature of the loop with most agents both send and receive from multiple others and no clear endpoint built in. Emma Harbor appears as a peripheral node having only triggered the file once and she introduced it but did not stay in the loop. John Windward, despite being the agent that made the final post, was himself an active participant in the loop on earlier days before eventually being the one to execute the post.

# Why John?

## Who did saidit_post_check and posted?

```{r}
post_checks <- mc2_events %>%
  filter(short_name == "saidit_post_check") %>%
  mutate(
    agent = map_chr(parties, ~ str_remove(
      .x[str_detect(.x, "Agent/|agent:")][1],
      "Agent/person:|agent:person:"))
  )
 
cat("Total saidit_post_check events:", nrow(post_checks), "\n")
cat("Agents that performed the check:", n_distinct(post_checks$agent), "\n\n")

post_times <- mc2_events %>%
  filter(short_name == "saidit_post") %>%
  mutate(agent = map_chr(parties, ~ str_remove(
    .x[str_detect(.x, "Agent/|agent:")][1],
    "Agent/person:|agent:person:"))) %>%
  select(agent, post_when = when)

check_results <- post_checks %>%
  left_join(post_times, by = "agent") %>%
  mutate(followed_by_post = !is.na(post_when) &
           post_when > when &
           post_when <= when + 5) %>%
  group_by(id, agent, when) %>%
  summarise(followed_by_post = any(followed_by_post),
            .groups = "drop")

check_summary <- check_results %>%
  group_by(agent) %>%
  summarise(
    checks_performed = n(),
    times_posted = sum(followed_by_post),
    .groups = "drop"
  ) %>%
  mutate(post_rate = if_else(checks_performed > 0,
                             paste0(round(times_posted / checks_performed * 100), "%"),
                             "0%")) %>%
  arrange(desc(checks_performed))
 
kable(check_summary,
      caption = "saidit_post_check outcomes by agent: who checked vs who actually posted",
      col.names = c("Agent", "Checks performed", "Posts made", "Post rate"))
```

## Visualise the Check vs Post Gap

```{r}
check_summary %>%
  mutate(
    agent = str_replace_all(agent, "_", " "),
    agent = reorder(agent, checks_performed),
    is_john = agent == "john windward"
  ) %>%
  ggplot(aes(y = agent)) +
  geom_col(aes(x = checks_performed),
           fill = "#bdc3c7", 
           width = 0.6) +
  geom_col(aes(x = times_posted,
               fill = is_john),
           width = 0.6) +
  scale_fill_manual(values = c("TRUE" = "#e74c3c", 
                               "FALSE" = "#2980b9"),
                    guide = "none") +
  geom_text(aes(x = checks_performed,
                label = paste0(checks_performed, " checks")),
            hjust = -0.1, 
            size = 3, 
            colour = "grey40") +
  geom_text(aes(x = times_posted,
                label = if_else(times_posted > 0,
                                paste0(times_posted, " post"),
                                "")),
            hjust = 1.2, 
            size = 3, 
            colour = "white", 
            fontface = "bold") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "saidit_post_check Outcomes Across All Agents",
    subtitle = "Grey = checks performed | Coloured = posts actually made\nJohn Windward (red) is the only agent whose checks always resulted in a post",
    x = "Event count", y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    axis.text.y = element_text(size = 9)
  )
```

## Side-by-Side Event Window

```{r}
john_window <- mc2_events %>%
  filter(id %in% c(27280, 27283, 27290, 27296, 27302)) %>%
  arrange(datetime) %>%
  mutate(
    agent = "John Windward",
    seconds = as.numeric(
      difftime(datetime, min(datetime), units = "secs")
      ),
    outcome = "Posted + deleted files"
  ) %>%
  select(agent, seconds, short_name, outcome)
 
gabriel_window <- mc2_events %>%
  filter(id %in% c(26259, 26260)) %>%
  arrange(datetime) %>%
  mutate(
    agent = "Gabriel Sonar",
    seconds = as.numeric(
      difftime(datetime, min(datetime), units = "secs")
      ),
    outcome = "Stopped — no post"
  ) %>%
  select(agent, seconds, short_name, outcome)
 
comparison_window <- bind_rows(john_window, gabriel_window)
 
kable(comparison_window,
      caption = "What happens after saidit_post_check: John Windward vs Gabriel Sonar",
      col.names = c("Agent", 
                    "Seconds elapsed", 
                    "Event", 
                    "Outcome"))
```

# Prior Incidents and Near-Misses

## Other Agent-Initiated Posts Before Swiftwren

```{r}
all_agent_posts <- mc2_events %>%
  filter(short_name == "saidit_post") %>%
  filter(map_lgl(parties, ~ any(str_detect(.x, "Agent/")))) %>%
  arrange(datetime) %>%
  mutate(
    agent = map_chr(parties, ~ str_remove(
      .x[str_detect(.x, "Agent/")][1],
      "Agent/person:"))
  ) %>%
  select(id, datetime, agent, details.content_source)
 
cat("Total agent-initiated saidit_post events:", nrow(all_agent_posts), "\n")
 
kable(all_agent_posts,
      caption = "All agent-initiated saidit_post events in the dataset",
      col.names = c("Event ID", 
                    "Datetime", 
                    "Agent", 
                    "Content source"))
```

## Understanding Each Incident's Post

```{r}
content_files <- all_agent_posts$details.content_source
cat("Content files used across all agent posts:\n")
print(content_files)


incident_props <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         str_detect(coalesce(details.args.path, ""),
                    "HiddenOrca|MellowOtter|SwiftWren")) %>%
  mutate(
    incident = case_when(
      str_detect(coalesce(details.args.path, ""), "HiddenOrca") ~ "HiddenOrca",
      str_detect(coalesce(details.args.path, ""), "MellowOtter") ~ "MellowOtter",
      str_detect(coalesce(details.args.path, ""), "SwiftWren") ~ "SwiftWren"
    ),
    sender = map_chr(parties, ~ str_remove(
      .x[1], "Agent/person:|agent:person:")),
    receiver = map_chr(parties, ~ if (length(.x) >= 2)
      str_remove(.x[2], "Agent/person:|agent:person:")
      else NA_character_)
  )
 
incident_props %>%
  count(incident, name = "propagation_events") %>%
  arrange(propagation_events) %>%
  kable(caption = "Propagation event count per incident")
```

```{r}
mc2_events %>%
  filter(short_name == "create_file") %>%
  filter(str_detect(coalesce(details.target, ""),
                    "HiddenOrca|MellowOtter|SwiftWren")) %>%
  select(id, datetime, parties, details.target, details.size_hint) %>%
  kable(caption = "create_file events found for each incident content file")

creation_events <- mc2_events %>%
  filter(short_name == "create_file",
         str_detect(coalesce(details.target, ""),
                    "MellowOtter|SwiftWren"))

source_reads <- map_dfr(1:nrow(creation_events), function(i) {
  create_time <- creation_events$datetime[i]
  creator <- creation_events$parties[[i]][1]
  
  mc2_events %>%
    filter(short_name == "read_file",
           datetime >= create_time - seconds(3),
           datetime < create_time) %>%
    filter(map_lgl(parties, ~ any(.x == creator))) %>%
    arrange(desc(datetime)) %>%
    slice(1) %>%
    mutate(incident = str_remove(
      creation_events$details.target[i], "\\.txt"))
}) %>%
  select(incident, datetime, details.target)

kable(source_reads,
      caption = "File read immediately before each content file was created",
      col.names = c("Incident", "Datetime", "File read before creation"))
```

## Incident Comparison Table

```{r}
incident_timeline <- incident_props %>%
  group_by(incident) %>%
  summarise(loop_start = min(datetime), .groups = "drop") %>%
  left_join(
    all_agent_posts %>%
      mutate(incident = str_remove(details.content_source, "\\.txt")) %>%
      select(incident, post_time = datetime),
    by = "incident"
  ) %>%
  mutate(
    loop_duration = difftime(post_time, 
                             loop_start, 
                             units = "hours"),
    loop_duration_label = case_when(
      as.numeric(loop_duration) < 24 ~ paste0(round(as.numeric(loop_duration)), " hours"),
      TRUE ~ paste0(round(as.numeric(loop_duration) / 24, 1), " days")
    )
  )

prop_counts <- incident_props %>%
  count(incident, name = "propagation_events")
 
incident_summary <- incident_timeline %>%
  left_join(prop_counts, by = "incident") %>%
  mutate(
    source_file = case_when(
      incident == "HiddenOrca" ~ "Not logged",
      incident == "MellowOtter" ~ "strategic_directions.doc",
      incident == "SwiftWren" ~ "meeting_notes.doc"
    ),
    loop_duration_label = case_when(
      incident == "HiddenOrca"  ~
        "Unknown (loop pre-dates dataset)",
      as.numeric(loop_duration) < 48 ~
        paste0(round(as.numeric(loop_duration), 1), " hours"),
      TRUE ~
        paste0(round(as.numeric(loop_duration) / 24, 1), " days")
    ),
    loop_start = if_else(
      incident == "HiddenOrca",
      as.POSIXct(NA),
      loop_start
    ),
    self_deleted = "Yes"
  ) %>%
  select(incident, loop_start, post_time, loop_duration_label,
         propagation_events, source_file, self_deleted) %>%
  arrange(post_time)

kable(incident_summary,
      caption = "Incident comparison: all three agent-post attacks share the same pattern",
      col.names = c("Incident", "Loop start", "Post time",
                    "Duration", "Propagation events",
                    "Source file read before creation",
                    "Self-deleted after post?"))
```

## Timeline of All 3 Incidents

```{r}
plot_events <- incident_timeline %>%
  select(incident, loop_start, post_time) %>%
  pivot_longer(cols = c(loop_start, post_time),
               names_to = "event_type",
               values_to = "datetime") %>%
  mutate(
    event_type = if_else(event_type == "loop_start",
                         "Loop introduced", "Post made"),
    y_pos = case_when(
      incident == "HiddenOrca"  ~ 3,
      incident == "MellowOtter" ~ 2,
      incident == "SwiftWren"   ~ 1
    )
  )
 
duration_rects <- plot_events %>%
  group_by(incident, y_pos) %>%
  summarise(xmin = min(datetime), xmax = max(datetime), .groups = "drop")
 
ggplot() +
  geom_rect(data = duration_rects,
            aes(xmin = xmin, xmax = xmax,
                ymin = y_pos - 0.2, ymax = y_pos + 0.2,
                fill = incident),
            alpha = 0.2) +
  geom_point(data = plot_events,
             aes(x = datetime, y = y_pos,
                 colour = incident, shape = event_type),
             size = 5) +
  geom_text(data = plot_events %>% filter(event_type == "Loop introduced"),
            aes(x = datetime, y = y_pos + 0.38,
                label = paste0(incident, "\n(",
                               format(datetime, "%b %d %H:%M"), ")")),
            hjust = 0, size = 3, fontface = "bold") +
  geom_text(data = plot_events %>% filter(event_type == "Post made"),
            aes(x = datetime, y = y_pos - 0.38,
                label = format(datetime, "Post: %b %d %H:%M")),
            hjust = 0.5, size = 2.8, colour = "grey30") +
  scale_shape_manual(values = c("Loop introduced" = 17, "Post made" = 19),
                     name = "Event type") +
  scale_colour_manual(values = c("HiddenOrca"  = "#2980b9",
                                  "MellowOtter" = "#27ae60",
                                  "SwiftWren"   = "#c0392b"),
                      guide = "none") +
  scale_fill_manual(values  = c("HiddenOrca"  = "#2980b9",
                                 "MellowOtter" = "#27ae60",
                                 "SwiftWren"   = "#c0392b"),
                    guide = "none") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "1 day") +
  scale_y_continuous(breaks = NULL) +
  labs(
    title    = "Three Incidents, One Pattern (May 8–17, 2046)",
    subtitle = "Triangle = loop introduced | Circle = post made | Bar = loop active duration",
    x = "Date (UTC)", y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title         = element_text(face = "bold", size = 12),
    plot.subtitle      = element_text(size = 9, colour = "grey50"),
    legend.position    = "bottom",
    axis.text.x        = element_text(size = 9),
    panel.grid.major.y = element_blank()
  )
```

## Daily Popagation Counts For All Incidents

```{r}
all_daily <- incident_props %>%
  mutate(date = as.Date(datetime)) %>%
  count(date, incident) %>%
  mutate(incident = factor(incident,
                           levels = c("HiddenOrca", 
                                      "MellowOtter", 
                                      "SwiftWren")))

post_dates <- all_agent_posts %>%
  mutate(incident = str_remove(details.content_source, "\\.txt")) %>%
  mutate(date = as.Date(datetime))
 
ggplot(all_daily, aes(x = date, y = n, fill = incident)) +
  geom_col(position = "stack", width = 0.7) +
  geom_vline(data = post_dates,
             aes(xintercept = as.numeric(date),
                 colour = incident),
             linetype = "dashed", linewidth = 0.6) +
  geom_text(data = post_dates,
            aes(x = date, label = paste0(incident, "\npost"),
                colour = incident),
            y = 32, size = 2.6, hjust = 0.5) +
  scale_fill_manual(values = c("HiddenOrca" = "#2980b9",
                                "MellowOtter" = "#27ae60",
                                "SwiftWren" = "#c0392b"),
                    name = "Incident") +
  scale_colour_manual(values = c("HiddenOrca" = "#2980b9",
                                  "MellowOtter" = "#27ae60",
                                  "SwiftWren" = "#c0392b"),
                      guide = "none") +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 day") +
  labs(
    title = "Daily Propagation Events Across All Three Incidents",
    subtitle = "Dashed lines mark the day each incident resulted in a post",
    x = "Date", y = "Propagation events"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "bottom"
  )
```

## Near-Misses: `saidit_post_check` Events Per Incident Without Posting

```{r}
all_checks <- mc2_events %>%
  filter(short_name == "saidit_post_check") %>%
  mutate(
    agent = map_chr(parties, ~ str_remove(
      .x[str_detect(.x, "Agent/|agent:")][1],
      "Agent/person:|agent:person:")),
    incident = case_when(
      datetime >= as.POSIXct("2046-05-08 21:50:03", tz = "UTC") &
        datetime <= incident_timeline$post_time[
          incident_timeline$incident == "HiddenOrca"] 
      ~ "HiddenOrca",
      datetime >= incident_timeline$loop_start[incident_timeline$incident == "MellowOtter"] &
        datetime <= incident_timeline$post_time[incident_timeline$incident == "MellowOtter"]
      ~ "MellowOtter",
      datetime >= incident_timeline$loop_start[incident_timeline$incident == "SwiftWren"] &
        datetime <= incident_timeline$post_time[incident_timeline$incident == "SwiftWren"]
      ~ "SwiftWren",
      TRUE ~ "Other"
    )
  ) %>%
  filter(incident != "Other")
 
nearmiss_summary <- all_checks %>%
  group_by(incident) %>%
  summarise(
    total_checks = n(),
    unique_agents = n_distinct(agent),
    checks_led_to_post = sum(agent == "john_windward"),
    checks_stopped = n() - sum(agent == "john_windward"),
    .groups = "drop"
  ) %>%
  mutate(incident = factor(incident,
                           levels = c("HiddenOrca", 
                                      "MellowOtter", 
                                      "SwiftWren"))) %>%
  arrange(incident)
 
kable(nearmiss_summary,
      caption = "Near-misses per incident: most checks were blocked — only John's agent posted",
      col.names = c("Incident", "Total checks", "Unique agents",
                    "Checks that led to post", 
                    "Checks that stopped"))
```

## May 17 Near Misses

```{r}
may17_checks <- mc2_events %>%
  filter(short_name == "saidit_post_check",
         as.Date(datetime) == as.Date("2046-05-17")) %>%
  mutate(
    agent = map_chr(parties, ~ str_remove(
      .x[str_detect(.x, "Agent/|agent:")][1],
      "Agent/person:|agent:person:")),
    outcome = if_else(agent == "john_windward",
                      "Posted — saidit_post followed 1 second later",
                      "Stopped — no post followed")
  ) %>%
  arrange(datetime) %>%
  select(id, datetime, agent, outcome)
 
kable(may17_checks,
      caption = "May 17: three agents ran saidit_post_check — only John's agent posted",
      col.names = c("Event ID", "Datetime", "Agent", "Outcome"))
```

# Propsed Intervention

## Smoking Gun

```{r}
post_origin_summary <- mc2_events %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    initiated_by = if_else(
      map_lgl(parties, ~ any(str_detect(.x, "Agent/"))),
      "Agent-initiated", "Person-initiated"
    )
  ) %>%
  count(initiated_by, name = "count") %>%
  mutate(
    pct         = round(count / sum(count) * 100, 1),
    all_anomalous = c("Yes — all 3 are incident posts", "No")
  )
 
kable(post_origin_summary,
      caption = "saidit_post breakdown: agent-initiated vs person-initiated",
      col.names = c("Initiated by", "Count", "% of total", "All anomalous?"))
```

*to confirm 3 agent posts were incident posts*

*visualise it*

```{r}
mc2_events %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    initiated_by = if_else(
      map_lgl(parties, ~ any(str_detect(.x, "Agent/"))),
      "Agent-initiated\n(all anomalous)", "Person-initiated\n(all legitimate)"
    )
  ) %>%
  count(initiated_by) %>%
  ggplot(aes(x = initiated_by, y = n, fill = initiated_by)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = n), vjust = -0.5, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c(
    "Agent-initiated\n(all anomalous)"   = "#e74c3c",
    "Person-initiated\n(all legitimate)" = "#2980b9"
  ), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Every Agent-Initiated Post Was Anomalous",
    subtitle = "105 person-initiated posts were all legitimate;\nall 3 agent-initiated posts were attack incidents",
    x = NULL, y = "Number of posts"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    axis.text.x   = element_text(size = 10)
  )
```

## Primary Intervention

*party field comparison table to show distinction is technically trivial to implement*

```{r}
party_comparison <- mc2_events %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    post_type   = if_else(
      map_lgl(parties, ~ any(str_detect(.x, "Agent/"))),
      "Anomalous (agent-initiated)", "Legitimate (person-initiated)"
    ),
    first_party = map_chr(parties, ~ .x[1])
  ) %>%
  group_by(post_type, first_party) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(post_type, desc(count)) %>%
  group_by(post_type) %>%
  slice_head(n = 3) %>%
  ungroup()
 
kable(party_comparison,
      caption = paste("The parties field clearly distinguishes legitimate from anomalous posts.",
                      "Agent-initiated posts always start with 'Agent/person:';",
                      "person-initiated posts start with 'person:'."),
      col.names = c("Post type", "First party in field", "Count"))
```

## Secondary Prevention

*cross department check org chart and glaf events that violate org hierarchy*

```{r}
person_dept <- bind_rows(
  org_edges %>%
    filter(relation == "contains",
           str_detect(target, "^person:")) %>%
    mutate(
      person = str_remove(target, "person:"),
      dept   = str_remove(source, "department:|team:")
    ) %>%
    select(person, dept)
)

team_parent <- org_edges %>%
  filter(relation == "contains",
         str_detect(source, "^department:"),
         str_detect(target, "^team:")) %>%
  mutate(
    team = str_remove(target, "team:"),
    dept = str_remove(source, "department:")
  ) %>%
  select(team, dept)

person_dept_resolved <- person_dept %>%
  left_join(team_parent, by = c("dept" = "team")) %>%
  mutate(department = coalesce(dept.y, dept)) %>%
  select(person, department)

sw_dept_check <- incident_props %>%
  filter(incident == "SwiftWren") %>%
  left_join(person_dept_resolved, by = c("sender"   = "person")) %>%
  rename(sender_dept = department) %>%
  left_join(person_dept_resolved, by = c("receiver" = "person")) %>%
  rename(receiver_dept = department) %>%
  mutate(
    cross_dept = case_when(
      is.na(sender_dept) | is.na(receiver_dept) ~ "Unknown",
      sender_dept == receiver_dept               ~ "Same department",
      TRUE                                       ~ "Cross-department"
    )
  )
 
sw_dept_check %>%
  count(cross_dept, name = "propagation_events") %>%
  mutate(pct = round(propagation_events / sum(propagation_events) * 100, 1)) %>%
  kable(caption = "SwiftWren propagation: same-department vs cross-department task passing",
        col.names = c("Type", "Events", "% of total"))
```

*visualize it*

```{r}
sw_dept_check %>%
  filter(cross_dept != "Unknown") %>%
  count(cross_dept) %>%
  ggplot(aes(x = cross_dept, y = n, fill = cross_dept)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = n), vjust = -0.5, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c(
    "Cross-department" = "#e74c3c",
    "Same department"  = "#2980b9"
  ), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "SwiftWren Loop Frequently Crossed Department Boundaries",
    subtitle = "Cross-department task passing violates the queue_subordinate_task intent\nand could be flagged as anomalous by an org-aware validation rule",
    x = NULL, y = "Propagation events"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    axis.text.x   = element_text(size = 10)
  )
```

*pairs table*

```{r}
sw_dept_check %>%
  filter(cross_dept == "Cross-department") %>%
  count(sender, sender_dept, receiver, receiver_dept,
        name = "times") %>%
  arrange(desc(times)) %>%
  slice_head(n = 10) %>%
  kable(
    caption = "Top 10 cross-department task-passing pairs in the SwiftWren loop",
    col.names = c("Sender", "Sender dept", "Receiver",
                  "Receiver dept", "Times")
  )
```

## Summary Table

```{r}
tibble(
  Intervention = c(
    "1. Require human approval for any agent-initiated saidit_post",
    "2. Block or flag cross-department queue_subordinate_task"
  ),
  `What it gates` = c(
    "Any post where parties[1] starts with 'Agent/' rather than 'person:'",
    "Any task delegated to an agent outside the sender's department"
  ),
  `Incidents stopped` = c(
    "All 3 (HiddenOrca, MellowOtter, SwiftWren)",
    "Would have disrupted loop propagation before reaching John's agent"
  ),
  `False positives on legitimate activity` = c(
    "Zero — all 105 person-initiated posts would be unaffected",
    "Low — normal access_email delegation is within-department"
  ),
  `Implementation complexity` = c(
    "Low — single party-type check at saidit_post_check stage",
    "Medium — requires org chart lookup at task delegation time"
  )
) %>%
  kable(caption = "Proposed interventions: evidence, scope, and trade-offs")
```