---
title: "Take-Home Exercise 2 - VAST Mini Challenge 2"
author: "Amelia Rosalind"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

# Getting Started

## Installing of required packages

```{r}
pacman::p_load(tidyverse, jsonlite, igraph, tidygraph, 
               ggraph, visNetwork, plotly, DT,
               gganimate, tidytext, stringdist, knitr)
```

## Importing data

```{r}
mc2_raw <- fromJSON("mc2_data/MC2_data.json", flatten = TRUE)
org_raw <- fromJSON("mc2_data/org_chart.json", flatten = TRUE)
```

# Inspect MC2 Raw Data

## Inspect structure

```{r}
str(mc2_raw, max.level = 2)
names(mc2_raw)
```

## Inspect desciption and events only
```{r}
mc2_desc <- mc2_raw$description
mc2_events <- as_tibble(mc2_raw$events)
```

## Parse timestamp into readable datetime
```{r}
mc2_events <- mc2_events %>%
  mutate(datetime = as_datetime(when))
```

## Quick sanity checks
```{r}
glimpse(mc2_events)
range(mc2_events$datetime)
n_distinct(mc2_events$short_name)
mc2_events %>% count(short_name, sort = TRUE) %>% print (n = 24)
```

## Find the anomalous SaidIT post
```{r}
saidit_post <- mc2_events %>%
  filter(!is.na(details.forum)) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, details.timestamp)

print(saidit_post)
```

## Confirm John Windwar's event on May 17
```{r}
john_may17 <- mc2_events %>%
  filter(map_lgl(parties, ~ "person:john_windward" %in% .x)) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, details.from, details.to,
         details.content, details.forum, details.poster_id,
         details.a2a.from, details.a2a.to, details.a2a.action)

datatable(john_may17,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "John Windward's Events on May 17, 2046")
```

## Narrow down to specific anomalous post
```{r}
anomalous_post <- mc2_events %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  filter(as.Date(datetime) == as.Date("2046-5-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, 
         details.content_source)

datatable(anomalous_post, 
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "SaidIT Posts on May 17, 2046")
```

::: callout-note
### Key Finding

The anomalous post is event ID 373902, posted at 2046-05-17T11:21:15Z but this is in UTC which in US Eastern time is on 4.21am, which matches the challenge description. In addition `details.forum`:`general` coincide with the challenge description that its posted in a random forum. 
:::

## Find difference between saidit_post and post_saidit
```{r}
bind_rows(
  mc2_events %>% filter(short_name == "saidit_post") %>% head(5),
  mc2_events %>% filter(short_name == "post_saidit") %>% head(5)
) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content,
         details.content_source) %>%
  datatable(options = list(scrollX = TRUE, pageLength = 5),
           caption = "Comparing saidit_post vs post_saidit action types")
```

::: callout-note
### Key Findings

`saidit_post` events show `parties` as `person:...` and `post_saidit` events show `parties` as `Agent/person:...,system:saidit`, which means `saidit_post` should be posted by the person themselves and `post_saidit` are posted by an agent on behalf of the person. 
:::

## Conclusion on the anomalous post

The post has `short_name` as `saidit_post` but under `parties` as `Agent/person:john_windward,system:saidit` which is weird as concluded that `saidit_post` should be posted by the person so John himself. But `parties` shows that its posted by John's agent instead. 

In addition based on John's event on May 17, there was only `check_email`. He was also not active on 4:21 am, which means his agent acted autonomously.

# Inspect 


