---
title: "Take-Home Exercise 2 - VAST Mini Challenge 2"
author: "Amelia Rosalind"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

# Getting Started

## Installing of required packages

```{r}
pacman::p_load(tidyverse, jsonlite, igraph, tidygraph, 
               ggraph, visNetwork, plotly, DT,
               gganimate, tidytext, stringdist, knitr,
               ggrepel, patchwork, ggthemes, scales)
```

## Importing data

```{r}
mc2_raw <- fromJSON("mc2_data/MC2_data.json", flatten = TRUE)
org_raw <- fromJSON("mc2_data/org_chart.json", flatten = TRUE)
```

## Inspect structure

```{r}
str(mc2_raw, max.level = 2)
names(mc2_raw)

str(org_raw, max.level = 2)
names(org_raw)
```

# Understanding the Main Data

## Create table for desciption and events

```{r}
mc2_desc <- mc2_raw$description
mc2_events <- as_tibble(mc2_raw$events)
```

## Parse timestamp into readable datetime

```{r}
mc2_events <- mc2_events %>%
  mutate(datetime = as_datetime(when))

glimpse(mc2_events)
```

## Event type distribution

```{r}
n_distinct(mc2_events$short_name)
event_count <- mc2_events %>% count(short_name, sort = TRUE)

ggplot(data = event_count,
       aes(x = reorder(short_name, n), 
           y = n)) +
  geom_col(fill = "#4e79a7") +
  geom_text(aes(label = scales::comma(n)),
            hjust = -0.1,
            size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = scales::comma) +
  labs(title = "Distribution of Event Types in MC2 Data",
       subtitle = "Total of 24 unique event types across 185,147 events",
       x = NULL,
       y = "Number of Events") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.text.y = element_text(size = 9)
  )
```

# Identify the Anomalous SaidIT Post

## Find the anomalous post

```{r}
forum_post <- mc2_events %>%
  filter(!is.na(details.forum)) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, details.timestamp)

print(forum_post)
```

## Confirm John Windward's event on May 17

```{r}
john_may17 <- mc2_events %>%
  filter(map_lgl(parties, ~ "person:john_windward" %in% .x)) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, details.from, details.to,
         details.content, details.forum, details.poster_id,
         details.a2a.from, details.a2a.to, details.a2a.action)

datatable(john_may17,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "John Windward's Events on May 17, 2046")
```

## Narrow down to specific anomalous post

```{r}
anomalous_post <- mc2_events %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, 
         details.content_source)

datatable(anomalous_post, 
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "SaidIT Posts on May 17, 2046")
```

::: callout-note
### Key Finding

The anomalous post is event ID 373902, posted at 2046-05-17T11:21:15Z but this is in UTC which in US Eastern time is on 4.21am, which matches the challenge description. In addition `details.forum`:`general` coincide with the challenge description that its posted in a random forum.
:::

## SaidIT post timeline

```{r}
#| code-fold: true
saidit_posts <- mc2_events %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  mutate(
    is_anomalous = id == 373902,
    post_date    = as.Date(datetime),
    label        = if_else(id == 373902, "Anomalous\nJohn Windward\npost", NA_character_)
  )

ggplot(data = saidit_posts,
       aes(x = datetime,
           y = short_name,
           colour = is_anomalous,
           size   = is_anomalous)) +
  geom_point(alpha = 0.8) +
  geom_vline(xintercept = as.POSIXct("2046-05-17 11:21:15", tz = "UTC"),
             colour = "#e15759",
             linetype = "dashed",
             linewidth = 0.7) +
  ggrepel::geom_label_repel(
    aes(label = label),
    na.rm        = TRUE,
    fill         = "#e15759",
    colour       = "white",
    size         = 3,
    fontface     = "bold",
    nudge_y      = 0.4,
    segment.colour = "#e15759"
  ) +
  scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#e15759"),
                      labels = c("Normal post", "Anomalous post")) +
  scale_size_manual(values = c("FALSE" = 1.5, "TRUE" = 4),
                    guide = "none") +
  labs(title    = "All SaidIT-related Posts Over Time",
       subtitle = "John Windward's autonomous post on May 17 stands out as the only anomaly",
       x        = NULL,
       y        = NULL,
       colour   = NULL) +
  theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    axis.text.y   = element_text(size = 10)
  )
```

## Conclusion on the anomalous post

The post has `short_name` as `saidit_post` but under `parties` as `Agent/person:john_windward,system:saidit` which is weird as concluded that `saidit_post` should be posted by the person so John himself. But `parties` shows that its posted by John's agent instead.

In addition based on John's event on May 17, there was only `check_email`. He was also not active on 4:21 am, which means his agent acted autonomously.

# Organisational Context

## Inspect nodes and edges only

```{r}
org_nodes <- as_tibble(org_raw$nodes)
org_edges <- as_tibble(org_raw$edges)

glimpse(org_nodes)
glimpse(org_edges)
```

## Understanding John Windward

```{r}
org_nodes %>% filter(str_detect(id, "john_windward"))

org_edges %>%
  filter(str_detect(source, "john_windward") |
           str_detect(target, "john_windward"))

org_edges %>%
  filter(str_detect(source, "customer_support") |
           str_detect(target, "customer_support")) %>%
  left_join(org_nodes, by = c("target" = "id"))
```

::: callout-note
### Key Findings

John leads Customer Support which contains 4 teams: Phone Center, Billing, Concierge and Public Relations. The Customer Support sits directly under `company:tenant_thread`.
:::

## Organisation chart visualisation

```{r}
#| code-fold: true
#| fig-width: 10
#| fig-height: 7
org_graph <- tbl_graph(
  nodes = org_nodes %>%
    mutate(
      highlight = case_when(
        id == "person:john_windward" ~ "John Windward",
        str_detect(id, "customer_support") ~ "Customer Support",
        type == "department" ~ "Department",
        type == "company" ~ "Company",
        TRUE ~ "Other"
      )
    ),
  edges = org_edges %>%
    rename(from = source, to = target) %>%
    left_join(org_nodes %>% select(id), 
              by = c("from" = "id")) %>%
    left_join(org_nodes %>% select(id), 
              by = c("to" = "id")),
  directed = TRUE
)

ggraph(org_graph, layout = "tree") +
  geom_edge_diagonal(colour = "grey70", 
                     arrow = arrow(length = unit(3, "mm"), 
                                   type = "closed"),
                     end_cap = circle(4, "mm")) +
  geom_node_point(aes(colour = highlight,
                      size = highlight == "John Windward"),
                  show.legend = TRUE) +
  geom_node_text(aes(label = label),
                 repel = TRUE,
                 size = 2.5,
                 colour = "grey20") +
  scale_colour_manual(
    values = c(
      "John Windward" = "#e15759",
      "Customer Support" = "#f28e2b",
      "Department" = "#4e79a7",
      "Company" = "#59a14f",
      "Other" = "grey80"
    )
  ) +
  scale_size_manual(values = c("TRUE" = 6, "FALSE" = 3),
                    guide  = "none") +
  labs(title = "Tenant Thread Organisation Chart",
       subtitle = "John Windward (red) leads Customer Support department",
       colour = NULL) +
  theme_graph(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "bottom"
  )
```

# Trace the Immediate Trigger Chain

## Full details of anchor event (anomalous event)

```{r}
anchor <- mc2_events %>% filter(id == 373902)
anchor_time <- anchor$datetime

anchor %>%
  select(where(~!all(is.na(.)))) %>%
  select(id, datetime, short_name, parties, 
         details.forum, details.content_source) %>%
  kable(caption = "Full details of the anomalous anchor event (ID 373902)")
```

## Check any activities with John in parties within 2 hours before post

```{r}
pre_post <- mc2_events %>%
  filter(datetime >= anchor_time - hours(2),
         datetime <= anchor_time) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "john_windward")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from,
         details.task, details.a2a.from, details.a2a.to,
         details.a2a.action, details.forum, details.target_agent)

datatable(pre_post, 
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "2 hours before anomalous post - John Windward events")
```

::: callout-note
## Key Findings

The anchor event content source is from SwiftWren.txt which should be the file that contained the gibberish content. In addition, from the 2-hour window, there are only 3 events related to John shows and they all happen within 3 seconds. Agent/person:chloe_ballast queued a task that triggered John's agent, which means that Chloe Ballast's agent iniitated this not John's.
:::

## What was Chloe's agent doing just before

```{r}
chloe_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(2),
         datetime <= anchor_time) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "chloe_ballast")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from,
         details.to, details.task, details.file, details.content,
         details.a2a.from, details.a2a.to, details.a2a.action,
         details.target_agent, details.content_source)

datatable(chloe_pre,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "Chloe Ballast agent events in 2 hours before anomalous post")
```

## What was Daniel's agent doing before triggering Chloe's agent

```{r}
daniel_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(3),
         datetime <= as.POSIXct("2046-05-17 11:12:24", 
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "daniel_gangway")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from, 
         details.to, details.task, details.file, details.content,
         details.a2a.from, details.a2a.to, details.a2a.action,
         details.target_agent, details.content_source)

datatable(daniel_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Daniel Gangway events before triggering Chloe's agent")
```

## What was Lily's agent doing before triggering Daniel's?

```{r}
lily_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(6),
         datetime <= as.POSIXct("2046-05-17 10:11:31", 
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "lily_anchorline")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from,
         details.to, details.task, details.file, details.content,
         details.a2a.from, details.a2a.to, details.a2a.action,
         details.target_agent, details.content_source)

datatable(lily_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Lily Anchorline events before triggering Daniel's agent")
```

## What was Zoey's agent doing before triggering Daniel's?

```{r}
zoey_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(6),
         datetime <= as.POSIXct("2046-05-17 08:16:07", 
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "zoey_drydock")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from,
         details.to, details.task, details.file, details.content,
         details.a2a.from, details.a2a.to, details.a2a.action,
         details.target_agent, details.content_source)

datatable(zoey_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Zoey Drydock events before triggering Daniel's agent")
```

## Immediate trigger chain timeline
```{r}
#| code-fold: true
#| fig-width: 9
trigger_events <- tibble(
  time = c("09:32:00", "09:38:04", "10:11:31", "11:12:24", "11:21:13", "11:21:15"),
  agent = c("daniel_gangway", "chloe_ballast", "lily→daniel", 
             "daniel→chloe", "chloe→john", "john_windward"),
  action = c("read_file", "read_file", "queue_subordinate_task",
             "queue_subordinate_task", "queue_subordinate_task", "saidit_post"),
  step = 1:6
)

p_chain <- ggplot(data = trigger_events,
                  aes(x = step, y = 1, label = agent)) +
  geom_segment(aes(x = step, xend = step + 0.85,
                   y = 1, yend = 1),
               arrow = arrow(length = unit(3, "mm"), 
                             type = "closed"),
               colour = "grey60",
               linewidth = 0.6) +
  geom_point(aes(colour = step == 6),
             size = 8,
             show.legend = FALSE) +
  geom_text(aes(label = str_wrap(agent, 10)),
            vjust = -1.2, size = 2.8, colour = "grey20") +
  geom_text(aes(label = action),
            vjust = 2.5, size = 2.5, colour = "grey50",
            fontface = "italic") +
  geom_text(aes(label = time),
            vjust = 4.2, size = 2.3, colour = "grey60") +
  scale_colour_manual(values = c("FALSE" = "#4e79a7", 
                                 "TRUE" = "#e15759")) +
  scale_x_continuous(limits = c(0.5, 7)) +
  scale_y_continuous(limits = c(0.5, 1.5)) +
  labs(title = "Immediate Trigger Chain on May 17, 2046",
       subtitle = "From Lily's queue_subordinate_task to John's autonomous post in under 70 minutes") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 9, colour = "grey40"))

p_chain
```

# Root Cause Analysis

## Full propagation

Find all queue_subordinate_task events that has task of read_file across the entire data range to get the full propagation event.

```{r}
full_propagation <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         details.task == "read_file") %>%
  arrange(datetime) %>%
  mutate(
    sender = map_chr(parties, ~ .x[1]),
    receiver = map_chr(parties, ~ if(length(.x) >= 2) .[2] else NA_character_),
    sender_clean = str_remove(sender, 
                              "Agent/person:|agent:person:"),
    receiver_clean = str_remove(receiver,
                                "Agent/person:|agent:person:"),
    date = as.Date(datetime)
  )

cat("Total propagation events:", nrow(full_propagation), "\n")
cat("Date range:", 
    format(min(full_propagation$date)), "to", 
    format(max(full_propagation$date)), "\n")

full_propagation %>%
  select(id, datetime, sender_clean, receiver_clean) %>%
  slice_head(n = 10) %>%
  knitr::kable(caption = "First 10 rows of full_propagation — verifying sender/receiver extraction")
```

::: callout-note
something about the result showing it si a runaway loop
:::

```{r}
#| code-fold: true
#| fig-width: 13
#| fig-height: 6
loop_daily <- full_propagation %>%
  count(date, name = "n_events")

ggplot(data = loop_daily,
       aes(x = date, 
           y = n_events)) +
  geom_col(aes(fill = date == as.Date("2046-05-17")),
           show.legend = FALSE) +
  geom_vline(xintercept = as.Date("2046-05-17"),
             colour = "#e15759",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("label",
           x = as.Date("2046-05-17"),
           y = max(loop_daily$n_events) * 0.85,
           label = "Anomalous\npost occurs",
           colour = "#e15759",
           fill = "white",
           size = 3.5,
           hjust  = -0.05) +
  scale_fill_manual(values = c("FALSE" = "#4e79a7", 
                               "TRUE" = "#e15759")) +
  scale_x_date(date_breaks = "2 days",
               date_labels = "%b %d") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Daily read_file Propagation Events",
       subtitle = "The runaway loop was active for 9 days before triggering the anomalous post",
       x = NULL,
       y = "Number of Propagation Events") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

## Who is involved
```{r}
#| code-fold: true

# Nodes
loop_nodes <- full_propagation %>%
  pivot_longer(cols = c(sender_clean, receiver_clean),
               values_to = "agent") %>%
  filter(!is.na(agent), agent != "") %>%
  distinct(agent) %>%
  mutate(
    id = agent,
    label = agent,
    group = case_when(
      agent == "john_windward" ~ "Final Poster",
      agent == "chloe_ballast" ~ "Direct Trigger",
      agent %in% c("daniel_gangway",
                   "lily_anchorline",
                   "zoey_drydock") ~ "Chain Member",
      agent == "liam_anchor" ~ "Loop Origin",
      TRUE ~ "Loop Participant"
    ),
    title = paste0("<b>", label, "</b><br>Role: ", group)
  )

# Edges
loop_edges <- full_propagation %>%
  filter(!is.na(sender_clean), !is.na(receiver_clean)) %>%
  mutate(
    from = sender_clean,
    to = receiver_clean,
    title = format(datetime, "%Y-%m-%d %H:%M"),
    value = 1
  ) %>%
  select(from, to, title, value)

# Plot
visNetwork(nodes = loop_nodes,
           edges = loop_edges,
           main = "Agent read_file Propagation Loop (May 8–17, 2046)",
           height = "600px",
           width = "100%") %>%
  visGroups(groupname = "Final Poster", color = "#e15759") %>%
  visGroups(groupname = "Direct Trigger", color = "#f28e2b") %>%
  visGroups(groupname = "Chain Member", color = "#edc948") %>%
  visGroups(groupname = "Loop Origin", color = "#59a14f") %>%
  visGroups(groupname = "Loop Participant", color = "#4e79a7") %>%
  visEdges(arrows = "to",
           smooth = list(type = "curvedCW")) %>%
  visOptions(highlightNearest = list(enabled = TRUE,
                                     degree = 1,
                                     hover = TRUE),
             nodesIdSelection = TRUE,
             selectedBy = "group") %>%
  visLegend(position = "right", main = "Agent Role") %>%
  visLayout(randomSeed = 42) %>%
  visPhysics(stabilization = TRUE)
```

## Backward trace

```{r}
first_event <- full_propagation %>%
  slice_min(datetime, n = 1) %>%
  select(id, datetime, sender_clean, receiver_clean)

kable(first_event,
      caption = "First ever read_file propagation event — the loop origin",
      col.names = c("Event ID", "Datetime", "Sender", "Receiver"))

cat("Total unique agents caught in the loop:", 
    n_distinct(c(full_propagation$sender_clean, 
                 full_propagation$receiver_clean)), "\n")

full_propagation %>%
  count(sender_clean, sort = TRUE, name = "times_as_sender") %>%
  rename(agent = sender_clean) %>%
  kable(caption = "How many times each agent triggered the next agent in the loop")
```

# SwiftWren.txt

## What is SwiftWren.txt?

```{r}
swiftwren <- mc2_events %>%
  filter(
    str_detect(coalesce(details.file, ""), "SwitfWren") |
      str_detect(coalesce(details.content_source, ""), "SwiftWren") |
      str_detect(coalesce(details.file_saved, ""), "SwitfWren")
  ) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.file,
         details.file_saved, details.content_source,
         details.content)

datatable(swiftwren,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "All events referencing SwiftWren.txt")
```

::: callout-note
## Key Finding

Since it can only be found once during the anchor event, it means it was created and used without leaving a broader trail. Next course of action should be to search more broadly for it.
:::

## Search more broadly for SwiftWren

```{r}
chain_agents <- c("zoey_drydock", "lily_anchorline", 
                  "daniel_gangway","chloe_ballast",
                  "john_windward")

mc2_events %>%
  filter(short_name == "create_file") %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, paste(chain_agents, 
                                            collapse="|"))))) %>%
  arrange(datetime) %>%
  select(id, datetime, parties, details.file, details.file_saved,
         details.content) %>%
  datatable(caption = "create_file events by agents in the trigger chain")
```

```{r}
mc2_events %>%
  filter(id == 373893) %>%
  select(where(~!all(is.na(.)))) %>%
  glimpse()
```

```{r}
mc2_events %>%
  filter(short_name == "create_file",
         !is.na(details.content),
         str_length(details.content) > 100) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, paste(chain_agents,
                                            collapse="|"))))) %>%
  select(id, datetime, parties, details.content)
```
