---
title: "Take-Home Exercise 2 - VAST Mini Challenge 2"
author: "Amelia Rosalind"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

# Getting Started

## Installing of required packages

```{r}
pacman::p_load(tidyverse, jsonlite, igraph, tidygraph, 
               ggraph, visNetwork, plotly, DT,
               gganimate, tidytext, stringdist)
```

## Importing data

```{r}
mc2_raw <- fromJSON("mc2_data/MC2_data.json", flatten = TRUE)
org_raw <- fromJSON("mc2_data/org_chart.json", flatten = TRUE)
```

# Inspect MC2 Data

## Inspect structure

```{r}
str(mc2_raw, max.level = 2)
names(mc2_raw)
```

## Inspect desciption and events only

```{r}
mc2_desc <- mc2_raw$description
mc2_events <- as_tibble(mc2_raw$events)
```

## Parse timestamp into readable datetime

```{r}
mc2_events <- mc2_events %>%
  mutate(datetime = as_datetime(when))
```

## Quick sanity checks

```{r}
glimpse(mc2_events)
range(mc2_events$datetime)
n_distinct(mc2_events$short_name)
mc2_events %>% count(short_name, sort = TRUE) %>% print (n = 24)
```

## Find the anomalous SaidIT post

```{r}
forum_post <- mc2_events %>%
  filter(!is.na(details.forum)) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, details.timestamp)

print(forum_post)
```

## Confirm John Windward's event on May 17

```{r}
john_may17 <- mc2_events %>%
  filter(map_lgl(parties, ~ "person:john_windward" %in% .x)) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, details.from, details.to,
         details.content, details.forum, details.poster_id,
         details.a2a.from, details.a2a.to, details.a2a.action)

datatable(john_may17,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "John Windward's Events on May 17, 2046")
```

## Narrow down to specific anomalous post

```{r}
anomalous_post <- mc2_events %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, 
         details.content_source)

datatable(anomalous_post, 
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "SaidIT Posts on May 17, 2046")
```

::: callout-note
### Key Finding

The anomalous post is event ID 373902, posted at 2046-05-17T11:21:15Z but this is in UTC which in US Eastern time is on 4.21am, which matches the challenge description. In addition `details.forum`:`general` coincide with the challenge description that its posted in a random forum.
:::

## Find difference between saidit_post and post_saidit

```{r}
bind_rows(
  mc2_events %>% filter(short_name == "saidit_post") %>% head(5),
  mc2_events %>% filter(short_name == "post_saidit") %>% head(5)
) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content,
         details.content_source) %>%
  datatable(options = list(scrollX = TRUE, pageLength = 5),
           caption = "Comparing saidit_post vs post_saidit action types")
```

::: callout-note
### Key Findings

`saidit_post` events show `parties` as `person:...` and `post_saidit` events show `parties` as `Agent/person:...,system:saidit`, which means `saidit_post` should be posted by the person themselves and `post_saidit` are posted by an agent on behalf of the person.
:::

## Conclusion on the anomalous post

The post has `short_name` as `saidit_post` but under `parties` as `Agent/person:john_windward,system:saidit` which is weird as concluded that `saidit_post` should be posted by the person so John himself. But `parties` shows that its posted by John's agent instead.

In addition based on John's event on May 17, there was only `check_email`. He was also not active on 4:21 am, which means his agent acted autonomously.

# Inspect organisation chart

## Inpect structure

```{r}
str(org_raw, max.level = 2)
names(org_raw)
```

## Inspect nodes and edges only

```{r}
org_nodes <- as_tibble(org_raw$nodes)
org_edges <- as_tibble(org_raw$edges)

glimpse(org_nodes)
glimpse(org_edges)
```

## Check relations type for edges 

```{r}
org_edges %>% count(relation, sort = TRUE)
```

## Check types of nodes

```{r}
org_nodes %>% count(type, sort = TRUE)
```

## Find John Windward

```{r}
org_nodes %>% filter(str_detect(id, "john_windward"))
```

## Check John's connections in the org

```{r}
org_edges %>%
  filter(str_detect(source, "john_windward") |
           str_detect(target, "john_windward"))
```

## Check all people in Tenant Thread with their titles and teams

```{r}
org_nodes %>%
  filter(type == "person") %>%
  select(id, label, title) %>%
  left_join(
    org_edges %>% select(source, target, relation),
    by = c("id" = "target")
  ) %>%
  mutate(department_or_team = source) %>%
  select(person_id = id, name = label, title,
         department_or_team) %>%
  arrange(department_or_team) %>%
  datatable(options = list(scrollX = TRUE, pageLength = 15),
            caption = "All People in Tenant Thread")
```

## What teams/people does John's department contain
```{r}
org_edges %>%
  filter(str_detect(source, "customer_support") |
           str_detect(target, "customer_support")) %>%
  left_join(org_nodes, by = c("target" = "id"))
```

## Context of John's World

John leads Customer Support which contains 4 teams: Phone Center, Billing, Concierge and Public Relations. The Customer Support sits directly under `company:tenant_thread`.

# Trace the Event Chain

## Full details of anchor event (anomalous event)
```{r}
anchor <- mc2_events %>% filter(id == 373902)
anchor_time <- anchor$datetime

anchor %>%
  select(where(~!all(is.na(.)))) %>%
  glimpse()
```

## Check any activities with John in parties within 2 hours before post

```{r}
pre_post <- mc2_events %>%
  filter(datetime >= anchor_time - hours(2),
         datetime <= anchor_time) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "john_windward")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from,
         details.task, details.a2a.from, details.a2a.to,
         details.a2a.action, details.forum, details.target_agent)

datatable(pre_post, 
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "2 hours before anomalous post - John Windward events")
```

::: callout-note
## Key Findings At This Point

The anchor event content source is from SwiftWren.txt which should be the file that contained the gibberish content. In addition, from the 2-hour window, there are only 3 events related to John shows and they all happen within 3 seconds. Agent/person:chloe_ballast queued a task that triggered John's agent, which means that Chloe Ballast's agent iniitated this not John's.   
:::

## Who is Chloe Ballast?

```{r}
org_nodes %>% filter(str_detect(id, "chloe_ballast"))
```

## What was Chloe's agent doing just before
```{r}
chloe_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(2),
         datetime <= anchor_time) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "chloe_ballast")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from,
         details.to, details.task, details.file, details.content,
         details.a2a.from, details.a2a.to, details.a2a.action,
         details.target_agent, details.content_source)

datatable(chloe_pre,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "Chloe Ballast agent events in 2 hours before anomalous post")
```

::: callout-note
## Key Findings At This Point

Daniel Gangway's agent triggered Chloe's agent, which triggered John's agent.
:::

## Who is Daniel Gangway?

```{r}
org_nodes %>% filter(str_detect(id, "daniel_gangway"))

org_edges %>%
  filter(str_detect(target, "daniel_gangway"))
```

## What was Daniel's agent doing before triggering Chloe's agent
```{r}
daniel_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(3),
         datetime <= as.POSIXct("2046-05-17 11:12:24", 
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "daniel_gangway")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from, 
         details.to, details.task, details.file, details.content,
         details.a2a.from, details.a2a.to, details.a2a.action,
         details.target_agent, details.content_source)

datatable(daniel_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Daniel Gangway events before triggering Chloe's agent")
```

## Who is Lily Anchorline?
```{r}
org_nodes %>% filter(str_detect(id, "lily_anchorline"))

org_edges %>% filter(str_detect(target, "lily_anchorline"))
```

## What was Lily's agent doing before triggering Daniel's?
```{r}
lily_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(6),
         datetime <= as.POSIXct("2046-05-17 10:11:31", 
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "lily_anchorline")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from,
         details.to, details.task, details.file, details.content,
         details.a2a.from, details.a2a.to, details.a2a.action,
         details.target_agent, details.content_source)

datatable(lily_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Lily Anchorline events before triggering Daniel's agent")
```

## Who is Zoey Drydock?
```{r}
org_nodes %>% filter(str_detect(id, "zoey_drydock"))

org_edges %>% filter(str_detect(target, "zoey_drydock"))
```

## What was Zoey's agent doing before triggering Daniel's?
```{r}
zoey_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(6),
         datetime <= as.POSIXct("2046-05-17 08:16:07", 
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "zoey_drydock")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.from,
         details.to, details.task, details.file, details.content,
         details.a2a.from, details.a2a.to, details.a2a.action,
         details.target_agent, details.content_source)

datatable(zoey_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Zoey Drydock events before triggering Daniel's agent")
```

## Get ALL queue_subordinate_task events with read_file task on May 16-17

```{r}
propagation_chain <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         details.task == "read_file",
         as.Date(datetime) %in% as.Date(c("2046-05-16", "2046-05-17"))) %>%
  arrange(datetime) %>%
  select(id, datetime, parties, details.target_agent) %>%
  mutate(
    sender = map_chr(parties, ~ .x[!str_detect(.x, as.character(details.target_agent[1]))][1]),
    receiver = details.target_agent
  )

datatable(propagation_chain,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Full read_file propagation chain")
```

```{r}
# Check ALL queue_subordinate_task with read_file task in entire dataset
full_propagation <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         details.task == "read_file") %>%
  arrange(datetime) %>%
  select(id, datetime, parties, details.target_agent) %>%
  mutate(
    sender = map_chr(parties, ~ {
      target <- details.target_agent[1]
      others <- .x[!str_detect(.x, fixed(str_remove(target, "Agent/")))]
      if(length(others) > 0) others[1] else NA_character_
    }),
    receiver = details.target_agent
  )

# How many total? And what's the date range?
nrow(full_propagation)
range(full_propagation$datetime)

# Who appears before May 16?
full_propagation %>%
  filter(datetime < as.POSIXct("2046-05-16 01:01:55", tz = "UTC")) %>%
  datatable(options = list(scrollX = TRUE, pageLength = 10),
            caption = "read_file propagation events BEFORE Liam Anchor's first event")
```

```{r}
# Check: did john_windward's agent appear in the loop before May 17
# and NOT trigger a post?
john_in_loop <- full_propagation %>%
  filter(str_detect(coalesce(sender, ""), "john_windward") |
         str_detect(coalesce(receiver, ""), "john_windward")) %>%
  arrange(datetime) %>%
  select(id, datetime, sender, receiver)

datatable(john_in_loop,
          caption = "All times John Windward appeared in the read_file loop")

# Check: on those prior dates, did john's agent also do saidit_post_check?
john_saidit_checks <- mc2_events %>%
  filter(short_name == "saidit_post_check",
         map_lgl(parties, ~ any(str_detect(.x, "john_windward")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties)

datatable(john_saidit_checks,
          caption = "All saidit_post_check events for John's agent")

# Check: what file did john's agent read on May 17 vs other dates
john_reads <- mc2_events %>%
  filter(short_name == "read_file",
         map_lgl(parties, ~ any(str_detect(.x, "john_windward")))) %>%
  arrange(datetime) %>%
  select(id, datetime, parties, details.file, 
         details.content, details.content_source)

datatable(john_reads,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "All read_file events for John Windward's agent")
```

#  Understanding SwiftWren.txt

## What is SwiftWren.txt?
```{r}
swiftwren <- mc2_events %>%
  filter(
    str_detect(coalesce(details.file, ""), "SwitfWren") |
      str_detect(coalesce(details.content_source, ""), "SwiftWren") |
      str_detect(coalesce(details.file_saved, ""), "SwitfWren")
  ) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.file,
         details.file_saved, details.content_source,
         details.content)

datatable(swiftwren,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "All events referencing SwiftWren.txt")
```

::: callout-note
## Key Finding

Since it can only be found once during the anchor event, it means it was created and used without leaving a broader trail. Next course of action should be to search more broadly for it. 
:::

## Search more broadly for SwiftWren
```{r}
mc2_events %>%
  filter(
    str_detect(coalesce(details.file, ""), 
               regex("swift|wren", ignore_case = TRUE)) |
      str_detect(coalesce(details.content_source, ""),
                 regex("swift|wren", ignore_case = TRUE)) |
      str_detect(coalesce(details.file_saved, ""),
                 regex("swift|wren", ignore_case = TRUE)) |
      str_detect(coalesce(details.content, ""), 
                 regex("swift|wren", ignore_case = TRUE))
  ) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.file, 
         details.file_saved, details.content_source, 
         details.content) %>%
  datatable(options = list(scrollX = TRUE, pageLength = 20),
            caption = "Broader search for SwiftWren references")
```


