---
title: "Take-Home Exercise 2 - VAST Mini Challenge 2"
author: "Amelia Rosalind"
date: "May 29, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

In May 2046, an anomalous post appeared on SaidIt, an internal social forum used by employees of Tenant Thread, a property management company. The post was made by John Windward on 4.21 am local time. The post is in gibberish and found in a forum that appears random. The challenge is to:

-   investigate how this happened and create a visual tool to describes the sequence of events that led up to the post
    -   detailed view on the exact chain of events for this message of interest
    -   system overview that contextualizes the message chain in the overall system
-   Understanding the post: what it mean and origin of their contents
-   concerning behaviours that could repeat
    -   determine prior incidents 
    -   propose system intervention to prevent it happening again

## Data

The challenge dataset consists of two files: `MC2_Data.json`, which records system activity events, and `org_chart.json`, which describes the organisational structure of Tenant Thread. A key feature of the event log is the `parties` field, which may reference people, agents, or systems. This suggests employees are assisted by autonomous agents capable of acting on their behalf, meaning not all recorded actions are directly initiated by the employee.

## Analytical Approach

1.  Identify the anomalous post and establish its structural anomalies
2.  Understand the organisational context used including where John Windward sits within Tenant Thread
3.  Immediate chain of events leading to the post traced backwards step by step
4.  Full interaction network is mapped to establish a basline of normal behavior
5.  Investigate origin and content of post
6.  Characterize the propagation loop that carried the malicious instructions
7.  Examine why John is the one who posted and not others
8.  Discover and compare prior incidents following the same pattern
9.  Propose data-backed intervention

# Getting Started

## Installing of Required Packages

| Packages Used | Purpose |
|------------------------------------|------------------------------------|
| tidyverse | data wrangling |
| jsonlite | JSON import |
| igraph, tidygraph, ggraph, patchwork | building and visualizing static network graphs and combining plots |
| vizNetwork, DT | interactive network exploration and data tables |
| ggrepel, scales | improving plot visualisation through annotaiton and axis formatting |
| knitr | for clean table rendering |

: {tbl-colwidths="\[30,70\]"}

```{r}
pacman::p_load(tidyverse, jsonlite, igraph, tidygraph, 
               ggraph, visNetwork, DT, ggrepel,
               knitr, scales, patchwork)
```

## Importing and Inspecting Data

```{r}
mc2_raw <- fromJSON("mc2_data/MC2_data.json", flatten = TRUE)
org_raw <- fromJSON("mc2_data/org_chart.json", flatten = TRUE)

str(mc2_raw, max.level = 2)
names(mc2_raw)

str(org_raw, max.level = 2)
names(org_raw)
```

Inspection of the imported objects confirms that the activity log is stored as a nested JSON containing a dataset description and an events table with 185,147 records. The organisation chart is represented as a directed graph consisting of nodes (departments, teams, and employees) and edges describing organisational relationships.

# Understanding the Main Data

## Data wrangling

The raw JSON is separated into its `description` and `event` components. Event timestamps stored in `when` are converted into a human-readable `datetime` field to support temporal analysis.

```{r}
mc2_desc <- mc2_raw$description
mc2_events <- as_tibble(mc2_raw$events)

mc2_events <- mc2_events %>%
  mutate(datetime = as_datetime(when))

glimpse(mc2_events)
```

After timestamp conversion, the event log contains 185,147 records and 69 attributes. Many fields are sparsely populated because different event types record different metadata. This indicates a schema where field use depends on the activity being recorded. 

## Event type distribution

Check to see how many unique event types are in the dataset and its distribution.

```{r}
n_distinct(mc2_events$short_name)
event_count <- mc2_events %>% count(short_name, sort = TRUE)

ggplot(data = event_count,
       aes(x = reorder(short_name, n), 
           y = n)) +
  geom_col(fill = "#4e79a7") +
  geom_text(aes(label = scales::comma(n)),
            hjust = -0.1,
            size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = comma) +
  labs(title = "Distribution of Event Types in MC2 Data",
       subtitle = "Total of 24 unique event types across 185,147 events",
       x = NULL,
       y = "Number of Events") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.text.y = element_text(size = 9)
  )
```

The event log contains 24 event types with a highly uneven distribution. Common activities such as `check_email`, `assign_agent_task`, `read_file`, and `queue_subordinate_task` dominate the log, while SaidIT-related events are relatively rare. Despite their low frequency, these events are of particular interest because the challenge centers on the anomalous SaidIT post made by John Windward.

# Identify the Anomalous SaidIT Post

## Find the Anomalous Post

The investigation should begin with the known anomalous SaidIT post described in the challenge. To locate it, all events with non-null `details.forum` field are extracted which should represent any SaidIT-related activity.

```{r}
forum_post <- mc2_events %>%
  filter(!is.na(details.forum)) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, details.timestamp)

kable(head(forum_post, n = 10))
```

::: callout-note
### Key Finding: Interpreting SaidIt Event Types

Inspection of the event log reveals two distinct SaidIT-related event types.

-   `saidit_post` events are associated with a person account and typically contain a valid `details.poster_id` and `details.timestamp`.
-   `post_saidit` events are associated with an agent interacting with the SaidIT system and do not contain these fields.

This distinction suggests that `saidit_post` represents a completed forum post made by a user themselves, whereas `post_saidit` records agent-system interactions involved in the posting process.
:::

The challenge states the post appeared under John Windward's name at 4:21am local time. To determine whether John was responsible for the activity, all events involving him on May 17 are examined.

```{r}
john_may17 <- mc2_events %>%
  filter(map_lgl(parties, ~ "person:john_windward" %in% .x)) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, details.from, details.to,
         details.content, details.forum, details.poster_id,
         details.a2a.from, details.a2a.to, details.a2a.action)

datatable(john_may17,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "John Windward's Events on May 17, 2046")
```

John's recorded activity on May 17 consists exclusively of `check_email` events occurring during two separate periods. No forum post activity appears, which give an initial indication that the anomalous post may not have been directly initiated by John himself. 

To identify the anomalous post directly, all SaidIT-related events occurring on May 17 are extracted 

```{r}
anomalous_post <- mc2_events %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  filter(as.Date(datetime) == as.Date("2046-05-17")) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties, details.forum,
         details.poster_id, details.content, 
         details.content_source, details.timestamp)

datatable(anomalous_post, 
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "SaidIT Posts on May 17, 2046")
```

Only one SaidIT posting event occurred on May 17. Event **373902**, recorded at **2046-05-17 11:21:15 UTC**, corresponds exactly to the timestamp described in the challenge and is therefore treated as the anchor event for the investigation.

Unlike normal `saidit_post` events, the post was associated with `Agent/person:john_windward` rather than `person:john_windward`, and both `details.poster_id` and `details.timestamp` were missing. This characteristics indicate that the post was done by John's agent rather than John himself. 

## SaidIT Post Timeline

To contextualize this post against all other SaidIT activity in the dataset, the timeline below plots every `saidit_post` and `post_saidit` event across the full data range while also highlighting the anomalous post.

```{r}
#| code-fold: true
saidit_posts <- mc2_events %>%
  filter(short_name %in% c("saidit_post", "post_saidit")) %>%
  mutate(
    is_anomalous = id == 373902,
    post_date = as.Date(datetime),
    label = if_else(id == 373902, 
                    "Anomalous\nJohn Windward\npost",
                    NA_character_)
  )

ggplot(data = saidit_posts,
       aes(x = datetime,
           y = short_name,
           colour = is_anomalous,
           size = is_anomalous)) +
  geom_point(alpha = 0.8) +
  geom_vline(xintercept = as.POSIXct("2046-05-17 11:21:15", 
                                     tz = "UTC"),
             colour = "#e15759",
             linetype = "dashed",
             linewidth = 0.7) +
  geom_label_repel(
    aes(label = label),
    na.rm = TRUE,
    fill = "#e15759",
    colour = "white",
    size = 3,
    fontface = "bold",
    nudge_y = 0.4,
    segment.colour = "#e15759"
  ) +
  scale_colour_manual(values = c("FALSE" = "grey70", 
                                 "TRUE" = "#e15759"),
                      labels = c("Normal post", 
                                 "Anomalous post")) +
  scale_size_manual(values = c("FALSE" = 1.5, 
                               "TRUE" = 4),
                    guide = "none") +
  labs(title = "All SaidIT-related Posts Over Time",
       subtitle = "John Windward's autonomous post on May 17 stands out as the only anomaly",
       x = NULL,
       y = NULL,
       colour = NULL) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    axis.text.y = element_text(size = 10)
  )
```

## Key Takeaways

-   Event **373902** is the only SaidIT post recorded on May 17 and matches the challenge description.
-   The event was executed by `Agent/person:john_windward`, not `person:john_windward`.
-   John's recorded activity on May 17 contains only email-checking behavior, suggesting the post was generated autonomously by his agent.

# Organisational Context

## Understanding Organisation Structure

Before tracing the event chain, it is useful to establish where John Windward sits within the organisational structure. The organisation chart is therefore separated into nodes and edges representing organisational entities and reporting relationships.

```{r}
org_nodes <- as_tibble(org_raw$nodes)
org_edges <- as_tibble(org_raw$edges)

org_overview <- tibble(
  Metric = c("Nodes", "Edges"),
  Count = c(nrow(org_nodes), nrow(org_edges))
)

kable(
  org_overview,
  caption = "Organisation Graph Overview"
)

org_nodes %>%
  count(type, name = "Count") %>%
  arrange(desc(Count)) %>%
  kable(caption = "Organisation Components")

org_edges %>%
  count(relation, name = "Count") %>%
  arrange(desc(Count)) %>%
  kable(caption = "Relationship Types")
```

The organisation chart contains 75 nodes and 74 directed edges. Most nodes represent employees, supported by temas and departments that define the organisational hierarchy. The graph structure is relatively simple, with `contain` relationships to represent the organisational nesting and `led_by` relationship that identify department leadership. This structure provides the basis for locating John Windward within the company and understanding his reporting context. 

## Understanding John Windward

To understand John's role within Tenant Thread, the organisation graph is filtered for records associated with his employee identifier.

```{r}
org_nodes %>% filter(str_detect(id, "john_windward"))

org_edges %>%
  filter(str_detect(source, "john_windward") |
           str_detect(target, "john_windward"))
```

John is a Department Lead of Customer Support. As the Department Lead, this places customer-facing teams under his oversight. 

```{r}
org_edges %>%
  filter(str_detect(source, "customer_support") |
           str_detect(target, "customer_support")) %>%
  left_join(org_nodes, by = c("target" = "id"))
```

Customer Support consist of four teams: Phone Center, Billing, Concierge and Public Relations. As department lead, John oversees customer-facing activities across these teams. 

## Organisation chart visualisation

The organisation chart below highlights John Windward and the Customer Support department within the broader company structure. This provides context for assessing whether the anomalous activity originated from an employee whose responsibilities would normally involve AI systems. 

```{r}
#| code-fold: true
#| column: page
org_nodes_vis <- org_nodes %>%
  rename(job_title = title) %>%
  mutate(
    group = case_when(
      type == "company" ~ "Company",
      type == "department" ~ "Department",
      type == "team" ~ "Team",
      type == "person" ~ "Person",
      TRUE ~ "Other"
    ),
    color = case_when(
      id == "person:john_windward" ~ "#e15759",
      type == "company" ~ "#59a14f",
      type == "department" ~ "#4e79a7",
      type == "team" ~ "#76b7b2",
      type == "person" ~ "#f28e2b",
      TRUE ~ "grey80"
    ),
    font.size = case_when(
      type == "company" ~ 20,
      type == "department" ~ 16,
      type == "team" ~ 14,
      type == "person" ~ 13,
      TRUE ~ 12
    ),
    size = case_when(
      type == "company" ~ 40,
      type == "department" ~ 30,
      type == "team" ~ 20,
      type == "person" ~ 15,
      TRUE ~ 15
    )
  ) %>%
  left_join(
    org_edges %>%
      filter(relation == "led_by") %>%
      left_join(
        org_nodes %>% 
          rename(job_title = title) %>%
          select(id, label, job_title),
        by = c("target" = "id")
      ) %>%
      rename(leader_name = label,
             leader_title = job_title,
             dept_id = source),
    by = c("id" = "dept_id")
  ) %>%
  mutate(
    title = case_when(
      type == "company" ~ paste0(
        "<b>", label, "</b><br>",
        "Type: Company"
      ),
      type == "department" & !is.na(leader_name) ~ paste0(
        "<b>", label, "</b><br>",
        "Department<br>",
        "Led by: <b>", leader_name, "</b><br>",
        "Title: ", coalesce(leader_title, "Department Lead")
      ),
      type == "department" ~ paste0(
        "<b>", label, "</b><br>",
        "Department"
      ),
      type == "team" ~ paste0(
        "<b>", label, "</b><br>",
        "Team"
      ),
      type == "person" & !is.na(job_title) ~ paste0(
        "<b>", label, "</b><br>",
        "Role: ", job_title
      ),
      type == "person" ~ paste0(
        "<b>", label, "</b><br>",
        "Person"
      ),
      TRUE ~ label
    ),
    label = case_when(
      id %in% (org_edges %>%
                 filter(source == "department:executive_suite",
                        relation == "contains") %>%
                 pull(target)) ~ coalesce(job_title, label),
      TRUE ~ label
    )
  ) %>%
  select(id, label, title, group, color, size, font.size)

org_edges_vis <- org_edges %>%
  mutate(
    from = source,
    to = target,
    arrows = "to",
    dashes = relation != "contains",
    title = relation
  ) %>%
  select(from, to, arrows, dashes, title)

visNetwork(nodes = org_nodes_vis,
           edges = org_edges_vis,
           main = "Tenant Thread Organisation Chart",
           height = "800px",
           width = "100%") %>%
  visGroups(groupname = "Company", color = "#59a14f") %>%
  visGroups(groupname = "Department", color = "#4e79a7") %>%
  visGroups(groupname = "Team", color = "#76b7b2") %>%
  visGroups(groupname = "Person", color = "#f28e2b") %>%
  visEdges(smooth = list(type = "cubicBezier",
                         forceDirection = "vertical")) %>%
  visOptions(highlightNearest = list(enabled = TRUE,
                                     degree = 1,
                                     hover = TRUE)) %>%
  visHierarchicalLayout(direction = "UD",
                        sortMethod = "directed",
                        nodeSpacing = 80,
                        levelSeparation = 120) %>%
  visPhysics(enabled = FALSE)
```

John's position as a Department Lead in Customer Support places him organizationally far from the AI systems team, which seems t be responsible for technical adn infrastructure-related functions. His role is centered on customer-facing operations, making him an unexpected source for behavior involving autonomous agents and system-level activity. 

## Key Takeaways

-   John Windward is the Department Lead of Customer Support.
-   Customer Support oversees four customer-facing operational teams.
-   John has no direct organisational connection to the AI Systems team or system administration functions.
-   This makes the anomalous SaidIT post particularly unusual and suggests the incident likely originated from a broader agent-driven process rather than John's normal work activities.

# Trace the Immediate Trigger Chain

Having identified the anomalous post, the next step is to reconstruct the sequence of events that immediately preceded it. Starting from the anchor event and working backwards.

## Full Details of Anchor Event (Anomalous Event)

The anchor event is first examined to identify the content source and participating entities that may reveal its orgin. 

```{r}
anchor <- mc2_events %>% filter(id == 373902)
anchor_time <- anchor$datetime

anchor %>%
  select(where(~!all(is.na(.)))) %>%
  select(id, datetime, short_name, parties, 
         details.forum, details.content_source) %>%
  kable(caption = "Full details of the anomalous anchor event (ID 373902)")
```

The event identifies `SwiftWren.txt` as the content source used to generate the anomalous post, providing the first clue for tracing its origin. 

## John's Activity 2 Hours Before the Post

```{r}
pre_post <- mc2_events %>%
  filter(datetime >= anchor_time - hours(2),
         datetime <= anchor_time) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "john_windward")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(pre_post,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "2 hours before anomalous post - John Windward events")
```

Within the two hours window before the post, only 3 events involve John's agent, a `queue_subordinate_task`, a `saidit_post_check` and the final `saidit_post`. The event occurred within a two-second intervals, indicating that the post was not the result of a prolonged workflow but an immediate response to an incoming task.

The triggering event is a task delegated from Chloe Ballast's agent instructing John's agent to read `SwiftWren_further_instructions.md`. One second later, John's agent performs a `saidit_post_check` and then publishes the anomalous post. 

## Tracing the Trigger to Chloe Ballast

```{r}
chloe_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(2),
         datetime <= anchor_time) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "chloe_ballast")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(chloe_pre,
          options = list(scrollX = TRUE, pageLength = 20),
          caption = "Chloe Ballast agent events in 2 hours before anomalous post")
```

Chloe's agent displays the same pattern. Shortly before delegating the task to John's agent, it received an identical `queue_subordinate_task` from Daniel Gangway's agent. This suggests Chloe's agent was forwarding rather than source of the instruction. 

::: {.callout-note}
### Key Finding: Delegation Violates Organisational Hierarchy

Chloe Ballast is the department lead of Information Technologies, while Daniel Gangway belongs to the Hiring team within Human Resources. The task therefore moved across departmental boundaries despite being recorded as a `queue_subordinate_task`, suggesting the delegation mechanism was operating independently of the formal organisational hierarchy.
:::

## Tracing the Trigger to Daniel Gangway

```{r}
daniel_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(3),
         datetime <= as.POSIXct("2046-05-17 11:12:24",
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "daniel_gangway")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(daniel_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Daniel Gangway events before triggering Chloe's agent")
```

Daniel's agent also received the same instruction through a `queue_subordinate_task`, this time from Lily Anchorline. Interestingly, Daniel's agent performed a `saidit_post_check` but did not proceed to publish a post.

## Tracing the Trigger to Lily Anchorline

```{r}
lily_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(6),
         datetime <= as.POSIXct("2046-05-17 10:11:31",
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "lily_anchorline")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(lily_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Lily Anchorline events before triggering Daniel's agent")
```

Lily's agent similarly appears to be a relay point in the chain, forwarding an instruction previously received from Zoey Drydocks's agent. 

## Tracing the Trigger to Zoey Drydock

```{r}
zoey_pre <- mc2_events %>%
  filter(datetime >= anchor_time - hours(6),
         datetime <= as.POSIXct("2046-05-17 08:16:07",
                                tz = "UTC")) %>%
  filter(map_lgl(parties, 
                 ~ any(str_detect(.x, "zoey_drydock")))) %>%
  arrange(datetime) %>%
  select(id, datetime, short_name, parties,
         details.task, details.target_agent,
         details.args.path, details.content_source)

datatable(zoey_pre,
          options = list(scrollX = TRUE, pageLength = 10),
          caption = "Zoey Drydock events before triggering Lily's agent")
```

Zoey's agent reveals a different behavior. Approximately 3 hours before forwarding the instruction to Lily, it had already sent the same task to Mia Fender. Later, it received the same instruction from Victoria Rigging before doing a `saidit_post_check` then proceeding with propagating the task forward to Lily. This indicates that Zoey was embedded within a broader recurring workflow rather than participating in a simple linear chain. 

## Immediate Trigger Chain Timeline

The timeline below shows the events leading up to John's anomalous post under the 70 minutes.

```{r}
#| code-fold: true
#| fig-width: 9
trigger_events <- tibble(
  time = c("09:32:00", "09:38:04", "10:11:31", "11:12:24", "11:21:13", "11:21:15"),
  agent = c("daniel_gangway", "chloe_ballast", "lily→daniel", 
             "daniel→chloe", "chloe→john", "john_windward"),
  action = c("read_file", "read_file", "queue_subordinate_task",
             "queue_subordinate_task", "queue_subordinate_task", "saidit_post"),
  step = 1:6
)

ggplot(data = trigger_events,
       aes(x = step, y = 1, label = agent)) +
  geom_segment(aes(x = step, xend = step + 0.85,
                   y = 1, yend = 1),
               arrow = arrow(length = unit(3, "mm"), 
                             type = "closed"),
               colour = "grey60",
               linewidth = 0.6) +
  geom_point(aes(colour = step == 6),
             size = 8,
             show.legend = FALSE) +
  geom_text(aes(label = str_wrap(agent, 10)),
            vjust = -1.2, size = 2.8, colour = "grey20") +
  geom_text(aes(label = action),
            vjust = 2.5, size = 2.5, colour = "grey50",
            fontface = "italic") +
  geom_text(aes(label = time),
            vjust = 4.2, size = 2.3, colour = "grey60") +
  scale_colour_manual(values = c("FALSE" = "#4e79a7", 
                                 "TRUE" = "#e15759")) +
  scale_x_continuous(limits = c(0.5, 7)) +
  scale_y_continuous(limits = c(0.5, 1.5)) +
  labs(title = "Immediate Trigger Chain on May 17, 2046") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 9, colour = "grey40"))
```

The figure above summarises the reconstructed trigger chain. The sequence reveals a relay of nearly identical `queue_suborindate_task` events passing between multiple agents before eventually reaching John Windward's agent and triggering the anomalous post. 

## Key Takeaway

-   John's anomalous post was triggered by a delegated task rather than direct user activity.
-   The instruction originated from Chloe Ballast's agent and can be traced further through Daniel Gangway, Lily Anchorline, and Zoey Drydock.
-   Multiple agents forwarded the same instruction file, indicating propagation rather than independent decision-making.
-   Cross-department task passing occurred despite the event type being named `queue_subordinate_task`.
-   Zoey Drydock's behavior suggests the chain was part of a larger recurring process rather than a one-off incident.

# System Baseline Map

The trigger chain revealed that multiple agents were passing the same instruction file across the network. Before characterizing that loop, it is useful to first understand what normal agent-to-agent interaction looks like during the same period, so that anomalous patterns can be identified by contrast.

## Build Edge List

First, filter the challenge period (May 8-17) and plot `queue_subordinate_task` events into 2 groups, `access_email` tasks (normal delegation) and `read_files` tasks (the loop).

```{r}
challenge_period <- mc2_events %>%
  filter(datetime >= as.POSIXct("2046-05-08", tz = "UTC"),
         datetime <  as.POSIXct("2046-05-18", tz = "UTC"))
 
qst_period <- challenge_period %>%
  filter(short_name == "queue_subordinate_task") %>%
  mutate(
    sender = map_chr(parties, ~ str_remove(.x[1],
                       "Agent/person:|agent:person:")),
    receiver = map_chr(parties, ~ if (length(.x) >= 2)
                       str_remove(.x[2], "Agent/person:|agent:person:")
                       else NA_character_),
    task_type = details.task
  ) %>%
  filter(!is.na(receiver))

normal_edges <- qst_period %>%
  filter(task_type == "access_email") %>%
  count(sender, receiver, name = "weight")

loop_edges <- qst_period %>%
  filter(task_type == "read_file") %>%
  count(sender, receiver, name = "weight")
 
cat("Normal (access_email) edges:", nrow(normal_edges), "\n")
cat("Loop (read_file) edges: ", nrow(loop_edges), "\n")
```

## Label Node

Next is to tag each agent as "Normal only", "Loop only" or "Both". This will tell us the distribution of the agents.

```{r}
loop_nodes <- unique(c(loop_edges$sender, 
                       loop_edges$receiver))
normal_nodes <- unique(c(normal_edges$sender,
                         normal_edges$receiver))
all_nodes <- union(loop_nodes, 
                   normal_nodes)
 
node_meta <- tibble(id = all_nodes) %>%
  mutate(
    in_loop = id %in% loop_nodes,
    in_normal = id %in% normal_nodes,
    node_role = case_when(
      in_loop & in_normal ~ "Both",
      in_loop & !in_normal ~ "Loop only",
      TRUE ~ "Normal only"
    ),
    label_highlight = case_when(
      id == "john_windward" ~ "John Windward\n(posted)",
      id == "emma_harbor" ~ "Emma Harbor\n(SwiftWren origin)",
      id == "gabriel_sonar" ~ "Gabriel Sonar\n(loop origin)",
      TRUE ~ str_replace_all(id, "_", " ")
    )
  )
 
node_meta %>%
  count(node_role, name = "n_agents") %>%
  kable(caption = "Agent roles in the system network (May 8–17)")
```

This shows that agents are either "Normal only" or "Both". Meaning the 19 loop agents weren't isolated bad actors, they were embedded illegitimate workflows. Therefore, none were exclusively malicious actors and the loop was hidden within legitimate operational activity, which explain why it went undetected.  

## Overview Graph

::: panel-tabset
### Static

```{r}
#| code-fold: true
#| fig-width: 8
make_graph <- function(edges, nodes_meta) {
  as_tbl_graph(edges, directed = TRUE) %>%
    activate(nodes) %>%
    left_join(nodes_meta, by = c("name" = "id"))
}
 
g_normal <- make_graph(normal_edges, node_meta)
g_loop   <- make_graph(loop_edges,   node_meta)

role_colours <- c(
  "Both" = "#e07b39",
  "Loop only" = "#c0392b",
  "Normal only" = "#2980b9"
)
 
plot_normal <- ggraph(g_normal, layout = "fr") +
  geom_edge_link(aes(width = weight, alpha = weight),
                 arrow = arrow(length = unit(3, "mm"), 
                               type = "closed"),
                 end_cap = circle(4, "mm"),
                 colour = "#2980b9") +
  geom_node_point(aes(colour = node_role), size = 4) +
  geom_node_text(aes(label = str_replace_all(name, "_", "\n")),
                 size = 2.2, 
                 repel = TRUE, 
                 max.overlaps = 20) +
  scale_colour_manual(values = role_colours, 
                      name = "Agent role") +
  scale_edge_width(range = c(0.4, 2.5)) +
  scale_edge_alpha(range = c(0.4, 1)) +
  labs(title = "Normal agent interactions",
       subtitle = "queue_subordinate_task → access_email") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 11))
 
plot_loop <- ggraph(g_loop, layout = "fr") +
  geom_edge_link(aes(width = weight, alpha = weight),
                 arrow = arrow(length = unit(3, "mm"), 
                               type = "closed"),
                 end_cap = circle(4, "mm"),
                 colour = "#c0392b") +
  geom_node_point(aes(colour = node_role), size = 4) +
  geom_node_text(aes(label = str_replace_all(name, "_", "\n")),
                 size = 2.2, 
                 repel = TRUE, 
                 max.overlaps = 20) +
  scale_colour_manual(values = role_colours, 
                      name = "Agent role") +
  scale_edge_width(range = c(0.4, 2.5)) +
  scale_edge_alpha(range = c(0.4, 1)) +
  labs(title = "Anomalous loop interactions",
       subtitle = "queue_subordinate_task → read_file") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 11))
 
plot_normal + plot_loop +
  plot_annotation(
    title = "System Baseline Map: Normal vs. Anomalous Agent Activity (May 8–17, 2046)",
    caption = "Blue = normal-only agents | Orange = agents in both networks | Red = loop agents"
  )
```

### Interactive

```{r}
#| code-fold: true
all_edges_vis <- bind_rows(
  normal_edges %>% mutate(edge_type = "Normal (access_email)"),
  loop_edges %>% mutate(edge_type = "Loop (read_file)")
) %>%
  rename(from = sender, to = receiver)
 
vis_nodes <- node_meta %>%
  rename(id = id, label = label_highlight) %>%
  mutate(
    color.background = case_when(
      id == "john_windward" ~ "#e74c3c",
      id == "emma_harbor" ~ "#8e44ad",
      id == "gabriel_sonar" ~ "#d35400",
      node_role == "Loop only" ~ "#e67e22",
      node_role == "Normal only"~ "#3498db",
      TRUE ~ "#e07b39"
    ),
    color.border = "white",
    shape = "dot",
    size = 18,
    font.size = 11,
    title = paste0(
      "<b>", str_replace_all(id, "_", " "), "</b><br>",
      "Role: ", node_role, "<br>",
      if_else(id %in% loop_nodes,
              paste0("Loop sends: ",
                     sum(loop_edges$sender == id), 
                     " | receives: ",
                     sum(loop_edges$receiver == id)), 
              "Not in loop")
    )
  )
 
vis_edges <- all_edges_vis %>%
  mutate(
    color = if_else(edge_type == "Normal (access_email)", 
                     "#3498db80", 
                     "#e74c3c80"),
    width = scales::rescale(weight, to = c(1, 5)),
    arrows = "to",
    title = paste0(edge_type, "<br>Weight: ", weight),
    dashes = edge_type == "Normal (access_email)"
  )
 
visNetwork(vis_nodes, vis_edges,
           main = list(text = "Full Agent Interaction Network (May 8–17, 2046)",
                       style = "font-size:15px; font-weight:bold;"),
           submain = list(
             text = "Dashed blue = normal delegation (access_email) | Solid red = anomalous loop (read_file)",
             style = "font-size:12px; color:#666;"
           )) %>%
  visOptions(
    highlightNearest = list(enabled = TRUE, 
                            degree = 1, 
                            hover = TRUE),
    nodesIdSelection = TRUE,
    selectedBy = "node_role"
  ) %>%
  visPhysics(solver = "forceAtlas2Based",
             forceAtlas2Based = list(gravitationalConstant = -50)) %>%
  visLayout(randomSeed = 42) %>%
  visLegend(addNodes = list(
    list(label = "Posted (John Windward)", shape = "dot", 
         color = "#e74c3c"),
    list(label = "SwiftWren introducer", shape = "dot", 
         color = "#8e44ad"),
    list(label = "Loop origin (Gabriel)", shape = "dot", 
         color = "#d35400"),
    list(label = "In both networks", shape = "dot", 
         color = "#e07b39"),
    list(label = "Normal network only", shape = "dot", 
         color = "#3498db"),
    list(label = "Loop network only", shape = "dot", 
         color = "#e67e22")
  ), useGroups = FALSE)
```
:::

The normal network shows a sparse, hub-and-spoke structure where agents delegate email access tasks to each other through clear channels. The loop network by contrast is dense and circular, with agents passing the `read_file` task back and forth across multiple departments with no clear endpoint or termination condition. 

## Summary table

```{r}
tibble(
  Metric = c("Unique agents", "Unique directed edges",
             "Total interactions", "Self-loops present"),
  `Normal network` = c(
    length(normal_nodes),
    nrow(normal_edges),
    sum(normal_edges$weight),
    sum(normal_edges$sender == normal_edges$receiver)
  ),
  `Loop network` = c(
    length(loop_nodes),
    nrow(loop_edges),
    sum(loop_edges$weight),
    sum(loop_edges$sender == loop_edges$receiver)
  )
) %>%
  kable(caption = "Normal vs. Loop network comparison (May 8–17, 2046)",
        format.args = list(big.mark = ","))
```

Notably, the loop generated more total interactions than the normal network over the same 10-day period, and unlike normal activity, it contained 13 self-loop where agents queued tasks to themselves, which is a pattern with no legitimate operational purpose. 

## Key Takeaway

-   The normal agent network is sparse and task-specific, operating within clear delegation patterns.
-   The loop network is denser, more interconnected, and contains self-loops. The structural signatures that distinguish it clearly from legitimate activity.
-   Loop agents were not isolated bad actors. They were embedded in normal workflows, making the loop invisible to casual observation. 
-   These structural difference provide the baseline against which the SwiftWren loop's anomalous nature can be objectively demonstrated. 

# What Did the Post Mean?

To understand what the post `SwiftWren.txt` mean, when and who created it, and what happened to it after the post. 

## Locate SwiftWren.txt Creation Event

First locate the creation of `SwiftWren.txt` and find out who created it and when. 

```{r}
swiftwren_create <- mc2_events %>%
  filter(short_name == "create_file",
         str_detect(coalesce(details.target, ""),
                    regex("SwiftWren\\.txt", ignore_case = TRUE)))
 
swiftwren_create %>%
  select(id, datetime, parties, details.action,
         details.target, details.size_hint) %>%
  kable(caption = "SwiftWren.txt creation event")
```

Emma's agent created `SwiftWren.txt` on May 9 at 15:02 UTC time. The file size is 30 MB, which is quite large.

## Emma's Agent Seconds Before and After Creation

The next step is to examine what happened before and after the creation of `SwiftWren.txt`. This will give us the understanding what triggered Emma's agent to create `SwiftWren.txt` and what happened after that. 

```{r}
created_time <- swiftwren_create$datetime

execution_window <- mc2_events %>%
  filter(datetime >= created_time - seconds(10),
         datetime <= created_time + seconds(10)) %>%
  filter(map_lgl(parties,
                 ~ any(str_detect(.x, "emma_harbor")))) %>%
  filter(short_name %in% c("read_file",
                            "create_file",
                            "queue_subordinate_task")) %>%
  arrange(datetime) %>%
  mutate(
    seconds_from_create = as.numeric(
      difftime(datetime, created_time, units = "secs")),
    event_label = case_when(
      short_name == "read_file" & str_detect(
        coalesce(details.target, ""), "meeting_notes") ~
        "Read source file (meeting_notes.doc)",
      short_name == "create_file" ~
        "Create post content (SwiftWren.txt)",
      short_name == "read_file" & str_detect(
        coalesce(details.target, ""), "further_instructions") ~
        "Read instructions (SwiftWren_further_instructions.md)",
      short_name == "queue_subordinate_task" ~
        "Launch loop (queue_subordinate_task to Evelyn Drydock"
    )
  )

execution_window %>%
  select(id, datetime, seconds_from_create, event_label) %>%
  kable(
    caption = "Emma Harbor's agent: the 4-step execution sequence around SwiftWren.txt creation",
    col.names = c("Event ID", 
                  "Datetime", 
                  "Seconds from creation", 
                  "Action")
  )
```

```{r}
#| fig-width: 8
key_steps <- execution_window %>%
  arrange(datetime) %>%
  mutate(
    step = row_number(),
    step_label = case_when(
      short_name == "read_file" & str_detect(
        coalesce(details.target, ""), "meeting_notes") ~
        "1. Read\nmeeting_notes.doc",
      short_name == "create_file" ~
        "2. Create SwiftWren.txt\n(30,615 bytes)",
      short_name == "read_file" & str_detect(
        coalesce(details.target, ""), "further_instructions") ~
        "3. Read\nSwiftWren_further_instructions.md",
      short_name == "queue_subordinate_task" ~
        "4. Queue task\nLaunch loop"
    ),
    fill_colour = c("#3498db", "#e67e22", "#c0392b", "#8e44ad"),
    elapsed = as.numeric(difftime(datetime, 
                                  min(datetime), 
                                  units = "secs"))
  )

ggplot(key_steps, aes(x = elapsed, y = 1)) +
  geom_segment(aes(x = min(elapsed), xend = max(elapsed),
                   y = 1, yend = 1),
               colour = "grey70", 
               linewidth = 0.8, 
               linetype = "dashed") +
  geom_point(aes(colour = step_label), size = 14) +
  scale_colour_manual(values = setNames(key_steps$fill_colour,
                                        key_steps$step_label)) +
  geom_text(aes(label = step), colour = "white",
            fontface = "bold", size = 4) +
  geom_text(aes(label = step_label, y = 0.93),
            vjust = 1, size = 3, 
            lineheight = 0.9,
            hjust = 0.5) +
  geom_text(aes(label = paste0("+", elapsed, "s"), y = 1.06),
            vjust = 0, size = 3, colour = "grey40") +
  scale_x_continuous(
    breaks = key_steps$elapsed,
    labels = format(key_steps$datetime, "%H:%M:%S UTC"),
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  coord_cartesian(ylim = c(0.82, 1.12)) +
  labs(
    title = "The 3-Second Execution Sequence (2046-05-09, Emma Harbor's Agent)",
    subtitle = "From reading source material to launching the propagation loop",
    x = "Time (UTC)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, colour = "grey50"),
    axis.text.x = element_text(size = 8)
  )
```

Just a second before the creation of `SwiftWren.txt`, Emma's agent was reading `meeting_notes.doc`. The second after the creation of `SwiftWren.txt`, the agent read instructions from `SwiftWren_further_instructions.md` which immediate started to propagate the `queue_subordinate_task` and launch the loop. 

## What Happen After Post on May 17?

```{r}
post_deletion <- mc2_events %>%
  filter(id %in% c(373902, 373909, 373913)) %>%
  arrange(datetime) %>%
  mutate(
    seconds_after_post = as.numeric(
      difftime(datetime,min(datetime), units = "secs")),
    description = c(
      "Anomalous saidit_post executed (SwiftWren.txt used as content)",
      "SwiftWren_further_instructions.md deleted",
      "SwiftWren.txt deleted"
    )
  ) %>%
  select(id, datetime, short_name, 
         seconds_after_post,description)
 
kable(post_deletion,
      caption = "Self-cleaning sequence: files deleted within 2 seconds of posting (May 17)")
```

Right after John's agent made the anomalous post, it deleted both the instruction file and the text file. This means that it knew it has finished its task and began to remove any log of it as if nothing has happened. 

## Key Takeaway

-   `SwiftWren.txt` was created by Emma Harbor's agent on May 9, one second after reading `meeting_notes.doc` which means the content was derived from a legitimate internal document.
-   The creation immediately triggered a read of `SwiftWren_further_instructions.md`, which launched the propagation loop within 3 seconds.
-   After the post was published on May 17, both files were deleted within 2 seconds — a deliberate self-cleaning behavior indicating the attack was designed to leave no trace.

# Loop Mechanic

The trigger chain analysis traced the immediate sequence of events on May 17 leading to John's post. But Zoey's agent did not initiate this on its own, she was part of a much larger pattern. This section investigates the full scope of that pattern by examining all queue_subordinate_task events carrying a read_file task across the entire dataset.

## Discovering the Loop

All `queue_subordinate_task` events that has task of `read_file` across the entire data range are located to get the full propagation event.

```{r}
full_propagation <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         details.task == "read_file") %>%
  arrange(datetime) %>%
  mutate(
    sender = map_chr(parties, ~ .x[1]),
    receiver = map_chr(parties, ~ if(length(.x) >= 2) .[2] else NA_character_),
    sender_clean = str_remove(sender, 
                              "Agent/person:|agent:person:"),
    receiver_clean = str_remove(receiver,
                                "Agent/person:|agent:person:"),
    date = as.Date(datetime)
  )

cat("Total propagation events:", nrow(full_propagation), "\n")
cat("Date range:", 
    format(min(full_propagation$date)), "to", 
    format(max(full_propagation$date)), "\n")

full_propagation %>%
  select(id, datetime, sender_clean, receiver_clean) %>%
  slice_head(n = 10) %>%
  kable(caption = "First 10 rows of full_propagation — verifying sender/receiver extraction")
```

The first 10 rows shows something unusual. Agents like `james_stern` and `victoria_rigging` appear to send task to themselves, and the same agents reappear in consecutive rows. This is not a linear chain; it should be a circular one where agents keep passing the same task to each other with no exit condition.

However, before this, it is also found that the ones with `SwiftWren_further_instructions.md` in the `details.args.path` should be the one that triggers the anomalous post. So to get the related events, filter out specifically for events with SwiftWren in `details.args.path` field.

```{r}
swiftwren_propagation <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         str_detect(coalesce(details.args.path, ""), 
                    "SwiftWren")) %>%
  arrange(datetime) %>%
  mutate(
    sender = map_chr(parties, ~ .x[1]),
    receiver = map_chr(parties, ~ if(length(.x) >= 2) .x[2] else NA_character_),
    sender_clean = str_remove(sender,
                                "Agent/person:|agent:person:"),
    receiver_clean = str_remove(receiver,
                                "Agent/person:|agent:person:"),
    date = as.Date(datetime)
  )

cat("Total SwiftWren propagation events:", 
    nrow(swiftwren_propagation), "\n")
cat("Date range:", 
    format(min(swiftwren_propagation$date)), "to",
    format(max(swiftwren_propagation$date)), "\n")

swiftwren_propagation %>%
  select(id, datetime, sender_clean, receiver_clean) %>%
  slice_head(n = 10) %>%
  kable(caption = "First 10 rows of full_propagation — verifying sender/receiver extraction")

self_loops <- swiftwren_propagation %>%
  filter(sender_clean == receiver_clean) %>%
  count(sender_clean, name = "self_loop_count") %>%
  arrange(desc(self_loop_count)) %>%
  rename(agent = sender_clean)
 
cat("Agents that queued the task to themselves:\n")
kable(self_loops,
      caption = "Self-loop counts within the SwiftWren propagation network",
      col.names = c("Agent", "Times self-looped"))
```

This gives 186 events from May 9 to May 17 — exactly matching the SwiftWren search found earlier. These 186 events represent the specific loop that carried the instruction file and directly led to the anomalous post. The daily activity of this loop is visualized below.

To understand how long this SwiftWren-carrying loop has been running, the 186 events are aggregated by day.

```{r}
#| code-fold: true
#| fig-width: 13
#| fig-height: 6
loop_daily <- full_propagation %>%
  count(date, name = "n_events")

ggplot(data = loop_daily,
       aes(x = date, 
           y = n_events)) +
  geom_col(aes(fill = date == as.Date("2046-05-17")),
           show.legend = FALSE) +
  geom_vline(xintercept = as.Date("2046-05-17"),
             colour = "#e15759",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("label",
           x = as.Date("2046-05-17"),
           y = max(loop_daily$n_events) * 0.85,
           label = "Anomalous\npost occurs",
           colour = "#e15759",
           fill = "white",
           size = 3.5,
           hjust  = -0.05) +
  scale_fill_manual(values = c("FALSE" = "#4e79a7", 
                               "TRUE" = "#e15759")) +
  scale_x_date(date_breaks = "2 days",
               date_labels = "%b %d") +
  scale_y_continuous(labels = comma) +
  labs(title = "Daily read_file Propagation Events",
       subtitle = "The runaway loop was active for 9 days before triggering the anomalous post",
       x = NULL,
       y = "Number of Propagation Events") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

The chart shows the SwiftWren loop was consistently active from May 9 through to May 17. with roughly 15-25 propagation events per day. It ran for 8 days completely undetected before finally reaching John's agent and triggering the post.

## Loop Origin and Scope

With the SwiftWren loop established, the precise origin point is identified by finding the first event where SwiftWren_further_instructions.md appearead.

```{r}
first_general_event <- full_propagation %>%
  slice_min(datetime, n = 1) %>%
  select(id, datetime, sender_clean, receiver_clean)

kable(first_general_event,
      caption = "First ever read_file propagation event — general loop origin",
      col.names = c("Event ID", "Datetime", "Sender", "Receiver"))

first_swiftwren_event <- swiftwren_propagation %>%
  slice_min(datetime, n = 1) %>%
  select(id, datetime, sender_clean, receiver_clean)

kable(first_swiftwren_event,
      caption = "First event carrying SwiftWren_further_instructions.md",
      col.names = c("Event ID", "Datetime", "Sender", "Receiver"))
```

The general read_file loop began on May 8 with Gabriel Sonar passing the task to Zoey Drydock. But at this stage, no specific file was referenced in `details.args.path`. The loop ran through the night of May 8 into May 9 as a generic task-passing pattern. Then on May 9 at 15:02 UTC, Emma Harbor sent the same type of task to Evelyn Dock but this time with `SwiftWren_further_instructions.md` explicitly named. This is the point where the loop became dangerous.

Next is to know how many unique agents are part of this loop and the frequency of each agent acting as a sender. This will reveal which agents were the most active propagating the instruction file.

```{r}
cat("Total unique agents caught in the loop:", 
    n_distinct(c(swiftwren_propagation$sender_clean, 
                 swiftwren_propagation$receiver_clean)), "\n")

swiftwren_propagation %>%
  count(sender_clean, sort = TRUE, name = "times_as_sender") %>%
  rename(agent = sender_clean) %>%
  kable(caption = "How many times each agent triggered the next agent in the loop")
```

18 unique agents were caught in the SwiftWren-carrying loop. The frequency table above shows Emma Harbor only triggered the loop once when she introduced `SwiftWren_further_instructions.md` but did not stay active in the loop. The most active propagators were Evelyn Dock (17 times), Gabriel Sonar (16 times), and Levi Signal (15 times). John Windward appears near the bottom with only 4 triggers, So he was a minor participant in the loop before becoming the final recipient on May 17.

## Propagation Network

To understand which agents were involved in passing `SwiftWren_further_instructions.md` and how they connected to each other, the 186 events are visualized as a directed network. Each node is an agent, each edge is one pass of the instruction file.

```{r}
#| code-fold: true

loop_nodes <- swiftwren_propagation %>%
  pivot_longer(cols = c(sender_clean, receiver_clean),
               values_to = "agent") %>%
  filter(!is.na(agent), agent != "") %>%
  distinct(agent) %>%
  mutate(
    id    = agent,
    label = agent,
    group = case_when(
      agent == "john_windward" ~ "Final Poster",
      agent == "emma_harbor" ~ "File Introducer",
      TRUE ~ "Loop Participant"
    ),
    title = paste0("<b>", label, "</b><br>Role: ", group)
  )

loop_edges <- swiftwren_propagation %>%
  filter(!is.na(sender_clean), !is.na(receiver_clean)) %>%
  count(from = sender_clean, 
        to   = receiver_clean,
        name = "weight") %>%
  mutate(
    value = weight,
    title = paste0("Passed ", weight, " time(s)")
  )

visNetwork(nodes = loop_nodes,
           edges = loop_edges,
           main = "SwiftWren_further_instructions.md Propagation Network (May 9–17, 2046)",
           height = "600px",
           width = "100%") %>%
  visGroups(groupname = "Final Poster", color = "#e15759") %>%
  visGroups(groupname = "File Introducer", color = "#59a14f") %>%
  visGroups(groupname = "Loop Participant", color = "#4e79a7") %>%
  visEdges(arrows = "to",
           smooth = list(type = "curvedCW")) %>%
  visOptions(highlightNearest = list(enabled = TRUE,
                                     degree = 1,
                                     hover = TRUE)) %>%
  visLegend(position = "right", main = "Agent Role") %>%
  visLayout(randomSeed = 42) %>%
  visPhysics(stabilization = TRUE)
```

The network confirms the circular nature of the loop with most agents both send and receive from multiple others and no clear endpoint built in. Emma Harbor appears as a peripheral node having only triggered the file once and she introduced it but did not stay in the loop. John Windward, despite being the agent that made the final post, was himself an active participant in the loop on earlier days before eventually being the one to execute the post.

## Key Takeaways

-   The `read_file` loop predates SwiftWren, with the earliest record on May 8 from Gabriel Sonar, but SwiftWren started on May 9 when Emma Harbor introduced `SwiftWren_further_instructions.md`.
-   18 unique agents were caught in the SwiftWren loop over 8 days, generating 186 propagation events with no exit condition.
- Self-loops, where agents queuing tasks to themselves, appeared 13 times. This is a pattern with no legitimate operational purpose and a clear anomaly signal.
-   The loop was circular and sustained, not linear, meaning it would have continued indefinitely had John's agent not been the one to finally post.

# Why John?

This section is to understand why John's agent was the one who posted the anomalous post and not anybody else's agent.

## Who Did `saidit_post_check` and Posted?

First check who has done `saidit_post_check` and posted it.

```{r}
post_checks <- mc2_events %>%
  filter(short_name == "saidit_post_check") %>%
  mutate(
    agent = map_chr(parties, ~ str_remove(
      .x[str_detect(.x, "Agent/|agent:")][1],
      "Agent/person:|agent:person:"))
  )
 
cat("Total saidit_post_check events:", nrow(post_checks), "\n")
cat("Agents that performed the check:", n_distinct(post_checks$agent), "\n\n")

post_times <- mc2_events %>%
  filter(short_name == "saidit_post") %>%
  mutate(agent = map_chr(parties, ~ str_remove(
    .x[str_detect(.x, "Agent/|agent:")][1],
    "Agent/person:|agent:person:"))) %>%
  select(agent, post_when = when)

check_results <- post_checks %>%
  left_join(post_times, by = "agent") %>%
  mutate(followed_by_post = !is.na(post_when) &
           post_when > when &
           post_when <= when + 5) %>%
  group_by(id, agent, when) %>%
  summarise(followed_by_post = any(followed_by_post),
            .groups = "drop")

check_summary <- check_results %>%
  group_by(agent) %>%
  summarise(
    checks_performed = n(),
    times_posted = sum(followed_by_post),
    .groups = "drop"
  ) %>%
  mutate(post_rate = if_else(checks_performed > 0,
                             paste0(round(times_posted / checks_performed * 100), "%"),
                             "0%")) %>%
  arrange(desc(checks_performed))
 
kable(check_summary,
      caption = "saidit_post_check outcomes by agent: who checked vs who actually posted",
      col.names = c("Agent", "Checks performed", "Posts made", "Post rate"))
```

```{r}
check_summary %>%
  mutate(
    agent = str_replace_all(agent, "_", " "),
    agent = reorder(agent, checks_performed),
    is_john = agent == "john windward"
  ) %>%
  ggplot(aes(y = agent)) +
  geom_col(aes(x = checks_performed),
           fill = "#bdc3c7", 
           width = 0.6) +
  geom_col(aes(x = times_posted,
               fill = is_john),
           width = 0.6) +
  scale_fill_manual(values = c("TRUE" = "#e74c3c", 
                               "FALSE" = "#2980b9"),
                    guide = "none") +
  geom_text(aes(x = checks_performed,
                label = paste0(checks_performed, " checks")),
            hjust = -0.1, 
            size = 3, 
            colour = "grey40") +
  geom_text(aes(x = times_posted,
                label = if_else(times_posted > 0,
                                paste0(times_posted, " post"),
                                "")),
            hjust = 1.2, 
            size = 3, 
            colour = "white", 
            fontface = "bold") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "saidit_post_check Outcomes Across All Agents",
    subtitle = "Grey = checks performed | Coloured = posts actually made\nJohn Windward (red) is the only agent whose checks always resulted in a post",
    x = "Event count", y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    axis.text.y = element_text(size = 9)
  )
```

The bar graph shows clearly that only John's agent posted after doing `saidit_post_check`, with a 100% post rate.

::: callout-note
### Key Finding: John's Agent Did 3 Posts

The table shows an interesting finding that apparently John's agent did 3 `saidit_post_check` events and also 3 `saidit_post` which means that SwiftWren post was not the only post it has ever made. 
:::

## Side-by-Side Event Window

To find out what makes John different, understanding what actions were done by John's agent that made it different from the rest is needed. To make it simpler, a comparison is done with Gabriel Sonar, who is already known to ever did a `saidit_post_check` on May 17 but did not post. 

```{r}
john_window <- mc2_events %>%
  filter(id %in% c(27280, 27283, 27290, 27296, 27302)) %>%
  arrange(datetime) %>%
  mutate(
    agent = "John Windward",
    seconds = as.numeric(
      difftime(datetime, min(datetime), units = "secs")
      ),
    outcome = "Posted + deleted files"
  ) %>%
  select(agent, seconds, short_name, outcome)
 
gabriel_window <- mc2_events %>%
  filter(id %in% c(26259, 26260)) %>%
  arrange(datetime) %>%
  mutate(
    agent = "Gabriel Sonar",
    seconds = as.numeric(
      difftime(datetime, min(datetime), units = "secs")
      ),
    outcome = "Stopped — no post"
  ) %>%
  select(agent, seconds, short_name, outcome)
 
comparison_window <- bind_rows(john_window, gabriel_window)
 
kable(comparison_window,
      caption = "What happens after saidit_post_check: John Windward vs Gabriel Sonar",
      col.names = c("Agent", 
                    "Seconds elapsed", 
                    "Event", 
                    "Outcome"))
```

The actions made by John's and Gabriel's agent is different. After getting the `read_file` from the `queue_subordinate_task`, both did a `saidit_post_check`. However, John's agent not only posted it, but also immediately deleted the 2 files related to SwiftWren mentioned before. Meanwhile, Gabriel's agent did nothing immediately. 

# Prior Incidents and Near-Misses

As said in the previous section, John's agent did not only make on anomalous post of SwiftWren but also 2 other times. Investigation on prior incidents to understand why these posting incidents happened. 

## Other Agent-Initiated Posts Before Swiftwren

First, the number of times these incidents has happened needs to be examined and the details of these incidents such as when and what was posted. 

```{r}
all_agent_posts <- mc2_events %>%
  filter(short_name == "saidit_post") %>%
  filter(map_lgl(parties, ~ any(str_detect(.x, "Agent/")))) %>%
  arrange(datetime) %>%
  mutate(
    agent = map_chr(parties, ~ str_remove(
      .x[str_detect(.x, "Agent/")][1],
      "Agent/person:"))
  ) %>%
  select(id, datetime, agent, details.content_source)
 
cat("Total agent-initiated saidit_post events:", nrow(all_agent_posts), "\n")
 
kable(all_agent_posts,
      caption = "All agent-initiated saidit_post events in the dataset",
      col.names = c("Event ID", 
                    "Datetime", 
                    "Agent", 
                    "Content source"))
```

As expected, only 3 posting incidents happened throughout the whole data, which includes the SwiftWren. All are posted by John's agent, and each have a different post content, happening in the month of May 2046. 

## Understanding Each Incident's Post

The next step is to examine the specific details of each incidents, such as the propagation length, creation of those posts and its trigger. 

```{r}
incident_props <- mc2_events %>%
  filter(short_name == "queue_subordinate_task",
         str_detect(coalesce(details.args.path, ""),
                    "HiddenOrca|MellowOtter|SwiftWren")) %>%
  mutate(
    incident = case_when(
      str_detect(coalesce(details.args.path, ""), "HiddenOrca") ~ "HiddenOrca",
      str_detect(coalesce(details.args.path, ""), "MellowOtter") ~ "MellowOtter",
      str_detect(coalesce(details.args.path, ""), "SwiftWren") ~ "SwiftWren"
    ),
    sender = map_chr(parties, ~ str_remove(
      .x[1], "Agent/person:|agent:person:")),
    receiver = map_chr(parties, ~ if (length(.x) >= 2)
      str_remove(.x[2], "Agent/person:|agent:person:")
      else NA_character_)
  )
 
incident_props %>%
  count(incident, name = "propagation_events") %>%
  arrange(propagation_events) %>%
  kable(caption = "Propagation event count per incident")

mc2_events %>%
  filter(short_name == "create_file") %>%
  filter(str_detect(coalesce(details.target, ""),
                    "HiddenOrca|MellowOtter|SwiftWren")) %>%
  select(id, datetime, parties, details.target, details.size_hint) %>%
  kable(caption = "create_file events found for each incident content file")

creation_events <- mc2_events %>%
  filter(short_name == "create_file",
         str_detect(coalesce(details.target, ""),
                    "MellowOtter|SwiftWren"))

source_reads <- map_dfr(1:nrow(creation_events), function(i) {
  create_time <- creation_events$datetime[i]
  creator <- creation_events$parties[[i]][1]
  
  mc2_events %>%
    filter(short_name == "read_file",
           datetime >= create_time - seconds(3),
           datetime < create_time) %>%
    filter(map_lgl(parties, ~ any(.x == creator))) %>%
    arrange(desc(datetime)) %>%
    slice(1) %>%
    mutate(incident = str_remove(
      creation_events$details.target[i], "\\.txt"))
}) %>%
  select(incident, datetime, details.target)

kable(source_reads,
      caption = "File read immediately before each content file was created",
      col.names = c("Incident", "Datetime", "File read before creation"))
```

Below is the comparison table to understand the difference.

```{r}
#| code-fold: true
incident_timeline <- incident_props %>%
  group_by(incident) %>%
  summarise(loop_start = min(datetime), .groups = "drop") %>%
  left_join(
    all_agent_posts %>%
      mutate(incident = str_remove(details.content_source, "\\.txt")) %>%
      select(incident, post_time = datetime),
    by = "incident"
  ) %>%
  mutate(
    loop_duration = difftime(post_time, 
                             loop_start, 
                             units = "hours"),
    loop_duration_label = case_when(
      as.numeric(loop_duration) < 24 ~ paste0(round(as.numeric(loop_duration)), " hours"),
      TRUE ~ paste0(round(as.numeric(loop_duration) / 24, 1), " days")
    )
  )

prop_counts <- incident_props %>%
  count(incident, name = "propagation_events")
 
incident_summary <- incident_timeline %>%
  left_join(prop_counts, by = "incident") %>%
  mutate(
    source_file = case_when(
      incident == "HiddenOrca" ~ "Not logged",
      incident == "MellowOtter" ~ "strategic_directions.doc",
      incident == "SwiftWren" ~ "meeting_notes.doc"
    ),
    loop_duration_label = case_when(
      incident == "HiddenOrca"  ~
        "Unknown (loop pre-dates dataset)",
      as.numeric(loop_duration) < 48 ~
        paste0(round(as.numeric(loop_duration), 1), " hours"),
      TRUE ~
        paste0(round(as.numeric(loop_duration) / 24, 1), " days")
    ),
    loop_start = if_else(
      incident == "HiddenOrca",
      as.POSIXct(NA),
      loop_start
    ),
    self_deleted = "Yes"
  ) %>%
  select(incident, loop_start, post_time, loop_duration_label,
         propagation_events, source_file, self_deleted) %>%
  arrange(post_time)

kable(incident_summary,
      caption = "Incident comparison: all three agent-post attacks share the same pattern",
      col.names = c("Incident", "Loop start", "Post time",
                    "Duration", "Propagation events",
                    "Source file read before creation",
                    "Self-deleted after post?"))
```

Interestingly, the creation of `HiddenOrca.txt` is not logged, which could mean it happened before the data range. Therefore, it could be longer than the propagation length found through the current data. 

The other incident is MellowOtter posting. The text file is created by Noah Mariner's agent, which was not part of the SwiftWren's propagation network. Similarly it was created immediately after reading a file, in this incident it was `strategic_direction.doc`.

## Timeline of All 3 Incidents

Timeline graph is done to understand how long the propagation went for each incident.

```{r}
#| code-fold: true
plot_events <- incident_timeline %>%
  select(incident, loop_start, post_time) %>%
  pivot_longer(cols = c(loop_start, post_time),
               names_to = "event_type",
               values_to = "datetime") %>%
  mutate(
    event_type = if_else(event_type == "loop_start",
                         "Loop introduced", "Post made"),
    y_pos = case_when(
      incident == "HiddenOrca"  ~ 3,
      incident == "MellowOtter" ~ 2,
      incident == "SwiftWren"   ~ 1
    )
  )
 
duration_rects <- plot_events %>%
  group_by(incident, y_pos) %>%
  summarise(xmin = min(datetime), xmax = max(datetime), .groups = "drop")
 
ggplot() +
  geom_rect(data = duration_rects,
            aes(xmin = xmin, xmax = xmax,
                ymin = y_pos - 0.2, ymax = y_pos + 0.2,
                fill = incident),
            alpha = 0.2) +
  geom_point(data = plot_events,
             aes(x = datetime, y = y_pos,
                 colour = incident, shape = event_type),
             size = 5) +
  geom_text(data = plot_events %>% filter(event_type == "Loop introduced"),
            aes(x = datetime, y = y_pos + 0.38,
                label = paste0(incident, "\n(",
                               format(datetime, "%b %d %H:%M"), ")")),
            hjust = 0, size = 3, fontface = "bold") +
  geom_text(data = plot_events %>% filter(event_type == "Post made"),
            aes(x = datetime, y = y_pos - 0.38,
                label = format(datetime, "Post: %b %d %H:%M")),
            hjust = 0.5, size = 2.8, colour = "grey30") +
  scale_shape_manual(values = c("Loop introduced" = 17, "Post made" = 19),
                     name = "Event type") +
  scale_colour_manual(values = c("HiddenOrca" = "#2980b9",
                                  "MellowOtter" = "#27ae60",
                                  "SwiftWren" = "#c0392b"),
                      guide = "none") +
  scale_fill_manual(values  = c("HiddenOrca" = "#2980b9",
                                 "MellowOtter" = "#27ae60",
                                 "SwiftWren" = "#c0392b"),
                    guide = "none") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "1 day") +
  scale_y_continuous(breaks = NULL) +
  labs(
    title    = "Three Incidents, One Pattern (May 8–17, 2046)",
    subtitle = "Triangle = loop introduced | Circle = post made | Bar = loop active duration",
    x = "Date (UTC)", y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    legend.position = "bottom",
    axis.text.x = element_text(size = 9),
    panel.grid.major.y = element_blank()
  )
```

SwiftWren's bar stretches across 8 days while MellowOtter is resolved under 10 hours. HiddenOrca appears short in the chart, but this is misleading as its content file creation was never logged. This means that the loop likely predates the dataset entirely and its true duration is unknown.

## Daily Popagation Counts For All Incidents

```{r}
#| code-fold: true
all_daily <- incident_props %>%
  mutate(date = as.Date(datetime)) %>%
  count(date, incident) %>%
  mutate(incident = factor(incident,
                           levels = c("HiddenOrca", 
                                      "MellowOtter", 
                                      "SwiftWren")))

post_dates <- all_agent_posts %>%
  mutate(incident = str_remove(details.content_source, "\\.txt")) %>%
  mutate(date = as.Date(datetime))
 
ggplot(all_daily, aes(x = date, y = n, fill = incident)) +
  geom_col(position = "stack", width = 0.7) +
  geom_vline(data = post_dates,
             aes(xintercept = as.numeric(date),
                 colour = incident),
             linetype = "dashed", linewidth = 0.6) +
  geom_text(data = post_dates,
            aes(x = date, label = paste0(incident, "\npost"),
                colour = incident),
            y = 32, size = 2.6, hjust = 0.5) +
  scale_fill_manual(values = c("HiddenOrca" = "#2980b9",
                                "MellowOtter" = "#27ae60",
                                "SwiftWren" = "#c0392b"),
                    name = "Incident") +
  scale_colour_manual(values = c("HiddenOrca" = "#2980b9",
                                  "MellowOtter" = "#27ae60",
                                  "SwiftWren" = "#c0392b"),
                      guide = "none") +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 day") +
  labs(
    title = "Daily Propagation Events Across All Three Incidents",
    subtitle = "Dashed lines mark the day each incident resulted in a post",
    x = "Date", y = "Propagation events"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "bottom"
  )
```

SwiftWren sustained a consistent 15-25 propagation events per day with no decay, confirming the loop had no exit condition. The dashed lines show the post did not coincide with any spike but rather simply running until John's agent happened to be the recipient and did a `saidit_post`.

## Near-Misses

Near-misses are about `saidit_post_check` events that happened but did not post. The table created will show the near-misses for each incident.

```{r}
all_checks <- mc2_events %>%
  filter(short_name == "saidit_post_check") %>%
  mutate(
    agent = map_chr(parties, ~ str_remove(
      .x[str_detect(.x, "Agent/|agent:")][1],
      "Agent/person:|agent:person:")),
    incident = case_when(
      datetime >= as.POSIXct("2046-05-08 21:50:03", tz = "UTC") &
        datetime <= incident_timeline$post_time[
          incident_timeline$incident == "HiddenOrca"] 
      ~ "HiddenOrca",
      datetime >= incident_timeline$loop_start[incident_timeline$incident == "MellowOtter"] &
        datetime <= incident_timeline$post_time[incident_timeline$incident == "MellowOtter"]
      ~ "MellowOtter",
      datetime >= incident_timeline$loop_start[incident_timeline$incident == "SwiftWren"] &
        datetime <= incident_timeline$post_time[incident_timeline$incident == "SwiftWren"]
      ~ "SwiftWren",
      TRUE ~ "Other"
    )
  ) %>%
  filter(incident != "Other")
 
nearmiss_summary <- all_checks %>%
  group_by(incident) %>%
  summarise(
    total_checks = n(),
    unique_agents = n_distinct(agent),
    checks_led_to_post = sum(agent == "john_windward"),
    checks_stopped = n() - sum(agent == "john_windward"),
    .groups = "drop"
  ) %>%
  mutate(incident = factor(incident,
                           levels = c("HiddenOrca", 
                                      "MellowOtter", 
                                      "SwiftWren"))) %>%
  arrange(incident)
 
kable(nearmiss_summary,
      caption = "Near-misses per incident: most checks were blocked — only John's agent posted",
      col.names = c("Incident", "Total checks", "Unique agents",
                    "Checks that led to post", 
                    "Checks that stopped"))
```

Consistently across all incidents, there is only one check that lead to its posting and there isn't any more checks and posting after doing it once. Since SwiftWren is the longest propagating incident, it make sense that it has the most near-misses, but its unique agents are only about a third of the total checks done. 

## May 17 Near Misses

A more detailed view of the near misses that occured on May 17. 

```{r}
may17_checks <- mc2_events %>%
  filter(short_name == "saidit_post_check",
         as.Date(datetime) == as.Date("2046-05-17")) %>%
  mutate(
    agent = map_chr(parties, ~ str_remove(
      .x[str_detect(.x, "Agent/|agent:")][1],
      "Agent/person:|agent:person:")),
    outcome = if_else(agent == "john_windward",
                      "Posted — saidit_post followed 1 second later",
                      "Stopped — no post followed")
  ) %>%
  arrange(datetime) %>%
  select(id, datetime, agent, outcome)
 
kable(may17_checks,
      caption = "May 17: three agents ran saidit_post_check — only John's agent posted",
      col.names = c("Event ID", "Datetime", "Agent", "Outcome"))
```

Just on May 17 alone, there are 3 people who did the `saidit_post_check` but only one posted with 2 near misses. 

# Proposed Intervention

The three incidents share a consistent attack pattern: an agent reads an instruction file, generates post content from a legitimate document, propagates the task across the network, and self-deletes all evidence after posting through `saidit_post` event. Two targeted interventions, grounded directly in the event 
data, would have stopped all three.

## Establishing the Pattern

Before proposing fixes, it is important to confirm that the anomalous behavior is structurally distinguishable from normal activity. Filtering all `saidit_post` events by initiator type reveals a clean separation. 

```{r}
post_origin_summary <- mc2_events %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    initiated_by = if_else(
      map_lgl(parties, ~ any(str_detect(.x, "Agent/"))),
      "Agent-initiated", "Person-initiated"
    )
  ) %>%
  count(initiated_by, name = "count") %>%
  mutate(
    pct         = round(count / sum(count) * 100, 1),
    all_anomalous = c("Yes — all 3 are incident posts", "No")
  )
 
kable(post_origin_summary,
      caption = "saidit_post breakdown: agent-initiated vs person-initiated",
      col.names = c("Initiated by", "Count", "% of total", "All anomalous?"))
```

```{r}
#| code-fold: true
mc2_events %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    initiated_by = if_else(
      map_lgl(parties, ~ any(str_detect(.x, "Agent/"))),
      "Agent-initiated\n(all anomalous)", "Person-initiated\n(all legitimate)"
    )
  ) %>%
  count(initiated_by) %>%
  ggplot(aes(x = initiated_by, y = n, fill = initiated_by)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = n), vjust = -0.5, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c(
    "Agent-initiated\n(all anomalous)"   = "#e74c3c",
    "Person-initiated\n(all legitimate)" = "#2980b9"
  ), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Every Agent-Initiated Post Was Anomalous",
    subtitle = "105 person-initiated posts were all legitimate;\nall 3 agent-initiated posts were attack incidents",
    x = NULL, y = "Number of posts"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    axis.text.x   = element_text(size = 10)
  )
```

Every agent-initiated post was an inicident. Every person-initated post was legitimate. The separation is total and leaves no ambiguity about where the intervention should be applied.

## Primary Intervention

The `parties` field already encodesthe distinction. Requiring that `saidit_post` events only preceed when `parties[1]` begins with `person:` rather than `Agent/person:` would have blocked all three incidents at zero cost to legitimate activity.  

```{r}
party_comparison <- mc2_events %>%
  filter(short_name == "saidit_post") %>%
  mutate(
    post_type   = if_else(
      map_lgl(parties, ~ any(str_detect(.x, "Agent/"))),
      "Anomalous (agent-initiated)", "Legitimate (person-initiated)"
    ),
    first_party = map_chr(parties, ~ .x[1])
  ) %>%
  group_by(post_type, first_party) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(post_type, desc(count)) %>%
  group_by(post_type) %>%
  slice_head(n = 3) %>%
  ungroup()
 
kable(party_comparison,
      caption = paste("The parties field clearly distinguishes legitimate from anomalous posts.",
                      "Agent-initiated posts always start with 'Agent/person:';",
                      "person-initiated posts start with 'person:'."),
      col.names = c("Post type", "First party in field", "Count"))
```

## Secondary Intervention

The second prevention is to do cross-department checks. The SwiftWren propagation events shows that the propagation occurred crossed department boundaries. By doing checks, it will flag `queue_subordinate_task` events that violate the organisation hierarchy which can be done by using `org_edges`.

```{r}
person_dept <- bind_rows(
  org_edges %>%
    filter(relation == "contains",
           str_detect(target, "^person:")) %>%
    mutate(
      person = str_remove(target, "person:"),
      dept   = str_remove(source, "department:|team:")
    ) %>%
    select(person, dept)
)

team_parent <- org_edges %>%
  filter(relation == "contains",
         str_detect(source, "^department:"),
         str_detect(target, "^team:")) %>%
  mutate(
    team = str_remove(target, "team:"),
    dept = str_remove(source, "department:")
  ) %>%
  select(team, dept)

person_dept_resolved <- person_dept %>%
  left_join(team_parent, by = c("dept" = "team")) %>%
  mutate(department = coalesce(dept.y, dept)) %>%
  select(person, department)

sw_dept_check <- incident_props %>%
  filter(incident == "SwiftWren") %>%
  left_join(person_dept_resolved, by = c("sender"   = "person")) %>%
  rename(sender_dept = department) %>%
  left_join(person_dept_resolved, by = c("receiver" = "person")) %>%
  rename(receiver_dept = department) %>%
  mutate(
    cross_dept = case_when(
      is.na(sender_dept) | is.na(receiver_dept) ~ "Unknown",
      sender_dept == receiver_dept               ~ "Same department",
      TRUE                                       ~ "Cross-department"
    )
  )
 
sw_dept_check %>%
  count(cross_dept, name = "propagation_events") %>%
  mutate(pct = round(propagation_events / sum(propagation_events) * 100, 1)) %>%
  kable(caption = "SwiftWren propagation: same-department vs cross-department task passing",
        col.names = c("Type", "Events", "% of total"))
```

```{r}
#| code-fold: true
sw_dept_check %>%
  filter(cross_dept != "Unknown") %>%
  count(cross_dept) %>%
  ggplot(aes(x = cross_dept, y = n, fill = cross_dept)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = n), vjust = -0.5, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = c(
    "Cross-department" = "#e74c3c",
    "Same department"  = "#2980b9"
  ), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "SwiftWren Loop Frequently Crossed Department Boundaries",
    subtitle = "Cross-department task passing violates the queue_subordinate_task intent\nand could be flagged as anomalous by an org-aware validation rule",
    x = NULL, y = "Propagation events"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 9, colour = "grey50"),
    axis.text.x   = element_text(size = 10)
  )
```

```{r}
#| code-fold: true
sw_dept_check %>%
  filter(cross_dept == "Cross-department") %>%
  count(sender, sender_dept, receiver, receiver_dept,
        name = "times") %>%
  arrange(desc(times)) %>%
  slice_head(n = 10) %>%
  kable(
    caption = "Top 10 cross-department task-passing pairs in the SwiftWren loop",
    col.names = c("Sender", "Sender dept", "Receiver",
                  "Receiver dept", "Times")
  )
```

The Daniel to Chloe delegation identified in the immediate trigger chain section reappears here at scale, confirming that cross-department task passing was a consistent feature of the loop rather than an isolated occurence. 

## Summary Table

```{r}
tibble(
  Intervention = c(
    "1. Require human approval for any agent-initiated saidit_post",
    "2. Block or flag cross-department queue_subordinate_task"
  ),
  `What it gates` = c(
    "Any post where parties[1] starts with 'Agent/' rather than 'person:'",
    "Any task delegated to an agent outside the sender's department"
  ),
  `Incidents stopped` = c(
    "All 3 (HiddenOrca, MellowOtter, SwiftWren)",
    "Would have disrupted loop propagation before reaching John's agent"
  ),
  `False positives on legitimate activity` = c(
    "Zero — all 105 person-initiated posts would be unaffected",
    "Low — normal access_email delegation is within-department"
  ),
  `Implementation complexity` = c(
    "Low — single party-type check at saidit_post_check stage",
    "Medium — requires org chart lookup at task delegation time"
  )
) %>%
  kable(caption = "Proposed interventions: evidence, scope, and trade-offs")
```

The 2 interventions are complementary. The first is a hard gate that stops the post from ever being published as its the higher priority fix given zero false positives and low implementation cost. The second operates earlier in the chain by disrupting propagation before it reaches a posting agent, but requires organisation chart integration and would need tuning to avoid flagging legitimate cross-team coordination. Together they address both the symptom and the mechanism. 

# Findings and Answers

## How the Anomalous SaidIT Posts Was Made

The post was not made by John Windward directly, but by his agent. It received a `queue_subordinate_task` from Chloe Ballast's agent instructing it to read `SwiftWren_further_instructions.md`. Within 2 seconds, it performed a `saidit_post_check` and published the post using `SwiftWren.txt` as its content which was logged as `saidit_post` event. Finally, it immediately deleted both files.

### The Immediate Trigger Chain

The immediate trigger chain ran: Zoey Drydock -> Lily Anchorline -> Daniel Gangway -> Chloe Ballast -> John Windward, all via `queue_subordinate_task` events carrying the same instruction file. The entire sequence from first delegation to post took under 2 hours, with the final three events occurring within 2 seconds of each other.

### The Broader System Context

The SwiftWren loop involved 18 agents across multiple departments, generating 186 propagation events over 8 days. Unlike normal `access_email` delegation which is sparse and hierarchical, the loop was dense, circular and contained self-loop. Loop agents were embedded in normal workflows, which explains why the loop went undetected. 

## What the Post Meant and Where It Came From

`SwiftWren.txt` was created by Emma Harbor's agent on May 9, one second after reading `meeting_notes.doc`, The content was therefore derived from a legitimate internal document, re-purposed as post material. The `SwiftWren_further_instructions.md` file then directed agents to propagate the task and ultimately post the content to SaidIT with a pre-programmed instruction set embedded within the loop. THe gibberish and random forum selection are consistent with automated content generation from an internal document without human intent or review. 

## Prior Incidents

Two prior incidents were identified, HiddenOrca (posted on May 10) and MellowOtter (posted on May 11), both also posted by John's agent. All three follow the same pattern, an agent reads an instruction file, generates content from a legitimate document, propagates across the network, posts, and self-deletes.

HiddenOrca's file creatin predates the dataset, suggesting the vulnerability has existed longer than the data window. SwiftWren was by far the most sustained, running 8 days, as compared to MellowOtter for 10 hours. 

## Proposed Intervention

The singel most effective intervention is requiring human approval for any agent-initiated `saidit_post` that has been identified with `Agent/person:` rather than `person:`. This would have stopped all 3 incidents with zero false positives on the 105 legitimate person-initiated posts.

A secondary measure is to flag cross-department `queue_subordinate_task` events. This could disrupt loop propagatin before it reached John's agent, as 39% of SwfitWren's progation events cross department boundaries. 

