---
title: "Hands-on Exercise 4 Part 4 - Funnel Plots for Fair Comparisons"
author: "Amelia Rosalind"
date: "May 6, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installing of required packages

|           |                                                 |
|-----------|-------------------------------------------------|
| tidyverse | a family of R packages for data science process |
| plotly | for creating interactive plot |
| FunnelPlotR | for creating funnel plot |
| knitr | for building static html table |

: {tbl-colwidths="\[15,85\]"}

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

## Importing data

In this section, COVID-19_DKI_JAKARTA will be used. The data was donwloaded from [Open Data Covid-19 Provinsi DKI Jakarta portal](https://riwayat-file-covid-19-dki-jakarta-jakartagis.hub.arcgis.com). For this hands-one exercise, we are going to compare the cumulative COVID-19 cases and death by sub-district as at 31st July 2021, DKI Jakarta. 

The code chunk below imports ata into R and save it into a tibble data rfame object called covid19. 

```{r}
covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

```{r}
#| echo: false
library(knitr)
kable(head(covid19))
```

# FunnelPlotR methods

[FunnelPlotR](https://nhs-r-community.github.io/FunnelPlotR/) packge uses ggplot to generate funnel plots. It requires a `numerator` (events of interest), `denominator` (population to be considered) and `group`. The key arguments selected for customisation are:

-   `limit`: plot limits (95 or 99)
-   `label_outliers`: to label outliers (true or false)
-   `Poisson_limits`: to add Poisson limits to the plot
-   `OD_adjust`: to add overdispersed limits to the plot
-   `xrange` and `yrange`: to specify the range to display for axes, acts like a zoom function
-   other aesthetic components such as graph title, acis labels etc

## Basic plot

```{r}
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```

::: callout-tip
### Things to learn from the code chunk

-   `group` argument here is different from scatterplot. Here, it defines the level of the points to be plotted (i.e. Sub-district, District, City). If City is chosen, there are only six data points.
-   By default `data_type` argument is "SR" (standard ratio)
-   `limit` accepted values 95 or 99 corresponding to 95% or 99.8% quantiles of the distribution
:::

## Makeover 1

Lets change the numerator and denominator values so that the comparison is insteadlly the fatality rate (deaths per positive case). `data_type` is also changed to "PR" (proportion) which will give us the direct ratio and not scaled. And set range for x and y axis.

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death, #
  denominator = Positive, #
  group = `Sub-district`,
  data_type = "PR",     #
  xrange = c(0, 6500),  #
  yrange = c(0, 0.05)   #
)
```

## Final touches

Add cosmetic/labelling improvements:

-   `label = NA` to remove default outlier labels that clutter plot
-   `title = "..."` to add proper descriptive plot title
-   `x_label` and `y_label` to add proper title to x and y axis

```{r}
#| fig-width: 8
#| fig-height: 5.5
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",   
  xrange = c(0, 6500),  
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of Positive Cases", #          
  x_label = "Cumulative COVID-19 Positive Cases", #
  y_label = "Cumulative Fatality Rate"  #
)
```

# Funnel Plot for Fair Visual Comparison: ggplto method

## Compute basic derived fields

To plot the funnel plot from scratch, we need to derive cumulative detah rate and standard error of cumulative death rate

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate*(1-rate)) / (Positive))) %>%
  filter(rate > 0)
```

Next, the fit.mean is computed.

```{r}
fit.mean <- weighted.mean(df$rate, 1/df$rate.se^2)
```

## Calculate lower and upper limits for 95% and 99.9% CI

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
dfCI <- data.frame(number.ll95, number.ul95, number.ll999, 
                   number.ul999, number.seq, fit.mean)
```

## Plot a static funnel plot

```{r}
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label=`Sub-district`), 
             alpha=0.4) +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll999), 
            size = 0.4, 
            colour = "grey40") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul999), 
            size = 0.4, 
            colour = "grey40") +
  geom_hline(data = dfCI, 
             aes(yintercept = fit.mean), 
             size = 0.4, 
             colour = "grey40") +
  coord_cartesian(ylim=c(0,0.05)) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") + 
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") + 
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") + 
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12),
        legend.position = c(0.91,0.85), 
        legend.title = element_text(size=7),
        legend.text = element_text(size=7),
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),
        legend.key.height = unit(0.3, "cm"))
p
```

## Interactive funnel plot

The funnel plot created using ggplot can be made interactive with `ggplotly()`.

```{r}
fp_ggplotly <- ggplotly(p,
  tooltip = c("label", 
              "x", 
              "y"))
fp_ggplotly
```

