---
title: "Hands-on Exercise 10: Information Dashboard Design: R Methods"
author: "Amelia Rosalind"
date: "June 26, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Getting Started

The following R packages are used in this exercise:

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
               reactablefmtr, gt, gtExtras, tidyverse)
```

-   [tidyverse](https://www.tidyverse.org/) is a collection of modern R packages for data import, tidying, wrangling, and visualisation, including `readr`, `tidyr`, `dplyr`, `ggplot2`, `tibble`, `stringr`, `forcats`, and `purrr`.
-   [lubridate](https://lubridate.tidyverse.org/) provides functions to work with dates and times more efficiently.
-   [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) extends ggplot2 with additional themes beyond its built-in options.
-   [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) offers helper functions for building richer tables with [gt](https://gt.rstudio.com/).
-   [reactable](https://glin.github.io/reactable/index.html) enables the creation of interactive data tables based on the React Table library.
-   [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) provides styling and formatting utilities to enhance `reactable` tables.

## Importing Microsoft Access Database

### The Data Set

A personal Microsoft Access database called *Coffee Chain* (`.mdb` format) is used throughout this exercise.

### Importing the Database into R

The `odbcConnectAccess2007()` function from the [RODBC](https://cran.r-project.org/web/packages/RODBC/index.html) package is used to connect to the database and pull the query table into R.

```{r}
#| eval: false
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

::: callout-note
Before running the code above, switch R to the 32-bit version, as `odbcConnectAccess()` is not compatible with 64-bit R.
:::

### Data Preparation

Load the saved `.rds` file:

```{r}
coffeechain <- read_rds("data/CoffeeChain.rds")
```

::: callout-note
This step can be skipped if `coffeechain` is already available in the R environment.
:::

Aggregate Sales and Budgeted Sales at the product level:

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

### Bullet Chart in ggplot2

The code below plots bullet charts for each product using ggplot2:

```{r}
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current),
           width=0.35,
           fill = "black") +
  geom_errorbar(aes(y = target,
                    x = Product,
                    ymin = target,
                    ymax= target),
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

## Plotting Sparklines Using ggplot2

### Preparing the Data

Filter and aggregate monthly sales per product:

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

Compute the minimum, maximum, and end-of-month sales values:

```{r}
mins <- group_by(sales_report, Product) %>%
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>%
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>%
  filter(Month == max(Month))
```

Compute the 25th and 75th quantiles:

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 0.25),
            quart2 = quantile(Sales, 0.75)) %>%
  right_join(sales_report)
```

### Sparklines in ggplot2

```{r}
ggplot(sales_report, aes(x=Month, y=Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2),
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank(), strip.text = element_blank())
```

## Static Information Dashboard Design: gt and gtExtras Methods

This section covers building a static information dashboard using the [gt](https://gt.rstudio.com/) and [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) packages. It is recommended to browse both package websites to get a feel for the range of functions available before proceeding.

### Plotting a Simple Bullet Chart

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current,
              target = target,
              width = 60,
              palette = c("lightblue",
                          "black")) %>%
  gt_theme_538()
```

## Sparklines: gtExtras Method

Before plotting the sales report with `gtExtras`, prepare the data:

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate(Month = month(Date,
                       label = TRUE,
                       abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

`gtExtras` functions require list columns. The code below converts the data accordingly:

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")
```

### Plotting the CoffeeChain Sales Report

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

### Adding Statistics

Calculate summary statistics for each product:

```{r}
report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### Combining the Data Frames

Create the sparkline list column:

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales),
            .groups = "drop")
```

Create the summary statistics table:

```{r}
sales <- report %>%
  group_by(Product) %>%
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

Join the two:

```{r}
sales_data = left_join(sales, spark)
```

### Plotting the Updated Data Table

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

### Combining Bullet Chart and Sparklines

Prepare the bullet chart data:

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup()
```

Join it with the existing sales data:

```{r}
sales_data = sales_data %>%
  left_join(bullet)
```

Plot the combined table with both sparklines and bullet chart:

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue",
                          "black")) %>%
  gt_theme_538()
```

## Interactive Information Dashboard Design: reactable and reactablefmtr Methods

This section covers building interactive dashboards using [reactable](https://glin.github.io/reactable/index.html) and [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html). As before, reviewing the documentation for both packages beforehand is strongly recommended.

Interactive sparklines require the `dataui` package, which must be installed from GitHub:

```{r}
remotes::install_github("timelyportfolio/dataui")
```

Then load it:

```{r}
library(dataui)
```

### Plotting Interactive Sparklines

Prepare the list column, then use `react_sparkline()` to render sparklines inside a `reactable` table:

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

```{r}
#| eval: false
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### Changing the Page Size

By default, `reactable` displays 10 rows per page. Use `defaultPageSize` to override this:

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### Adding Points and Labels

Use `highlight_points` to mark the minimum and maximum values, and `labels` to annotate the first and last data points:

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

### Adding a Reference Line

The `statline` argument overlays a reference line — in this case, the mean:

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

### Adding a Bandline

Instead of a reference line, a bandline can be used to shade a range. Here, the inner quartile range is highlighted:

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

### Changing from Sparkline to Sparkbar

To display values as bars instead of a line, switch from `react_sparkline()` to `react_sparkbar()`:

```{r}
#| eval: false
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```
