---
title: "Hands-on Exercise 3 - Interactive Data Visualisation"
author: "Amelia Rosalind"
date: "April 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installation of required libraries

R packages needed:

-   ggiraph: making ggplot graphic interactive
-   plotly: plotting interactive statistical graphs
-   DT: provide a R interface to JavaScript library DataTables that create interactive table on html page
-   tidyverse: support data science, analysis and communication task including creating static graphs
-   patchwork: combine multiple ggplot graphs into one figure

```{r}
pacman::p_load(ggiraph, plotly, patchwork,
               DT, tidyverse)
```

## Importing data

Exam_data.csv is used for this exercise. `read_csv()` is used to read the csv file and import Exam_data into R.

```{r}
exam_data <- read_csv("Exam_data.csv")
```

# Interactive Data Visualisation - ggiraph methods

[ggiraph](https://davidgohel.github.io/ggiraph/index.html) is an htmlwidget and a ggplot2 extension. It allows ggplot graphics to be interactive. Interactivity is made with ggplot geometries that can understand three arguments:

-   Tooltip: a column of data-sets that contain tooltips to be displayed when the mouse is over elements
-   OnClick: a column of data-sets that contain a JavaScript function to be executed when elements are clicked
-   Data_id: a column of data-sets that contain an id to be associated with elements

If it used within a shiny application, elements associated with an id (data_id) can be selected and manipulated on client and server sides.

# Tooltip Effect With tooltip

The code chunk below is a typical code chunk to plot an interactive statistical graph using `tooltip` argument from ggiraph package. By hovering the mouse pointer on a data point of interest, the student's ID will be displayed. You can try it on the graph below.

```{r}
p <- ggplot(data=exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

::: callout-note
Notice that the code chunk consists of two parts. First, an interactive version of ggplot object will be created. In this case, `geom_dotplot_interactive()` is used to create the basic graph. Next, [girafe()](https://davidgohel.github.io/ggiraph/reference/girafe.html) of ggiraph will be used to create an interactive svg object to be displayed on an html page.
:::

## Displaying multiple information on tooltip

The content of the tooltip can be customized by including a list object as shown in the code chunk below.

```{r}
exam_data$tooltip <- c(paste0(     
  "Name = ", exam_data$ID,         
  "\n Class = ", exam_data$CLASS)) 

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip), 
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8*0.618
)
```

The first 3 lines in the code is create a column called tooltip in Exam_data which is filled with the combined text of ID and CLASS values. Now this will act as the label for each student. Then this field is then called in line 7 `tooltip = exam_data(tooltip)` so that when building the chart it will use the new tooltip column instead.

## Customizing tooltip sytle

opts_tooltip() of ggiraph is used to customize tooltip rendering by add css declaration. Refer to the code chunk below.

```{r}
tooltip_css <- "background-color:white; #<<
font-style:bold; color:black;" #<<

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = ID),                   
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(    #<<
    opts_tooltip(    #<<
      css = tooltip_css)) #<<
)                                        
```

Notice that the background color of the tooltip is black adn font color is white and bold.

Refer to [this article](https://www.ardata.fr/ggiraph-book/customize.html) to learn more about customizing girafe objects.

## Displaying statistics on tooltip

Code chunk below shows an advanced way to customise tooltip. Int his example, a function is used to cimpute 90% confident interval of the mean. The derived statistics are then displayed in the tooltip.

```{r}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths score:", mean, "+/-", sem)
}

gg_point <- ggplot(data=exam_data, aes(x = RACE)) +
  stat_summary(aes(y = MATHS, 
                   tooltip=after_stat(
                     tooltip(y, ymax))),
               fun.data = "mean_se",
               geom = GeomInteractiveCol,
               fill = "light blue") +
  stat_summary(aes(y = MATHS), 
               fun.data = mean_se,
               geom = "errorbar", width = 0.2, size = 0.2)

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

# Hover Effect with data_id

Code chunk below make use of another interactive feature of ggiraph, namely `data_id`.

```{r}
p <- ggplot(data=exam_data,
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

This creates a hover effect that highlights elements associated with the same data_id, which in this class is CLASS.

## Styling hover effect

The code chunk below will change the css code for the highlighting effect.

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

::: callout-note
The default css style is an orange fill. In this example, the css in `opts_hover()` would mean the data points that are hovered over will have a close to black fill. But the css style of other data points will have their opacity reduced to 0.2 as set in `opts_hover_inv()`.
:::

# Combine Tooltip and Hover Effect

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = CLASS,
        data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

In this case, there is interactivity when a data point is hover over and elements that are associated with the same data_id, in this case CLASS, will be highlighted too. At the same time, the tooltip will also show the CLASS value.

# Click Efect With onclick

`onclick` argument of ggiraph provides hotlink interactivity on the web. The code chunk below is an example of onclick. A web document link with a data object will be displayed on the web browser upon a mouse click on any of the data points. Note that the hotlink is for illustration purpose only.

```{r}
exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder/primary%20school",
as.character(exam_data$ID))

p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(onclick = onclick),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618)
```

::: callout-warning
There are 2 requirements for onlick column to work.

1.  It must be a string column so it must contain text or character values not numbers or decimals. Hence the use of `as.character(exam_data$ID)` to make sure the ID is converted into string before pasted into URL.
2.  Text needs to be a valid JavaScript as the text will be executed as Javascript in the browser. Hence the use of `window.open(...)` which is standard JavaScript command that opens a URL.
:::

# Coordinated Multiple Views with ggiraph

It is possible to create coordinated multiple views. To do this, 2 things are needed:

1.  Appropriate interactive functions of ggiraph will be used to create the multiple views
2.  patchword function of patchwork package will be used inside girafe function to create the interactive coordinated multiple views

```{r}
p1 <- ggplot(data=exam_data,
             aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim=c(0,100)) +
  scale_y_continuous(NULL,
                     breaks = NULL)

p2 <- ggplot(data=exam_data,
             aes(x = ENGLISH)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim=c(0,100)) +
  scale_y_continuous(NULL,
                     breaks = NULL)

girafe(code = print(p1 + p2),
       width_svg = 6,
       height_svg = 6*0.618,
       options = list(
         opts_hover(css = "fill: #202020"),
         opts_hover_inv(css = "opacity:0.2;")
       )
      )
```

::: callout-note
Notice that when a data point of one of the dotplot is selected, the corresponding data point ID on the second data visualisation will be highted too. As such data_id aesthetic is critical to link observations between plots. Adding tooltip aesthetic is optional but nice to have when mouse over a point.
:::

# Interactive Data Visualisation - plotly methods

Plotly's R graphing library create interactive web graphics from ggplot graphs and/or custom interface to JacaScript library plotly.js inspired by the grammar of graphics.

There are two ways to create interactive graph by using plotly: - using `plot_ly()` - using `ggplotly()`

## `plot_ly()` method

::: panel-tabset
### Basic

```{r}
plot_ly(data = exam_data, 
             x = ~MATHS, 
             y = ~ENGLISH)
```

### With visual argument

```{r}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```

If hovered over the data points, it will also say the Race.
:::

# `ggplotly()` method

```{r}
p <- ggplot(data=exam_data, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```

Notice that the only extra line in the code chunk is `ggplotly()`

## Coordinate multiple views with plotly

The creation of coordinated linked plot by using plotly incolves three steps:

-   highlight_key() of plotly package is used as shared data
-   two scatterplots will be created by using ggplot functions
-   subplot() of plotly package is used to place them next to each other side-by-side

```{r}
d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

subplot(ggplotly(p1),
        ggplotly(p2))
```

In terms of interactivity, by clicking a data points of one of the scatterplot will also select the corresponding point on the other scatterplot.

# Interactive Data Visualisation - crosstalk method

[Crosstalk](https://rstudio.github.io/crosstalk/index.html) is an add-on to the htmlwidgets package. It extends htmlwidgets with a set of classes, functions and conventions for implementing cross-widget interactions (currentlu, linked brushing and filtering).

## Interactive data table

Data objects in R can be rendered as HTML tables using the JavaScript library [DataTables](https://datatables.net).

```{r}
DT::datatable(exam_data, class= "compact")
```

## Linked brushing

It is possible to create interactive coordinated visalisations where selecting data points in one widget automatically highlights or filters the corresponding data in another widget. Brush here mean click or drag to select data points in one chart. So a linked brushing means that by dragging to select data points in one chart, the same data points are simultaneously highlighted in another chart or table.

:::: panel-tabset

### Code
```{r}
#| eval: false
d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        
```

::: callout-note
-   `highlight()` is a function of plotly package. It sets a variety of optins for brushing multiple plots. These options are primarily designed for linking multuple ploly graphs, and may not behave as expected when linking plotly to another htmlwidget package via crosstalk. In some cases, other htmlwidgets will respect these options, such as persistent selection in leaflet.
-   `bscols()` is a helper function of crosstalk package It makes it easy to put HTML elements side by side. It can be called directly from the console but is especially designed to work in a R Markdown document. **Warning**: This will bring in all of Bootstrap!
:::

### Output plot
```{r}
#| echo: false
d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        
```
::::
