---
title: "Hands-on Exercise 5 Part 1 - Creating Ternary Plot with R"
author: "Amelia Rosalind"
date: "May 13, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installing of required packages

|  |  |
|------------------------|------------------------------------------------|
| ggtern | ggplot extension specially designed to plot ternanry diagrams |
| Plotly R | R package for creating interactive web-based graphs via plotly's JavaScript graphing library |
| tidyverse | a family of modern R packages specially designed to support data science, analysis and communication task including creating static statistical graphs |

: {tbl-colwidths="\[15,85\]"}

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## Importing data

For this exercise, [Singapore Residents by Planning AreaSubzone, Age Group, Sex adn Type of Dwelling](https://www.singstat.gov.sg/find-data/explore-data-themes/population/geographic-distribution/latest-news-data) from June 2000-2018 data will be used. 

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv")
```

## Preparing data
Use `mutate()` function of dplyr package to derive three new measures, namely: young, active and old.

```{r}
agpop_mutated <- pop_data %>%
  dplyr::mutate(Year = as.character(Year))%>%
  spread(AG, Population)%>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))%>%
  mutate(OLD = rowSums(.[17:21]))%>%
  mutate(TOTAL = rowSums(.[22:24]))%>%
  filter(Year == "2018")%>%
  filter(TOTAL > 0)
```

# Plotting Ternary Diagram with R

## Static ternary diagram

Use `ggern()` function fo fftern package to create a simple ternary plot.

```{r}
ggtern(data=agpop_mutated, aes(x=YOUNG, y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

## Interactive ternary diagram

```{r}
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```
