---
title: "Take-Home Exercise 1 - Technical Summary"
author: "Amelia Rosalind"
date: "May 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
  cache: true
---

# Overview

The data used is a real-world vehicle insurance data from [Spanish insurance company](https://data.mendeley.com/datasets/sw4jmdb2sm/1), comprising 354,140 policy records spanning from 2022 to 2024. The objective is to gain deeper insights into data distributions using univariate analysis and understand key factors influencing profitability and volatility using bivariate analysis. This technical report will cover the technical working of this project, such as data preprocessing and all analyses done.

# Getting Started

## Installing of required packages

```{r}
pacman::p_load(ggstatsplot, tidyverse, knitr,
               patchwork, ggthemes, ggdist,
               car)
```

## Importing data

```{r}
insurance <- read_delim("Motor_insurance.csv",
                        delim = ";",
                        col_types = cols(.default = "c"))
```

# Checking Dataset

Before cleaning, first examine the dataset's structure, summary statistics, and missing values to identify any data quality issues.

```{r}
glimpse(insurance)
summary(insurance)
colSums(is.na(insurance))[colSums(is.na(insurance)) > 0]
kable(head(insurance))
```

## Problems

1.  All features are now characters so need to change the data type of the numerical values into numeric.

2.  Some features have nulls: vehicle_value has 513 NAs, fuel_type has 1287 NAs, and both vehicle_age and age_driving_licence has 2 NAs each.

3.  Categoricals (policy_type, policy_status, business_type, payment_frequency, bonus_score, fuel_type, municipality_type, circulation_area, vehicle_brand, year) are stored as characters <chr> instead of factors <fct>. With data type factors, R will understand that it is categoricals and not some string.

# Data Wrangling

## Cleaning the data

```{r}
insurance <- insurance %>%
  # Fix numerical columns' data type
  mutate(across(c(insured_id, driver_age, vehicle_age,
                  age_driving_licence, vehicle_value, 
                  seats, power_to_weight_ratio,
                  total_premium, total_claims,
                  liability_claims, 
                  liability_property_claims,
                  liability_injury_claims, property_claims,
                  theft_claims, fire_claims, glass_claims,
                  legal_protection_claims, occupants_claims,
                  total_incurred, liability_incurred,
                  liability_property_incurred,
                  liability_injury_incurred,
                  property_incurred, theft_incurred,
                  fire_incurred, glass_incurred,
                  legal_protection_incurred,
                  occupants_incurred, liability_premium,
                  property_damage_premium, theft_premium,
                  fire_premium, glass_premium,
                  legal_protection_premium,
                  occupants_premium, total_exposure,
                  liability_exposure),
                ~ as.numeric(.))) %>%
  # Convert categoricals to factors
  mutate(across(c(policy_type, policy_status, business_type,
                  payment_frequency,bonus_score, fuel_type,
                  municipality_type, circulation_area,
                  vehicle_brand, year),
                as.factor)) %>%
  drop_na()

glimpse(insurance)
colSums(is.na(insurance))[colSums(is.na(insurance)) > 0]
```

## Feature Engineering

### driving_experience

driving_experience is needed to get how long the driver has been driving for, which is more meaningful as compared to the age when the driver receives the driving licence.

First check if age_driving_licence means the year driving licence is issued or the age the driver got the driving licence.

```{r}
insurance %>%
  filter(age_driving_licence == 80) %>%
  select(driver_age, age_driving_licence, year)
```

Combining with the first row value shown in the head summary, it can be concluded that age_driving_licence is the age of the driver when driving licence is issued. Hence to get driving_experience the formula should be driver_age - age_driving_licence.

```{r}
insurance <- insurance %>%
  mutate(driving_experience = 
           driver_age - age_driving_licence)
```

### loss_ratio

This feature will show the ratio of total_incurred to total_premium. This is important as it is the standard insurance profitability metric.

```{r}
insurance <- insurance %>%
  mutate(loss_ratio = total_incurred / total_premium)
```

There is a possibility that the total_premium is 0 which will produce Inf or NaN instead of a value, so check next.

```{r}
sum(is.infinite(insurance$loss_ratio))
sum(is.nan(insurance$loss_ratio))
```

Since there is only a total of 49 values Inf or NaN, dropping rows with them is negligible.

```{r}
insurance <- insurance %>%
  mutate(loss_ratio = ifelse(is.infinite(loss_ratio) | is.nan(loss_ratio), 
                             NA, loss_ratio)) %>%
  drop_na(loss_ratio)

sum(is.infinite(insurance$loss_ratio))
sum(is.nan(insurance$loss_ratio))
nrow(insurance)
```

## Check final data

```{r}
kable(head(insurance))
```

The final dataset now have 352,289 rows remaining, and 49 columns in total with the new column made. The data types are also all correct.

# Variable selection

Profitability is about how much profit the insurance company can make from a policy, which is addressed with `loss_ratio`. Volatility concerns the spread and unpredictability of losses, which can be addressed by `total_incurred`. As such the target variables are `loss_ratio` and `total_incurred`.

Other variables that can help explain profitability and volatility are listed on the table below

| Variable | Description | Purpose |
|-----------------------|----------------------------|---------------------|
| `total_premium` | Revenue collected per policy | Provides context on income side of portfolio |
| `total_claims` | Number of claims made | Indicates claim frequency which directly drives costs |
| `driver_age` | Age of driver | Older or younger drivers may present different risk profiles |
| `vehicle_age` | Age of the insured vehicle | Older vehicles may be more probe to certain types of claims |
| `driving_experience` | Years of driving experience | More experienced drivers may have lower claim rates |
| `policy_type` | Types coverage held | Different coverage types carry different risk exposures and premium levels |
| `bonus_score` | Claims history indicator | Drivers with poor claims history are likely higher risk |
| `circulation_area` | Urban vs rural driving environment | Urban areas typically have higher accident rates |

: {tbl-colwidths="\[20,20, 60\]"}

Year will also be taken as the temporal variable to be observed.

# Univariate Analysis

For univariate analysis, there are 2 types of models to be done based on if the values are numeric or categorical.

The numeric ones will use histogram and boxplot together. Histogram will be used to show the shape of the distribution. Boxplot will complement it by highlighting median, IQR and outliers. Both are combined using patchwork for a more complete view.

The categorical values will use bar chart. This type of graph is able to show frequency distribution of each level.

The table below will show the which features are numerical and categorical and the type of univariate analysis done.

| Variable             | Type of value | Univariate analysis |
|----------------------|---------------|---------------------|
| `total_incurred`     | Numerical     | histogram + boxplot |
| `loss_ratio`         | Numerical     | histogram + boxplot |
| `total_premium`      | Numerical     | histogram + boxplot |
| `total_claims`       | Numerical     | histogram + boxplot |
| `driver_age`         | Numerical     | histogram + boxplot |
| `vehicle_age`        | Numerical     | histogram + boxplot |
| `driving_experience` | Numerical     | histogram + boxplot |
| `policy_type`        | Categorical   | bar chart           |
| `bonus_score`        | Categorical   | bar chart           |
| `circulation_area`   | Categorical   | bar chart           |
| `year`               | Categorical   | bar chart           |

## Distribution of target variables

:::: panel-tabset
### Volatility

```{r}
summary(insurance$total_incurred)
```

```{r}
box_cost <- ggplot(insurance, aes(y = total_incurred, x = "")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$total_incurred)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-1e+05, 6e+05)) +
  labs(title = "Distribution of Total Incurred (Cost)") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hist_cost <- ggplot(insurance, aes(x = total_incurred)) +
  geom_histogram(bins = 50,
                 colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(total_incurred)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(total_incurred)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-1e+05, 6e+05)) +
  labs(x = "total_incurred") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_cost / hist_cost) +
  plot_layout(heights = c(1.5, 3.5))

hbox_final <- hbox & theme(
  plot.background =element_rect(
    fill="white",colour="white"
    )
  )

hbox_final
```

```{r}
#| echo: false
if (!dir.exists("plots")) dir.create("plots")
ggsave("plots/hist_box_total_incurred.png", 
       plot = hbox_final,
       width = 8, 
       height = 5,
       dpi = 150)
```

The distribution is heavily right-skewed with severe zero-inflation, as the median is 0 and the mean is 208.4. This indicates the majority of policies incurred no claims, with a small proportion of policies driving disproportinately high losses - a pattern typical in insurance portofolios.

### Profitability

```{r}
summary(insurance$loss_ratio)
```

```{r}
insurance %>%
  filter(loss_ratio > 100) %>%
  select(total_premium, total_incurred, loss_ratio) %>%
  arrange(desc(loss_ratio))
```

::: callout-note
There are 243 policies with extreme loss ratios exceeding 100, indicating cases where claim cost far outweigh the premium collected. As these are outliers, the distribution is plotted within the range of 0 and 100 to better visualise the majority of policies.
:::

```{r}
box_ratio <- ggplot(insurance, aes(y = loss_ratio, x = "")) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(x = "", 
                 y = mean(insurance$loss_ratio)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-1, 10)) +
  labs(title = "Distribution of Loss Ratio") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_ratio <- ggplot(insurance, aes(x = loss_ratio)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(loss_ratio)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(loss_ratio)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-1, 10)) +
  labs(x = "loss_ratio") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_ratio / hist_ratio) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

```{r}
#| echo: false
ggsave("plots/hist_box_loss_ratio.png", 
       plot = last_plot(),
       width = 8, 
       height = 5,
       dpi = 150)
```

Similar to `total_incurred`, `loss_ratio` extreme right skew with a median of 0 and 243 policies exceeding a ratio of 100. The wide gap between mean (0.788) and median (0) suggest the average is heavily pulled by catastrophic outlier policies.
::::

## Distribution of other risk factors

::: panel-tabset
### total_premium

```{r}
summary(insurance$total_premium)
```

```{r}
box_rev <- ggplot(insurance, aes(y = total_premium, x = "")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$total_premium)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-500, 6000)) +
  labs(title = "Distribution of Total Premium (Revenue)") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hist_rev <- ggplot(insurance, aes(x = total_premium)) +
  geom_histogram(bins = 50,
                 colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(total_premium)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(total_premium)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-500, 6000)) +
  labs(x = "total_premium") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_rev / hist_rev) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

```{r}
#| echo: false
ggsave("plots/hist_box_total_premium.png", 
       plot = last_plot(),
       width = 8, 
       height = 5,
       dpi = 150)
```

Premiums are right-skewed with a median of 257.75 and mean of 296.29, indicating most policies are moderately prices with a tail of hig-value policies. The distribution suggests a fairly broad but concentrated mid-range portfolio.

### total_claims

```{r}
summary(insurance$total_claims)
```

```{r}
box_claims <- ggplot(insurance, aes(y = total_claims, x="")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$total_claims)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-5, 25)) +
  labs(title = "Distribution of Total Claims (Frequency)") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_claims <- ggplot(insurance, aes(x = total_claims)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(total_claims)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(total_claims)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-5, 25)) +
  labs(x = "total_claims") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_claims / hist_claims) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

```{r}
#| echo: false
ggsave("plots/hist_box_total_claims.png", 
       plot = last_plot(),
       width = 8, 
       height = 5,
       dpi = 150)
```

Like `total_incurred`, claim frequency is zero-inflated with a median of 0 and mean of 0.22, confirming that most policyholders make no claims within a given year.

### driver_age

```{r}
summary(insurance$driver_age)
```

```{r}
box_da <- ggplot(insurance, aes(y = driver_age, x="")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$driver_age)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(15, 95)) +
  labs(title = "Distribution of Driver's Age") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_da <- ggplot(insurance, aes(x = driver_age)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(driver_age)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(driver_age)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(15, 95)) +
  labs(x = "driver_age") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_da / hist_da) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

```{r}
#| echo: false
ggsave("plots/hist_box_driver_age.png", 
       plot = last_plot(),
       width = 8, 
       height = 5,
       dpi = 150)
```

Driver age is approximately normally distributed with a median of 48 and mean of 49, ranging from 18 to 90. The portfolio skews towards middle-aged and older driver, which may reflect lower-risk segments self-selecting into this insurer.

### vehicle_age

```{r}
summary(insurance$vehicle_age)
```

```{r}
box_va <- ggplot(insurance, aes(y = vehicle_age, x="")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$vehicle_age)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-5, 75)) +
  labs(title = "Distribution of Vehicle Age") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_va <- ggplot(insurance, aes(x = vehicle_age)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(vehicle_age)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(vehicle_age)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-5, 75)) +
  labs(x = "vehicle_age") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_va / hist_va) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

```{r}
#| echo: false
ggsave("plots/hist_box_vehicle_age.png", 
       plot = last_plot(),
       width = 8, 
       height = 5,
       dpi = 150)
```

Vehicle age is broadly distributed with a median of 25 years and mean of 25.9, suggesting the portfolio is dominated by older vehicles. This i notable as older vehicles may have different risk profiles for certain claim types.

### driving_experience

```{r}
summary(insurance$driving_experience)
```

```{r}
box_de <- ggplot(insurance, aes(y = driving_experience, x="")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$driving_experience)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-5, 75)) +
  labs(title = "Distribution of Driving Experience") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_de <- ggplot(insurance, aes(x = driving_experience)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(driving_experience)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(driving_experience)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-5, 75)) +
  labs(x = "driving_experience") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_de / hist_de) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

```{r}
#| echo: false
ggsave("plots/hist_box_driving_experience.png", 
       plot = last_plot(),
       width = 8, 
       height = 5,
       dpi = 150)
```

Driving experience closely mirrors vehicle age in distribution (median 25, mean 26), which is expected given it si derived from driver age. Most insured drivers are experienced, with the distribution peaking around 20-30 years of experience.

### policy_type

```{r}
p_policy_type <- ggplot(
  data=insurance, 
  aes(x = policy_type)
  ) +
  geom_bar(fill = "grey90") +
  geom_text(stat="count",
            aes(label=paste0(after_stat(count))),
            vjust = -0.2,
            size = 3) +
  labs(title = "Distribution of Policy Types") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

p_policy_type
```

```{r}
#| echo: false
ggsave("plots/bar_policy_type.png", 
       plot = p_policy_type,
       width = 8, 
       height = 5,
       dpi = 150)
```

The portfolio is concentrated in a small number of policy types, with type CC being the most dominant. This shows limited product diversification.

### bonus_score

```{r}
p_bonus_score <- ggplot(
  data=insurance, 
  aes(x = bonus_score)
  ) +
  geom_bar(fill = "grey90") +
  geom_text(stat="count",
            aes(label=paste0(after_stat(count))),
            vjust = -0.2,
            size = 3) +
  labs(title = "Distribution of Good, Bad, Neutral Claims") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

p_bonus_score
```

```{r}
#| echo: false
ggsave("plots/bar_bonus_score.png", 
       plot = p_bonus_score,
       width = 8, 
       height = 5,
       dpi = 150)
```

The majority of policyholders hold a "G" (good) bonus score, indicating a portfolio with predominantly low-risk claims history. Bad and neutral scores are a small minority.

### circulation_area

```{r}
p_circulation_area <- ggplot(
  data=insurance, 
  aes(x = circulation_area)
  ) +
  geom_bar(fill = "grey90") +
  geom_text(stat="count",
            aes(label=paste0(after_stat(count))),
            vjust = -0.2,
            size = 3) +
  labs(title = "Distribution of Urban vs Rural") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

p_circulation_area
```

```{r}
#| echo: false
ggsave("plots/bar_circulation_area.png", 
       plot = p_circulation_area,
       width = 8, 
       height = 5,
       dpi = 150)
```

Urban and rural circulation environment are quite distributed, with urban being more dominant. This implies that the expected claim frequency for vehicles predominantely in the urban areas typically carry higher accident exposure.

### year

```{r}
p_year <- ggplot(
  data=insurance, 
  aes(x = year)
  ) +
  geom_bar(fill = "grey90") +
  geom_text(stat="count",
            aes(label=paste0(after_stat(count))),
            vjust = -0.2,
            size = 3) +
  labs(title = "Distribution of Policy Record by Year") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

p_year
```

```{r}
#| echo: false
ggsave("plots/bar_year.png", 
       plot = p_year,
       width = 8, 
       height = 5,
       dpi = 150)

p_combined_categorical <- (p_policy_type | p_bonus_score) / 
                          (p_circulation_area | p_year)

ggsave("plots/categorical_2x2.png",
       plot = p_combined_categorical,
       width = 10,
       height = 8,
       dpi = 150)
```

The dataset contains records across three years (2022–2024), with broadly similar policy counts per year. This suggests a stable portfolio size over the observation period, making year-on-year comparisons meaningful.
:::

# Bivariate Analysis

`ggbetweenstats` and `ggscatterstats` are used to do statistical bivariate analysis.

## Relationship between bonus score and profitability

One-way ANOVA test is suitable to find the relationship between bonus score and profitability since `bonus_score` has more than 2 groups. But first check first if parametric or non-parametric using Shapiro test on `loss_ratio`.

```{r}
# for reproducibility
set.seed(123)
shapiro.test(sample(insurance$loss_ratio, 5000))
```

Since result shows that it is significant, then Kruskal-Wallis test is used.

```{r}
p <- ggbetweenstats(
  data = insurance,
  x = bonus_score,
  y = loss_ratio,
  type = "np",
  title = "Relationship between bonus score and profitability"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_bonus_profitability.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

The Kruskal-Wallis test reveals a statically significant difference in loss ratios across bonus score groups. Policyholders with poor claims history (bad bonus score) tend to exhibit higher loss ratios, consistent with expectation that past claims behavior is a strong predictor of future profitability.

## Relationship between policy type and profitability

One-way ANOVA is also used to find the relationship between policy type and profitability since `policy_type` has 5 groups. Since the Shapiro test of `loss_ratio` is significant, then the test done should be non-parametric Kruskal-Wallis test.

```{r}
p <- ggbetweenstats(
  data = insurance,
  x = policy_type,
  y = loss_ratio,
  type = "np",
  title = "Relationship between policy type and profitability"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_policytype_profitability.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

A statistically significant difference in loss ratios exists across policy types. Certain coverage types carry notably higher median loss ratios, suggesting that product mix is a meaningful driver of portfolio profitability.

## Relationship between circulation area and profitability

Since circulation area only has 2 groups, the test done should be a two-sample test. In addition, since the Shapiro test for `loss_ratio` is significant, use the non-parametric version which is the Mann-Whitney U test.

```{r}
p <- ggbetweenstats(
  data = insurance,
  x = circulation_area,
  y = loss_ratio,
  type = "np",
  title = "Relationship between circulation area and profitability"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_circulation_profitability.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

The Mann-Whitney U test indicates a statistically significant difference in loss ratios between urban and rural policyholders. Urban policies tend toward higher loss ratios, consistent with the higher accident exposure in density populated areas.

## Relationship between years of driving experience and profitability

Since driving experience is a numeric value, the test done should be a Significant Test of Correlation. This is done using `ggscatterstats`. Looking at the univariate analysis, even though `driving_experience` is normally distributed, `loss_ratio` is very skewed. So the non-parametric version is used instead, which is the Spearman's rank correlation coefficient.

```{r}
p <- ggscatterstats(
  data = insurance,
  x = driving_experience,
  y = loss_ratio,
  type = "nonparametric",
  title = "Relationship between years of driving experience and profitability"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_drivingexp_profitability.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

Spearman's rank correlation reveals a weak but statistically significant negative relationship between driving experience and loss ratio. More experienecd drivers tend to have marginally lower loss ratios, though the effect size suggests experience alone is not a strong stabalone predictor.

## Relationship between driver age and profitability

Driver age is also numeric value that is normally distributed based on the univariate analysis. However, `loss_ratio` is very skewed, so the test to be done should be Spearman's rank correlation coefficient.

```{r}
p <- ggscatterstats(
  data = insurance,
  x = driver_age,
  y = loss_ratio,
  type = "nonparametric",
  title = "Relationship between driver's age and profitability"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_driverage_profitability.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

Spearman's rank correlation reveals a statistically significant but weak relationship between driver age and loss ratio. The direction and effect size suggest age alone has limited standalone predictive power for profitability, though it may interact with other factors such as driving experience.

## Relationship between vehicle age and profitability

Vehicle age is also numeric value that is normally distributed based on the univariate analysis. However, `loss_ratio` is very skewed, so the test to be done should be Spearman's rank correlation coefficient.

```{r}
p <- ggscatterstats(
  data = insurance,
  x = vehicle_age,
  y = loss_ratio,
  type = "nonparametric",
  title = "Relationship between vehicle age and profitability"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_vehicleage_profitability.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

Spearman's rank correlation shows a statistically significant relationship between vehicle age and loss ratio. Older vehicles tend to be associated with differing loss patterns, potentially reflecting differences in repair costs or the types of claims made on ageing vehicles.

## Relationship between bonus score and volatility

```{r}
# for reproducibility
set.seed(123)
shapiro.test(sample(insurance$total_incurred, 5000))
```

Like against `loss_ratio`, One-Way ANOVA is used. Since, the Shapiro test shows `total_incurred` is significant which means `total_incurred` is not normally distributed. Therefore, non-parametric test, Kruskal-Wallis, is used.

```{r}
p <- ggbetweenstats(
  data = insurance,
  x = bonus_score,
  y = total_incurred,
  type = "np",
  title = "Relationship between bonus score and volatility"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_bonus_volatility.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

Similar to the profitability analysis, the Kruskal-Wallis test shows a significant difference in incurred costs across bonus score groups. Policyholders with poor claims history are associated with higher and more volatile incurred losses.

## Relationship between policy type and volatility

Policy type as mentioned before has 5 groups and it is confirmed that `total_incurred` is not normally distributed from Shapiro test, therefore Kruskal-Wallis is used.

```{r}
p <- ggbetweenstats(
  data = insurance,
  x = policy_type,
  y = total_incurred,
  type = "np",
  title = "Relationship between policy type and volatility"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_policytype_volatility.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

Significant difference in total incurred costs exist across policy types, mirroring the profitability findings. Certain policy types no only affect lsos ratios but also drive higher absolute claim costs, contributing to portfolio volatility.

## Relationship between circulation area and volatility

Since circulation area only has 2 groups, the test done should be a two-sample test. In addition, since the `total_incurred` is not normally distributed, use the non-parametric version which is the Mann-Whitney U test.

```{r}
p <- ggbetweenstats(
  data = insurance,
  x = circulation_area,
  y = total_incurred,
  type = "np",
  title = "Relationship between circulation area and volatility"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_circulation_volatility.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

The Mann-Whitney U test indicates a statistically significant difference in total incurred costs between urban and rural policyholders. Urban policies are associated with higher incurred losses, consistent with greater accident exposure in densely populated areas, reinforcing the profitability finding.

## Relationship between years of driving experience and volatility

Since driving experience is a numeric value and is normally distributed, but `total_incurred` is very skewed, Spearman's rank correlation coefficient is done.

```{r}
p <- ggscatterstats(
  data = insurance,
  x = driving_experience,
  y = total_incurred,
  type = "nonparametric",
  title = "Relationship between years of driving experience and volatility"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_drivingexp_volatility.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

The Spearman correlation between driving experience and total incurred costs mirrors the profitability finding — a weak but statistically significant negative relationship. More experienced drivers tend to incur lower claim costs, though the effect size suggests it is not a dominant driver of volatility on its own.

## Relationship between driver age and volatility

Driver age is also numeric value that is normally distributed based on the univariate analysis. However, `total_incurred` is very skewed, so Spearman's rank correlation coefficient is done.

```{r}
p <- ggscatterstats(
  data = insurance,
  x = driver_age,
  y = total_incurred,
  type = "nonparametric",
  title = "Relationship between driver's age and volatility"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_driverage_volatility.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

Consistent with the profitability analysis, driver age shows a statistically significant but weak correlation with total incurred costs. The result suggests age has a modest influence on claim severity, likely working in conjunction with other risk factors.

## Relationship between vehicle age and volatility

Vehicle age is also numeric value that is normally distributed. However, `total_incurred` is very skewed, so the test to be done should be Spearman's rank correlation coefficient.

```{r}
p <- ggscatterstats(
  data = insurance,
  x = vehicle_age,
  y = total_incurred,
  type = "nonparametric",
  title = "Relationship between vehicle age and volatility"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_vehicleage_volatility.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

Vehicle age shows a statistically significant relationship with total incurred costs. Older vehicles appear associated with higher incurred losses, suggesting vehicle age contributes to claims volatility, possibly due to higher repair costs or increased mechanical vulnerability.

# Temporal Analysis: Profitability and Volatility Across Years

To examine temporal trends, year is treated as a categorical grouping variable given only three time points are available (2022–2024). A line chart of mean and median values is first used to visualise the direction of change, followed by Kruskal-Wallis tests via ggbetweenstats to statistically validate whether differences across years are significant.

::: panel-tabset
## Profitability across years

```{r}
insurance %>%
  group_by(year) %>%
  summarise(mean_lr = mean(loss_ratio),
            median_lr = median(loss_ratio)) %>%
  pivot_longer(cols = c(mean_lr, median_lr),
               names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(x = year, y = value,
             colour = metric, group = metric)) +
  geom_line() +
  geom_point() +
  labs(title = "Mean and Median Loss Ratio by Year") +
  theme_minimal()
```

```{r}
#| echo: false
ggsave("plots/line_lossratio_year.png", 
       plot = last_plot(),
       width = 8, 
       height = 5,
       dpi = 150)
```

```{r}
p <- ggbetweenstats(
  data = insurance,
  x = year,
  y = loss_ratio,
  type = "np",
  title = "Profitability across Years (2022–2024)"
)

p
```

```{r}
#| echo: false
ggsave("plots/biv_year_profitability.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

The mean loss ratio shows a consistent upward trend from 2022 to 2024 while the median remains at 0 across all years, indicating that rising averages are driven by outlier policies rather than broad portfolio deterioration. The Kruskal-Wallis test confirms this year-on-year difference is statistically significant (χ² = 196.05, p = 2.69e-43), with pairwise comparisons showing all three years differ significantly from each other.

## Volatility across years

```{r}
insurance %>%
  group_by(year) %>%
  summarise(mean_ti = mean(total_incurred),
            median_ti = median(total_incurred)) %>%
  pivot_longer(cols = c(mean_ti, median_ti),
               names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(x = year, y = value,
             colour = metric, group = metric)) +
  geom_line() +
  geom_point() +
  labs(title = "Mean and Median Total Incurred by Year") +
  theme_minimal()
```

```{r}
#| echo: false
ggsave("plots/line_totalincurred_year.png", 
       plot = last_plot(),
       width = 8, 
       height = 5,
       dpi = 150)
```

```{r}
ggbetweenstats(
  data = insurance,
  x = year,
  y = total_incurred,
  type = "np",
  title = "Volatility across Years (2022–2024)"
)
```

```{r}
#| echo: false
ggsave("plots/biv_year_volatility.png", 
       plot = p,
       width = 8, 
       height = 5,
       dpi = 150)
```

Mean total incurred costs similarly increase steadily from 2022 to 2024 while the median stays at 0, mirroring the loss ratio trend. The Kruskal-Wallis test confirms statistically significant differences across years (χ² = 215.90, p = 1.31e-47), with 2024 exhibiting the widest spread of extreme incurred values — suggesting claim severity has been worsening year on year.
:::

::: callout-note
The significant difference in sample size across years — 2022 (n = 66,889), 2023 (n = 118,264), and 2024 (n = 167,136) — should be interpreted with caution, as the smaller 2022 sample may underrepresent the true distribution for that year.
:::
