---
title: "Take-Home Exercise 1"
author: "Amelia Rosalind"
date: "May 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installing of required packages

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse, knitr)
```

## Importing data

```{r}
insurance <- read_delim("Motor_insurance.csv",
                        delim = ";",
                        col_types = cols(.default = "c"))
```

# Checking Dataset

Before cleaning, first examine the dataset's structure, summary statistics, and missing values to identify any data quality issues.

```{r}
glimpse(insurance)
summary(insurance)
colSums(is.na(insurance))[colSums(is.na(insurance)) > 0]
kable(head(insurance))
```

## Problems

1.  All features are now characters so need to change the data type of the numerical values into numeric. 

2.  Some features have nulls: vehicle_value has 513 NAs, fuel_type has 1287 NAs, fire_incurred has 94 NAs, occupants_incurred has 25 NAs.

3. Categoricals (policy_type, policy_status, business_type, payment_frequency, bonus_score, fuel_type, municipality_type, circulation_area, vehicle_brand) are stored as characters <chr> instead of factors <fct>. With data type factors, R will understand that it is categoricals and not some string. 

# Data Wrangling

## Cleaning the data

```{r}
insurance <- insurance %>%
  # Fix numerical columns' data type
  mutate(across(c(insured_id, year, driver_age, vehicle_age,
                  age_driving_licence, vehicle_value, 
                  seats, power_to_weight_ratio,
                  total_premium, total_claims,
                  liability_claims, 
                  liability_property_claims,
                  liability_injury_claims, property_claims,
                  theft_claims, fire_claims, glass_claims,
                  legal_protection_claims, occupants_claims,
                  total_incurred, liability_incurred,
                  liability_property_incurred,
                  liability_injury_incurred,
                  property_incurred, theft_incurred,
                  fire_incurred, glass_incurred,
                  legal_protection_incurred,
                  occupants_incurred, liability_premium,
                  property_damage_premium, theft_premium,
                  fire_premium, glass_premium,
                  legal_protection_premium,
                  occupants_premium, total_exposure,
                  liability_exposure),
                ~ as.numeric(gsub(",", ".", .)))) %>%
  # Convert categoricals to factors
  mutate(across(c(policy_type, policy_status, business_type,
                  payment_frequency,bonus_score, fuel_type,
                  municipality_type, circulation_area,
                  vehicle_brand),
                as.factor)) %>%
  drop_na()

glimpse(insurance)
colSums(is.na(insurance))[colSums(is.na(insurance)) > 0]
```

## Feature Engineering

### driving_experience

driving_experience is needed to get how long the driver has been driving for, which is more meaningful as compared to the age when the driver receives the driving licence.

First check if age_driving_licence means the year driving licence is issued or the age the driver got the driving licence. 

```{r}
insurance %>%
  filter(age_driving_licence == 80) %>%
  select(driver_age, age_driving_licence, year)
```

Combining with the first row value shown in the head summary, it can be concluded that age_driving_licence is the age of the driver when driving licence is issued. Hence to get driving_experience the formula should be driver_age - age_driving_licence. 

```{r}
insurance <- insurance %>%
  mutate(driving_experience = 
           driver_age - age_driving_licence)
```

### loss_ratio

This feature will show the ratio of total_incurred to total_premium. This is important as it is the standard insurance profitability metric. 

```{r}
insurance <- insurance %>%
  mutate(loss_ratio = total_incurred / total_premium)
```

There is a possibility that the total_premium is 0 which will produce Inf or NaN instead of a value, so check next.

```{r}
sum(is.infinite(insurance$loss_ratio))
sum(is.nan(insurance$loss_ratio))
```

Since there is only a total of 49 values Inf or NaN, dropping rows with them is negligible.

```{r}
insurance <- insurance %>%
  mutate(loss_ratio = ifelse(is.infinite(loss_ratio) | is.nan(loss_ratio), 
                             NA, loss_ratio)) %>%
  drop_na(loss_ratio)

sum(is.infinite(insurance$loss_ratio))
sum(is.nan(insurance$loss_ratio))
nrow(insurance)
```

## Check final data

```{r}
kable(head(insurance))
```

The final dataset now have 352,219 rows remaining, and 48 columns in total with the new column made. The data types are also all correct. 

# Univariate Analysis

## Distribution of key variables

::: panel-tabset
## Profitability

Key metrics that can influence profitability are variables that concern how much is the cost and revenue. So the most relevant variables to look at first are:

-   total_incurred: the cost side
-   total_premium: the revenue
-   total_claims: frequency of losses
-   loss_ratio: standard insurance profitability metric


```{r}
ggplot(data=insurance, aes(x = total_incurred)) +
  geom_histogram(bins=100) +
  ggtitle("Distribution of Total Incurred (Cost)")
```
:::

