---
title: "Take-Home Exercise 1"
author: "Amelia Rosalind"
date: "May 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Overview

The data used is a real-world vehicle insurance data from [Spanish insurance company](https://data.mendeley.com/datasets/sw4jmdb2sm/1), comprising 354,140 policy records spanning from 2022 to 2024. The objective is to gain deeper insights into data distributions using univariate analysis and understand key factors influencing profitability and volatility using bivariate analysis. This technical report will cover the technical working of this project, such as data preprocessing and all analyses done.

# Getting Started

## Installing of required packages

```{r}
pacman::p_load(ggstatsplot, tidyverse, knitr,
               patchwork, ggthemes, ggdist,
               car)
```

## Importing data

```{r}
insurance <- read_delim("Motor_insurance.csv",
                        delim = ";",
                        col_types = cols(.default = "c"))
```

# Checking Dataset

Before cleaning, first examine the dataset's structure, summary statistics, and missing values to identify any data quality issues.

```{r}
glimpse(insurance)
summary(insurance)
colSums(is.na(insurance))[colSums(is.na(insurance)) > 0]
kable(head(insurance))
```

## Problems

1.  All features are now characters so need to change the data type of the numerical values into numeric.

2.  Some features have nulls: vehicle_value has 513 NAs, fuel_type has 1287 NAs, fire_incurred has 94 NAs, occupants_incurred has 25 NAs.

3.  Categoricals (policy_type, policy_status, business_type, payment_frequency, bonus_score, fuel_type, municipality_type, circulation_area, vehicle_brand) are stored as characters <chr> instead of factors <fct>. With data type factors, R will understand that it is categoricals and not some string.

# Data Wrangling

## Cleaning the data

```{r}
insurance <- insurance %>%
  # Fix numerical columns' data type
  mutate(across(c(insured_id, year, driver_age, vehicle_age,
                  age_driving_licence, vehicle_value, 
                  seats, power_to_weight_ratio,
                  total_premium, total_claims,
                  liability_claims, 
                  liability_property_claims,
                  liability_injury_claims, property_claims,
                  theft_claims, fire_claims, glass_claims,
                  legal_protection_claims, occupants_claims,
                  total_incurred, liability_incurred,
                  liability_property_incurred,
                  liability_injury_incurred,
                  property_incurred, theft_incurred,
                  fire_incurred, glass_incurred,
                  legal_protection_incurred,
                  occupants_incurred, liability_premium,
                  property_damage_premium, theft_premium,
                  fire_premium, glass_premium,
                  legal_protection_premium,
                  occupants_premium, total_exposure,
                  liability_exposure),
                ~ as.numeric(.))) %>%
  # Convert categoricals to factors
  mutate(across(c(policy_type, policy_status, business_type,
                  payment_frequency,bonus_score, fuel_type,
                  municipality_type, circulation_area,
                  vehicle_brand),
                as.factor)) %>%
  drop_na()

glimpse(insurance)
colSums(is.na(insurance))[colSums(is.na(insurance)) > 0]
```

## Feature Engineering

### driving_experience

driving_experience is needed to get how long the driver has been driving for, which is more meaningful as compared to the age when the driver receives the driving licence.

First check if age_driving_licence means the year driving licence is issued or the age the driver got the driving licence.

```{r}
insurance %>%
  filter(age_driving_licence == 80) %>%
  select(driver_age, age_driving_licence, year)
```

Combining with the first row value shown in the head summary, it can be concluded that age_driving_licence is the age of the driver when driving licence is issued. Hence to get driving_experience the formula should be driver_age - age_driving_licence.

```{r}
insurance <- insurance %>%
  mutate(driving_experience = 
           driver_age - age_driving_licence)
```

### loss_ratio

This feature will show the ratio of total_incurred to total_premium. This is important as it is the standard insurance profitability metric.

```{r}
insurance <- insurance %>%
  mutate(loss_ratio = total_incurred / total_premium)
```

There is a possibility that the total_premium is 0 which will produce Inf or NaN instead of a value, so check next.

```{r}
sum(is.infinite(insurance$loss_ratio))
sum(is.nan(insurance$loss_ratio))
```

Since there is only a total of 49 values Inf or NaN, dropping rows with them is negligible.

```{r}
insurance <- insurance %>%
  mutate(loss_ratio = ifelse(is.infinite(loss_ratio) | is.nan(loss_ratio), 
                             NA, loss_ratio)) %>%
  drop_na(loss_ratio)

sum(is.infinite(insurance$loss_ratio))
sum(is.nan(insurance$loss_ratio))
nrow(insurance)
```

## Check final data

```{r}
kable(head(insurance))
```

The final dataset now have 352,289 rows remaining, and 48 columns in total with the new column made. The data types are also all correct.

# Variable selection

Profitability is about how much profit the insurance company can make from a policy, which is addressed with `loss_ratio`. Volatility concerns the spread and unpredictability of losses, which can be addressed by `total_incurred`. As such the target variables are `loss_ratio` and `total_incurred`.

Other variables that can help explain profitability and volatility are listed on the table below

| Variable | Description | Purpose |
|-----------------------|-----------------------------|---------------------|
| `total_premium` | Revenue collected per policy | Provides context on income side of portfolio |
| `total_claims` | Number of claims made | Indicates claim frequency which directly drives costs |
| `driver_age` | Age of driver | Older or younger drivers may present different risk profiles |
| `vehicle_age` | Age of the insured vehicle | Older vehicles may be more probe to certain types of claims |
| `driving_experience` | Years of driving experience | More experienced drivers may have lower claim rates |
| `policy_type` | Types coverage held | Different coverage types carry different risk exposures and premium levels |
| `bonus_score` | Claims history indicator | Drivers with poor claims history are likely higher risk |
| `circulation_area` | Urban vs rural driving environment | Urban areas typically have higher accident rates |

: {tbl-colwidths="[20,20, 60]"}

# Univariate Analysis

For univariate analysis, there are 2 types of models to be done based on if the values are numeric or categorical. 

The numeric ones will use histogram and boxplot together. Histogram will be used to show the shape of the distribution. Boxplot will complement it by highlighting median, IQR and outliers. Both are combined using patchwork for a more complete view.

The categorical values will use bar chart. This type of graph is able to show frequency distribution of each level. 

The table below will show the which features are numerical and categorical and the type of univariate analysis done.

| Variable | Type of value | Univariate analysis |
|-----------------------|-----------------------------|---------------------|
| `total_incurred` | Numerical | histogram + boxplot |
| `loss_ratio` | Numerical | histogram + boxplot |
| `total_premium` | Numerical | histogram + boxplot |
| `total_claims` | Numerical | histogram + boxplot |
| `driver_age` | Numerical | histogram + boxplot |
| `vehicle_age` | Numerical | histogram + boxplot |
| `driving_experience` | Numerical | histogram + boxplot |
| `policy_type` | Categorical | bar chart |
| `bonus_score` | Categorical | bar chart |
| `circulation_area` | Categorical | bar chart |

## Distribution of target variables

::::: panel-tabset
### Volatility

```{r}
summary(insurance$total_incurred)
```

```{r}
box_cost <- ggplot(insurance, aes(y = total_incurred, x = "")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$total_incurred)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-1e+05, 6e+05)) +
  labs(title = "Distribution of Total Incurred (Cost)") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hist_cost <- ggplot(insurance, aes(x = total_incurred)) +
  geom_histogram(bins = 50,
                 colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(total_incurred)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(total_incurred)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-1e+05, 6e+05)) +
  labs(x = "total_incurred") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_cost / hist_cost) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

::: callout-note
There are a lot of zeros as compared to the others, as majority have no claims made. This is known as zero-inflation. 
:::

### Profitability

```{r}
summary(insurance$loss_ratio)
```

```{r}
insurance %>%
  filter(loss_ratio > 100) %>%
  select(total_premium, total_incurred, loss_ratio) %>%
  arrange(desc(loss_ratio))
```

::: callout-note
There are 243 policies with extreme loss ratios exceeding 100, indicating cases where claim cost far outweigh the premium collected. As these are outliers, the distribution is plotted within the range of 0 and 100 to better visualise the majority of policies. 
:::

```{r}
box_ratio <- ggplot(insurance, aes(y = loss_ratio, x = "")) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(x = "", 
                 y = mean(insurance$loss_ratio)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-1, 10)) +
  labs(title = "Distribution of Loss Ratio") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_ratio <- ggplot(insurance, aes(x = loss_ratio)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(loss_ratio)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(loss_ratio)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-1, 10)) +
  labs(x = "loss_ratio") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_ratio / hist_ratio) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```
:::::

## Distribution of other risk factors

:::: panel-tabset
### total_premium

```{r}
summary(insurance$total_premium)
```

```{r}
box_rev <- ggplot(insurance, aes(y = total_premium, x = "")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$total_premium)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-500, 6000)) +
  labs(title = "Distribution of Total Premium (Revenue)") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hist_rev <- ggplot(insurance, aes(x = total_premium)) +
  geom_histogram(bins = 50,
                 colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(total_premium)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(total_premium)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-500, 6000)) +
  labs(x = "total_premium") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_rev / hist_rev) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

### total_claims

```{r}
summary(insurance$total_claims)
```

```{r}
box_claims <- ggplot(insurance, aes(y = total_claims, x="")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$total_claims)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-5, 25)) +
  labs(title = "Distribution of Total Claims (Frequency)") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_claims <- ggplot(insurance, aes(x = total_claims)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(total_claims)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(total_claims)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-5, 25)) +
  labs(x = "total_claims") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_claims / hist_claims) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

::: callout-note
Similar to total_incurred, the majority have no claim, resulting in zero-inflation at value zero.
:::

### driver_age

```{r}
summary(insurance$driver_age)
```

```{r}
box_da <- ggplot(insurance, aes(y = driver_age, x="")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$driver_age)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(15, 95)) +
  labs(title = "Distribution of Driver's Age") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_da <- ggplot(insurance, aes(x = driver_age)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(driver_age)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(driver_age)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(15, 95)) +
  labs(x = "driver_age") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_da / hist_da) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

### vehicle_age

```{r}
summary(insurance$vehicle_age)
```

```{r}
box_va <- ggplot(insurance, aes(y = vehicle_age, x="")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$vehicle_age)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-5, 75)) +
  labs(title = "Distribution of Vehicle Age") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_va <- ggplot(insurance, aes(x = vehicle_age)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(vehicle_age)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(vehicle_age)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-5, 75)) +
  labs(x = "vehicle_age") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_va / hist_va) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

### driving_experience

```{r}
summary(insurance$driving_experience)
```

```{r}
box_de <- ggplot(insurance, aes(y = driving_experience, x="")) +
  geom_boxplot(outlier.colour="lightgreen") +
  geom_point(aes(x = "", 
                 y = mean(insurance$driving_experience)),
             colour = "red", 
             size = 2, 
             shape = 18) +
  coord_flip() +
  scale_y_continuous(limits = c(-5, 75)) +
  labs(title = "Distribution of Driving Experience") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid.major = element_blank(),
     plot.background = element_rect(fill="white",
                                    colour="white")
  )

hist_de <- ggplot(insurance, aes(x = driving_experience)) +
  geom_histogram(colour = "grey25",
                 fill = "grey90") +
  geom_vline(aes(xintercept = mean(driving_experience)), 
             color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = median(driving_experience)), 
             color = "blue", linetype = "dashed") +
  scale_x_continuous(limits = c(-5, 75)) +
  labs(x = "driving_experience") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )

hbox <- (box_de / hist_de) +
  plot_layout(heights = c(1.5, 3.5))

hbox & theme(plot.background = element_rect(fill="white",colour="white"))
```

### policy_type

```{r}
ggplot(data=insurance, aes(x = policy_type)) +
  geom_bar(fill = "grey90") +
  geom_text(stat="count",
            aes(label=paste0(after_stat(count))),
            vjust = -0.2,
            size = 3) +
  labs(title = "Distribution of Policy Types") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )
```

### bonus_score

```{r}
ggplot(data=insurance, aes(x = bonus_score)) +
  geom_bar(fill = "grey90") +
  geom_text(stat="count",
            aes(label=paste0(after_stat(count))),
            vjust = -0.2,
            size = 3) +
  labs(title = "Distribution of Good, Bad, Neutral Claims") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )
```

### circulation_area

```{r}
ggplot(data=insurance, aes(x = circulation_area)) +
  geom_bar(fill = "grey90") +
  geom_text(stat="count",
            aes(label=paste0(after_stat(count))),
            vjust = -0.2,
            size = 3) +
  labs(title = "Distribution of Urban vs Rural") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    plot.background = element_rect(fill="white",
                                   colour="white")
  )
```
::::

# Bivariate Analysis

`ggbetweenstats` and `ggscatterstats` are used to do statistical bivariate analysis. 

## Relationship between bonus score and profitability

One-way ANOVA is used to find the relationship between bonus score and profitability since `bonus_score` has more than 2 groups.

```{r}
# for reproducibility
set.seed(123)
shapiro.test(sample(insurance$loss_ratio, 5000))
```

Since both shows that it is significant, then Kursal-Wallis test is used.

```{r}
ggbetweenstats(
  data = insurance,
  x = bonus_score,
  y = loss_ratio,
  type = "np",
  title = "Relationship between bonus score and profitability"
)
```

## Relationship between policy type and profitability

One-way ANOVA is also used to find the relationship between policy type and profitability since `policy_type` has 5 groups. Since the Shapiro test of `loss_ratio` is significant, then the test done should be non-parametric Kurskal-Wallis test. 

```{r}
ggbetweenstats(
  data = insurance,
  x = policy_type,
  y = loss_ratio,
  type = "np",
  title = "Relationship between policy type and profitability"
)
```

## Relationship between circulation area and profitability

Since circulation area only has 2 groups, the test done should be a two-sample test. In addition, since the Shapiro test for `loss_ratio` is significant, use the non-parametric version which is the Mann-Whitney U test. 

```{r}
ggbetweenstats(
  data = insurance,
  x = circulation_area,
  y = loss_ratio,
  type = "np",
  title = "Relationship between circulation area and profitability"
)
```

## Relationship between years of driving experience and profitability

Since driving experience is a numeric value, the test done should be a Significant Test of Correlation. This is done using `ggscatterplot`. Looking at the univariate analysis, even though `driving_experience` is normally distributed, `loss_ratio` is very skewed. So the non-parametric version is used instead, which is the Spearman's rank correlation coefficient.

```{r}
ggscatterstats(
  data = insurance,
  x = driving_experience,
  y = loss_ratio,
  type = "nonparametric",
  title = "Relationship between years of driving experience and profitability"
)
```

## Relationship between bonus score and volatility

```{r}
# for reproducibility
set.seed(123)
shapiro.test(sample(insurance$total_incurred, 5000))
```

Like against `loss_ratio`, One-Way ANOVA is used. Since, the Shapiro test shows `total_incurred` is significant which means `total_incurred` is not normally distributed. Therefore, non-parametric test, Kruskal-Wallis, is used. 

```{r}
ggbetweenstats(
  data = insurance,
  x = bonus_score,
  y = loss_ratio,
  type = "np",
  title = "Relationship between bonus score and profitability"
)
```

## Relationship between policy type and volatility

Non-parametric version of One-Way ANOVA is also used for this. 

```{r}
ggbetweenstats(
  data = insurance,
  x = bonus_score,
  y = loss_ratio,
  type = "np",
  title = "Relationship between bonus score and profitability"
)
```