---
title: "Take-Home Exercise 1"
author: "Amelia Rosalind"
date: "May 22, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installing of required packages

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse, knitr)
```

## Importing data

```{r}
insurance <- read_csv2("Motor_insurance.csv")
```

# Checking Dataset

```{r}
glimpse(insurance)
summary(insurance)
colSums(is.na(insurance))[colSums(is.na(insurance)) > 0]
kable(head(insurance))
```

## Problems

1.  Some features have incorrect data type: liability_premium, property_damage_premium, theft_premium, fire_premium, glass_premium, legal_protection_premium, occupants_premium, total_exposure and liability_exposure. It should be numeric but registered as characters.

2.  Some features have nulls: vehicle_value has 513 NAs, fuel_type has 1287 NAs, fire_incurred has 94 NAs, occupants_incurred has 25 NAs.

3. Categoricals (policy_type, policy_status, business_type, payment_frequency, bonus_score, fuel_type, municipality_type, circulation_area) are stored as characters <chr> instead of factors <fct>. With data type factors, R will understand that it is categoricals and not some string. 

# Data Wragling

## Cleaning the data

```{r}
insurance <- insurance %>%
  # Fix columns with wrong data types
  mutate(across(c(liability_premium, property_damage_premium, theft_premium,
                  fire_premium, glass_premium, legal_protection_premium,
                  occupants_premium, total_exposure, liability_exposure),
                ~ as.numeric(gsub(",", ".", .)))) %>%
  # Convert categoricals to factors
  mutate(across(c(policy_type, policy_status, business_type, payment_frequency,
                  bonus_score, fuel_type, municipality_type, circulation_area),
                as.factor))  %>%
  drop_na()

glimpse(insurance)
colSums(is.na(insurance))[colSums(is.na(insurance)) > 0]
```

## Feature Engineering

```{r}
insurance %>%
  filter(age_driving_licence == 80) %>%
  select(driver_age, age_driving_licence, year)
```

```{r}
insurance <- insurance %>%
  mutate(driving_experience = 
           driver_age - age_driving_licence)
```

## Check final data

```{r}
kable(head(insurance))
```

