---
title: "Hands-on Ex 7 Part 1 - Modelling, Visualising and Analysing Network Data with R"
author: "Amelia Rosalind"
date: "June 5, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installing of required packages

```{r}
pacman::p_load(igraph, tidygraph, ggraph, 
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts, 
               concaveman, ggforce)
```

# The data

The data sets used in this hands-on exercise is from an oil exploration and extraction company. There are two data sets. One contains the nodes data and the other contains the edges (also know as link) data.

## Importing data

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

## Review imported data

```{r}
glimpse(GAStech_edges)
```

::: callout-warning
The output report of GAStech_edges above reveals that the SentDate is treated as “Character” data type instead of date data type. This is an error! Before we continue, it is important for us to change the data type of SentDate field back to “Date” data type.
:::

## Data Wrangling

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

::: callout-note
-   Both dmy() and wday() are functions of lubridate package. lubridate is an R package that makes it easier to work with dates and times.
-   dmy() transforms the SentDate to Date data type.
-   wday() returns the day of the week as a decimal number or an ordered factor if label is TRUE. The argument abbr is FALSE keep the daya spells in full, i.e. Monday. The function will create a new column in the data.frame i.e. Weekday and the output of wday() will save in this newly created field.
-   The values in the Weekday field are in ordinal scale.
:::

```{r}
glimpse(GAStech_edges)
```

## Wrangling attributes

GAStech_edges data.frame consists of individual e-mail flow records which is not very useful for vsualisation. So aggregate individual by date, senders, receivers, main subject and day of the week.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

::: callout-note
-   Four functions from dplyr package are used. They are: filter(), group(), summarise(), and ungroup().
-   The output data.frame is called GAStech_edges_aggregated.
-   A new field called Weight has been added in GAStech_edges_aggregated.
:::

## Review revised edges file

```{r}
glimpse(GAStech_edges)
```

# Create network objects using tidygraph

Useful information about tidygraph:

-   [Introducing tidygraph](https://www.data-imaginist.com/posts/2017-07-07-introducing-tidygraph/)
-   [tidygraph 1.1 - A tidy hope](https://www.data-imaginist.com/2018/tidygraph-1-1-a-tidy-hope/)

## The tbl_graph object

2 functions of tidygraph package that can be used to create network objects are:

-   tbl_graph(): create tbl_graph network object from nodes and edges data.
-   as_tbl_graph(): convert network data and objects to tbl_graph network. Below are network data and objects supported by `as_tbl_graph()`
    -   a node data.frame and an edge data.frame,
    -   data.frame, list, matrix from base,
    -   igraph from igraph,
    -   network from network,
    -   dendrogram and hclust from stats,
    -   Node from data.tree,
    -   phylo and evonet from ape, and
    -   graphNEL, graphAM, graphBAM from graph (in Bioconductor).

## The dplyr verbs in tidyggraph

activate() verb from tidygraph serves as a switch between tibbles for nodes and edges. All dplyr verbs applied to tbl_graph object are applied to the active tibble.

*.N()* function is used to gain access to the node data while manipulating the edge data. *.E()* will give you the edge data. *.G()* will give you the tbl_graph object itself.

## Using tbl_graph() to build tidygraph data model

Reference guide: [A data structure for tidy graph manipulation](https://tidygraph.data-imaginist.com/reference/tbl_graph.html)

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
```

## Reviewing the output tidygraph's graph object

```{r}
GAStech_graph
```

-   The output above reveals that GAStech_graph is a tbl_graph object with 54 nodes and 4541 edges.
-   The command also prints the first six rows of “Node Data” and the first three of “Edge Data”.
-   It states that the Node Data is active. The notion of an active tibble within a tbl_graph object makes it possible to manipulate the data in one tibble at a time

## Changing the active object

The nodes tibble data frame is activated by default, but you can change which tibble data frame is active with the activate() function. Thus, if we wanted to rearrange the rows in the edges tibble to list those with the highest “weight” first, we could use activate() and then arrange().

For example,

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

Visit reference guide to [activate()](https://tidygraph.data-imaginist.com/reference/activate.html) to find out more.

# Plotting Static Netwoek Graphs with ggaph package

[ggraph](https://ggraph.data-imaginist.com/) is an extension of ggplot2, making it easier to carry over basic ggplot skills to the design of network graphs.

As in all network graph, there are three main aspects to a ggraph’s network graph, they are:

-   [nodes](https://cran.r-project.org/web/packages/ggraph/vignettes/Nodes.html),
-   [edges](https://cran.r-project.org/web/packages/ggraph/vignettes/Edges.html) and
-   [layouts](https://cran.r-project.org/web/packages/ggraph/vignettes/Layouts.html).

## Plotting basic network graph

The code chunk below uses [ggraph()](https://ggraph.data-imaginist.com/reference/ggraph.html), [geom-edge_link()](https://ggraph.data-imaginist.com/reference/geom_edge_link.html) and [geom_node_point()](https://ggraph.data-imaginist.com/reference/geom_node_point.html) to plot a network graph by using GAStech_graph. Before your get started, it is advisable to read their respective reference guide at least once.

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

::: callout-note

The basic plotting function is ggraph(), which takes the data to be used for the graph and the type of layout desired. Both of the arguments for ggraph() are built around igraph. Therefore, ggraph() can use either an igraph object or a tbl_graph object.

:::

## Changing the default netowrk graph theme

In this section, you will use [theme_graph()](https://ggraph.data-imaginist.com/reference/theme_graph.html) to remove the x and y axes. Before your get started, it is advisable to read it’s reference guide at least once.

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```


::: callout-note

-   ggraph introduces a special ggplot theme that provides better defaults for network graphs than the normal ggplot defaults. theme_graph(), besides removing axes, grids, and border, changes the font to Arial Narrow (this can be overridden).
-   The ggraph theme can be set for a series of plots with the set_graph_style() command run before the graphs are plotted or by using theme_graph() in the individual plots.

:::

## Changing the coloting of the plot

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

## Working with ggraph's layout

ggraph support many layout for standard used, they are: star, circle, nicely (default), dh, gem, graphopt, grid, mds, spahere, randomly, fr, kk, drl and lgl. Figures below and on the right show layouts supported by ggraph().


## Fruchterman adn Reingold layout

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

::: callout-note

`layout` argument is used to define the layout to be used.

:::

## Modifying network nodes

Colour each node by referring to their respective departments.

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

::: callout-note

geom_node_point is equivalent in functionality to geo_point of ggplot2. It allows for simple plotting of nodes in different shapes, colours and sizes. In the codes chnuks above colour and size are used.

:::

## Modifying edges

Thickness of edges will be mapped wih the Weight varibale

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

::: callout-note

geom_edge_link draws edges in the simplest way - as straight lines between the start and end nodes. But, it can do more that that. In the example above, argument width is used to map the width of the line in proportional to the Weight attribute and argument alpha is used to introduce opacity on the line.

:::


# Cating Facet Graphs

Another very useful feature of ggraph is faceting. In visualising network data, this technique can be used to reduce edge over-plotting in a very meaning way by spreading nodes and edges out based on their attributes. In this section, you will learn how to use faceting technique to visualise network data.

There are three functions in ggraph to implement faceting, they are:

-   [facet_nodes()](https://ggraph.data-imaginist.com/reference/facet_nodes.html) whereby edges are only draw in a panel if both terminal nodes are present here,
-   [facet_edges()](https://ggraph.data-imaginist.com/reference/facet_edges.html) whereby nodes are always drawn in al panels even if the node data contains an attribute named the same as the one used for the edge facetting, and
-   [facet_graph()](https://ggraph.data-imaginist.com/reference/facet_graph.html) faceting on two variables simultaneously.

## Working with facet_edges()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```

## Working with facet_edges()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

## A framed facet graph

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## Working with facet_nodes()

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

# Network Metrics Analysis

## Computing centrality indices

Centrality measures are a collection of statistical indices use to describe the relative important of the actors are to a network. There are four well-known centrality measures, namely: degree, betweenness, closeness and eigenvector. It is beyond the scope of this hands-on exercise to cover the principles and mathematics of these measure here. Students are encouraged to refer to Chapter 7: Actor Prominence of A User’s Guide to Network Analysis in R to gain better understanding of theses network measures.

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

::: callout-note

mutate() of dplyr is used to perform the computation.
the algorithm used, on the other hand, is the centrality_betweenness() of tidygraph.

:::

## Visualising network metrics

It is important to note that from ggraph v2.0 onward tidygraph algorithms such as centrality measures can be accessed directly in ggraph calls. This means that it is no longer necessary to precompute and store derived node and edge centrality measures on the graph in order to use them in a plot.

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

## Visualising community

tidygraph package inherits many of the community detection algorithms imbedded into igraph and makes them available to us, including Edge-betweenness (group_edge_betweenness), Leading eigenvector (group_leading_eigen), Fast-greedy (group_fast_greedy), Louvain (group_louvain), Walktrap (group_walktrap), Label propagation (group_label_prop), InfoMAP (group_infomap), Spinglass (group_spinglass), and Optimal (group_optimal). Some community algorithms are designed to take into account direction or weight, while others ignore it. Use this [link](https://tidygraph.data-imaginist.com/reference/group_graph.html) to find out more about community detection functions provided by tidygraph,

In the code chunk below group_edge_betweenness() is used.

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(
    group_edge_betweenness(
      weights = Weight, 
      directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(
    aes(
      width=Weight), 
    alpha=0.2) +
  scale_edge_width(
    range = c(0.1, 5)) +
  geom_node_point(
    aes(colour = community))  

g + theme_graph()
```


In order to support effective visual investigation, the community network above has been revised by using `geom_mark_hull()` of ggforce package.

::: callout-warning

Please be reminded that you must to install and include ggforce and concaveman packages before running the code chunk below.

:::

```{r}
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
         betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y, 
        group = community, 
        fill = community),  
    alpha = 0.2,  
    expand = unit(0.3, "cm"),  # Expand
    radius = unit(0.3, "cm")  # Smoothness
  ) + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                      size = betweenness_measure),
                      color = "black",
                      shape = 21)
  
g + theme_graph()
```

# Building Interactive Network Graph with visNetwork

visNetwork() is a R package for network visualization, using vis.js javascript library. visNetwork() function uses a nodes list and edges list to create an interactive graph. The nodes list must include an “id” column, and the edge list must have “from” and “to” columns. The function also plots the labels for the nodes, using the names of the actors from the “label” column in the node list.

The resulting graph is fun to play around with. You can move the nodes and the graph will use an algorithm to keep the nodes properly spaced. You can also zoom in and out on the plot and move it around to re-center it.

## Data preparation

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

## Plotting the first interative network graph

The code chunk below will be used to plot an interactive network graph by using the data prepared.

```{r}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```

## Workig with layout

In the code chunk below, Fruchterman and Reingold layout is used.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

Visit [Igraph](http://datastorm-open.github.io/visNetwork/igraph.html) to find out more about visIgraphLayout’s argument.

## Working with visual attributes - Nodes

visNetwork() looks for a field called “group” in the nodes object and colour the nodes according to the values of the group field.

The code chunk below rename Department field to group.

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

When we rerun the code chunk below, visNetwork shades the nodes by assigning unique colour to each category in the group field.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## Working with visual attributes - Edges

In the code run below visEdges() is used to symbolise the edges.

-   The argument arrows is used to define where to place the arrow.
-   The smooth argument is used to plot the edges using a smooth curve.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

Visit [Option](http://datastorm-open.github.io/visNetwork/edges.html) to find out more about visEdges’s argument.

## Interactivity

In the code chunk below, visOptions() is used to incorporate interactivity features in the data visualisation.

-   The argument highlightNearest highlights nearest when clicking a node.
-   The argument nodesIdSelection adds an id node selection creating an HTML select element.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

